### 2023-10-12 reservoirnet : an R package for reservoir computing/reservoirnet : an R package for reservoir computing.Rmd
initial commit

---
documentclass: jss
author:
  - name: FirstName LastName
    orcid: 0000-0000-0000-0000
    affiliation: University/Company
    # use this syntax to add text on several lines
    address: |
      | First line
      | Second line
    email: \email{name@company.com}
    url: https://posit.co
  - name: Second Author
    orcid: 0000-0000-0000-0000
    affiliation: 'Affiliation \AND'
    # To add another line, use \AND at the end of the previous one as above
  - name: Third Author
    orcid: 0000-0000-0000-0000
    address: |
      | Department of Statistics and Mathematics,
      | Faculty of Biosciences,
      | Universitat Autònoma de Barcelona
    affiliation: |
      | Universitat Autònoma 
      | de Barcelona
    # use a different affiliation in adress field (differently formated here)
    affiliation2: Universitat Autònoma de Barcelona
title:
  formatted: "A Capitalized Title: Something about a Package \\pkg{foo}"
  # If you use tex in the formatted title, also supply version without
  plain:     "A Capitalized Title: Something about a Package foo"
  # For running headers, if needed
  short:     "\\pkg{foo}: A Capitalized Title"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```

# Introduction

## Reservoir networks presentation

Reservoir Computing (RC) is a prominent machine learning methodology that has gained significant attention in recent years for its ability to effectively process information generated by dynamical systems. This innovative approach leverages the dynamics of a high-dimensional reservoir to perform complex computations and solve various tasks based on the response of this dynamical system to input signals.
Reservoir Computing is characterized by its unique architecture, which includes a dynamic system, often referred to as a reservoir, that projects temporal input signals onto a high-dimensional feature space. This high-dimensional representation of input data enables Reservoir Computing to excel in tasks such as time series prediction, pattern recognition, and signal processing.

## Why reservoir are interesting compared to standard approaches (penalised regression) and deep learning approaches?

RC exhibits several advantages over its counterparts. It excels in tasks that require the processing of temporal data, such as time series forecasting and speech recognition, thanks to its inherent memory capabilities. RC offers computational efficiency, reduced susceptibility to overfitting, and a higher degree of interpretability compared to deep learning models. Moreover, RC is able to capture intricate temporal patterns compared to statistical methods such as penalized regression.

## examples of previous use

TODO:
- Ghosh S, Senapati A, Mishra A, Chattopadhyay J, Dana SK, Hens C, et al. Reservoir computing on epidemic spreading: A case study on COVID-19 cases. Phys Rev E. 16 juill 2021;104(1):014308.
- Existing review?

## challenges of using reservoir

RC use is stille challenging for several reasons. First, its stochastic nature due to the reservoir random part makes the results of a given RC unpredictable even if the hyperparameters are well chosen. Second, eventhough, RC relies on several hyperparameter to determine the connections between neurons. There exist few guidance about the way to set the hyperparameters and most often the user must rely on a wrapper approach were combinations of hyperparameters performance are evaluated on a train set before chosing the ideal combination one. Third, most published paper are based on a low dimensional setting and few guidance exist in the context of high dimensional data, especially considering the choice of hyperparameters. Fourth, the use of reservoir computing in an epidemic setting with highly non-stationary time serie is scarce whereas it is of tremendous interest for epidemiologists. Fifth, there is currently no implementation available on R making the use of this method challenging for user not familiar with Python.

## what we do

In this paper we propose to explore those challenges with extensive guidance in order to help new user to fully benefit from RC. First a general introduction to reservoir computing is provided with a tutorial for its use using reservoirnet, a R package based on reservoirPy Python module. Second, we explore the different challenges of the use of RC in a non-stationary high dimensional setting based on the use case of covid-19 hospitalisations forecast with extensive guidance on the modelling strategy, the choice of hyperparameters using a genetic algorithm and the implementation.

# RC presentation

RC is a machine learning algorithm composed of an input layer denoted $u(t)$ which is randomy transformed by the reservoir into an activation state denoted $x(t)$ which is fed to a ridge penalized linear regression trained to foreast the outcome denoted as $y(t)$ as depicted at figure \@ref(fig:rcpresentation).

```{r rcpresentation, echo=FALSE, fig.cap="my caption", out.width = '90%', fig.cap="Reservoir computing is composed of an input layer, a reservoir and an output layer. Connection between input layer and reservoir and inside reservoir are random. Only the output layer is optimized based on a ridge penalized linear regression."}
knitr::include_graphics(here::here("reservoirnet : an R package for reservoir computing/images/schema.png/image1.png"))
```

The input layer $u(t)$ is an $M$-dimension vector which corresponds to the values of the input time series at time $t$ where $t = 1, …, T$. The reservoir layer $x(t)$ is an $Nres$-dimensional vector where $Nres$ is the number of nodes in the reservoir. The value $x(t)$ is defined as follow:

where

The leaking rate alpha define the rate of update of the nodes. The closer alpha is to 1, the higher the reservoir is sensitive to new inputs (i.e u(t)). Therefore, the reservoir state at time t+1 denoted x(t+1) depends on the reservoir state at the previous time (i.e x(t)) and the new inputs (i.e u(t+1)). Both Win and W are random matrices of size Nres x M and Nres x Nres respectively.

Win is a dense matrix generated using a bernouilli distribution where each value can be either -IS(m) or IS(m) with an equal probability where m = 1, …, M corresponds to a given feature in the input layer. The input scaling, denoted IS, is an hyperparameter coefficient wich can be common to all features from the input layer or specific to each feature m. In that case, the more important the feature is, the greater should be its input scaling.  W is a sparse matrix where values are generated base on a gaussian distribution with mean 0 and standard deviation equal to 1. Then the matrix W is rescaled according to the spectral radius, an hyperparameter defining the highest eigen value of W.

The final layer is a linear regression with ridge penalization where the explanatory features are the reservoir state and the variable to be explain is the outcome to predict such that:

where x(t) and y(t) are accumulated in X and Y respectively such that:
and

The parameter lambda is the ridge penalisation which aims to prevent overfitting.

Overall, there are four main hyperparameters to be chosen by the user: the leaking rate wich defines the memory of the RC, the input scaling wich define the relative importance of the features, the spectral radius wich define the connections of the neurons inside the reservoir and the ridge penalization wich defines the degree of overfitting. The choice of hyperparameter is difficult and the use of a wrapper approach where the performance of the RC with different combinations of hyperparameters is evaluated on a train set, the best combination is chosen for the use on the test set.

RC in high dimension

The interest of RC in high dimensional setting is unclear for at least two reasons: (i) RC relies primarily on the projection of the low dimension input layer to the high dimension space of the reservoir neurons and (ii) the high number of features for which each can be associated with a different input scaling increases the hyperparameter search space making standard approaches such as random search or grid search inefficient.

To tackle both problems, we proposed a genetic algorithm for hyperparameter optimisation including feature selection.

    1. Initialize:  
        - Population parameters : Npop = 200, Ne = 100, Ngeneration = 30
        - Cross-over parameters : Ntournament = 2 
        - Mutation parameters : pmutQuant = .5, pmutCat = .25, sigma = 1 (0.1 for leaking rate as the hyperparameter range is shorter)
        - Define the hyperparameter_list, a list containing each param.
        - For each param in hyperparameter_list define hyperparameter search_space: define the lower and upper bound, the type (numeric, integer, categorical) and if it should be log transformed (True, False)

    2. Define tournamentSelection(Npop, Ntournament): 
        - Get the best Npop individuals from previous generations 
        - Randomly select Ntournament challengers among them
        - Get the best individuals from the challengers
        - Repeat twice to get a father and a mother 
        - return father, mother

    3. Define crossoverMutation(pere.param.value, mere.param.value, param.search_space): 
        - sample alp and mut from a Uniform(0,1) distribution
        - If search_space.log is True then  pere.param.value, mere.param.value = log10(pere.param.value), log10(mere.param.value)
        - Cross-over, and define child.param.value as:
            - Numerical : alp*pere.param.value + (1-alp)*mere.param.value
            - Integer : round( alp*pere.param.value + (1-alp)*mere.param.value)
            - Categorical : random choice (list = [pere.param.value, mere.param.value], prob = [alp, 1-alp]) 
        - Mutation if mut < pmutQuant or mut < pmutCat for quantitative and categorical features respectively and update child.param.value
            - Categorical : Randomly selected an other modality 
            - Integer : Randomly add or substract 1 
            - Numerical : Add sigma * N(0,1) 
        - If search_space.log is True then child.param.value = 10^child.param.value
        - Return child.param.value

    4. Define geneticSearch(Npop, Ntournament):
            - pere, mere = tournamentSelection(Npop, Ntournament)
            - for param in  hyperparameter_list:
                - child.param.value = crossoverMutation(pere.param.value, mere.param.value, param.search_space)
            - return child.param

Step 1: Initialize genetic algorithm

Sample Npop individuals (i.e set of hyperparameters) using random search

Step 2: Run genetic algorithm
generation = 1
While generation <  Ngeneration:	
	generation = generation + 1
	sample Npop individuals using  geneticSearch method

3) Basic package use

In this section, we will cover the basics of reservoirnet use including installation, classification and regression. We will provide brief overview of the package use and more in depth description is provided in section 4 with the covid-19 forecast use case.

3.1) Installation

reservoirnet is an R package api making the python module reservoirPy easily callable from R. It is available on CRAN (see https://cran.r-project.org/package=reservoirnet) and can be installed using:

install.packages("reservoirnet")

3.2) Classification

Vignette:
https://cran.r-project.org/web/packages/reservoirnet/vignettes/Classification_with_RC.html

3.3) Regression

Vignette:
https://cran.r-project.org/web/packages/reservoirnet/vignettes/basic_usage_01.html

4) Study case: Prediction covid-19

what is covid-19 and covid-19 pandemic

hospital is a central actor during pandemic, importance to anticipate high number of patients

several models tried to forecast hospitalisations but task is difficult and many models struggled to perform well.

In this setting, reservoir computing seem to be an appropriate approach regarding the task.

4.1) Study case

This use case is based on data from Bordeaux University Hospital. During the pandemic a predictive model was elaborate and regularly use to forecast 14 days ahead the number of hospitalized patients at the hospital. In this work we will compare the first approach based on linear regression with elastic-net penalisation to reservoir computing approach.

4.2) Evaluation framework

- Train and test set

The task was to forecast 14 days ahead the number of hospitalised patients. The dataset was separated into two periods. First period from xx to xx served to identify relevant hyperparameters. Second period after xx was used to evaluate the model performance. As the time serie is not stationary and most appropriate hyperparameter value might change over time, we also investigated the effect of hyperparameter update every month.

The performance of the model was evaluated according to several metrics: the median absolute error, the median relative error, the absolute error relative to baseline, the relative error relative to baseline. For each metric, the median was chosen over the mean to be robust to outliers. For instance, for relative error, when the outcome is low (e.g equal to 1 hospitalisation), an error of 5 hospitalisations could results to a 500% relative error.

Median absolute error (AE): 
Median relative error (RE):

- Baseline model (forecast at T+14 = observed at T)
- Metrics

4.3) Hyperparameter optimisation using GA

- Recall hyperparameters
- Present GA algorithm + explain why it is important here (large number of hp + assumption free compared to other algorithm, cons = expensive)

4.4) Present results from GA

- Present results and detail them

5) Discussion and conclusion

<!-- ## Code formatting -->

<!-- In general, don't use Markdown, but use the more precise LaTeX commands instead: -->

<!-- * \proglang{Java} -->
<!-- * \pkg{plyr} -->

<!-- One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax). -->

<!-- If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute. You can also provide a custom identifier if necessary. See the header of Section \ref{r-code} for example. -->

<!-- # \proglang{R} code {short-title="R code" #r-code} -->

<!-- Can be inserted in regular R markdown blocks. -->

<!-- ```{r} -->
<!-- x <- 1:10 -->
<!-- x -->
<!-- ``` -->

<!-- ## Features specific to \pkg{rticles} {short-title="Features specific to rticles"} -->

<!-- * Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter). This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409). -->
<!-- * Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format. -->

### 2023-10-12 reservoirnet : an R package for reservoir computing/reservoirnet : an R package for reservoir computing.Rmd
update jss

@@
 The input layer $u(t)$ is an $M$-dimension vector which corresponds to the values of the input time series at time $t$ where $t = 1, …, T$.
-The reservoir layer $x(t)$ is an $Nres$-dimensional vector where $Nres$ is the number of nodes in the reservoir.
+The reservoir layer $x(t)$ is an $N_{res}$-dimensional vector where $N_{res}$ is the number of nodes in the reservoir.
 The value $x(t)$ is defined as follow:
 
-where
+$$x( t+1 ) = ( 1 - \alpha )  x ( t) + \alpha \: tanh( W x(t) + W_{in} u(t+1) ) \text{ where } \alpha \in [0, 1 ]$$
 
 The leaking rate alpha define the rate of update of the nodes.
-The closer alpha is to 1, the higher the reservoir is sensitive to new inputs (i.e u(t)).
-Therefore, the reservoir state at time t+1 denoted x(t+1) depends on the reservoir state at the previous time (i.e x(t)) and the new inputs (i.e u(t+1)).
-Both Win and W are random matrices of size Nres x M and Nres x Nres respectively.
+The closer $\alpha$ is to $1$, the higher the reservoir is sensitive to new inputs (i.e $u(t)$).
+Therefore, the reservoir state at time $t+1$ denoted $x(t+1)$ depends on the reservoir state at the previous time (i.e $x(t)$) and the new inputs (i.e $u(t+1)$).
+Both $W_{in}$ and $W$ are random matrices of size $Nres \times M$ and $Nres \times Nres$ respectively.
 
-Win is a dense matrix generated using a bernouilli distribution where each value can be either -IS(m) or IS(m) with an equal probability where m = 1, …, M corresponds to a given feature in the input layer.
-The input scaling, denoted IS, is an hyperparameter coefficient wich can be common to all features from the input layer or specific to each feature m.
+$W_{in}$ is a dense matrix generated using a bernouilli distribution where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an equal probability where $m = 1, …, M$ corresponds to a given feature in the input layer.
+The input scaling, denoted $I_{scale}$, is an hyperparameter coefficient which can be common to all features from the input layer or specific to each feature $m$.
 In that case, the more important the feature is, the greater should be its input scaling.
-W is a sparse matrix where values are generated base on a gaussian distribution with mean 0 and standard deviation equal to 1.
-Then the matrix W is rescaled according to the spectral radius, an hyperparameter defining the highest eigen value of W.
+$W$ is a sparse matrix where values are generated base on a gaussian distribution $\mathcal{N}(0,1)$.
+Then the matrix $W$ is rescaled according to the spectral radius, an hyperparameter defining the highest eigen value of $W$.
 
@@
 
-where x(t) and y(t) are accumulated in X and Y respectively such that:
-and
+$$W_{out} = YX^T ( XX^T + \lambda  I)^{ -1 }$$
 
-The parameter lambda is the ridge penalisation which aims to prevent overfitting.
+Where x(t) and y(t) are accumulated in X and Y respectively such that:
+
+$$X = \begin{bmatrix} x(1) \\ x(2) \\ ... \\ x(T) \end{bmatrix}
+\text{ and } Y = \begin{bmatrix} y(1) \\ y(2) \\ ... \\ y(T) \end{bmatrix}$$
+
+The parameter $\lambda$ is the ridge penalisation which aims to prevent overfitting.
 
@@
 
-    1. Initialize:  
+1. Initialize :
   - Population parameters : Npop = 200, Ne = 100, Ngeneration = 30
@@
   - Define the hyperparameter_list, a list containing each param.
-        - For each param in hyperparameter_list define hyperparameter search_space: define the lower and upper bound, the type (numeric, integer, categorical) and if it should be log transformed (True, False)
+  - For each param in hyperparameter_list define hyperparameter search_space : define the lower and upper bound, the type (numeric, integer, categorical) and if it should be log transformed (True, False)
 
@@
     - return father, mother 
-
 3. Define crossoverMutation(pere.param.value, mere.param.value, param.search_space): 
@@
     - If search_space.log is True then  pere.param.value, mere.param.value = log10(pere.param.value), log10(mere.param.value)
-        - Cross-over, and define child.param.value as:
+    - Cross-over, and define child.param.value as :
         - Numerical : alp*pere.param.value + (1-alp)*mere.param.value
@@
 
-    4. Define geneticSearch(Npop, Ntournament):
+4. Define geneticSearch(Npop, Ntournament) :
     - pere, mere = tournamentSelection(Npop, Ntournament)
-            - for param in  hyperparameter_list:
+    - for param in  hyperparameter_list :
         - child.param.value = crossoverMutation(pere.param.value, mere.param.value, param.search_space)
@@
 
-Step 1: Initialize genetic algorithm
+Step 1 : Initialize genetic algorithm
 
-Sample Npop individuals (i.e set of hyperparameters) using random search
+- Sample Npop individuals (i.e set of hyperparameters) using random search
 
-Step 2: Run genetic algorithm
-generation = 1
-While generation <  Ngeneration:	
+Step 2 : Run genetic algorithm
+- generation = 1
+While generation <  Ngeneration :	
     generation = generation + 1

### 2023-10-13 draft_jss/draft_jss.Rmd
rename jss

--- 2023-10-12 reservoirnet : an R package for reservoir computing/reservoirnet : an R package for reservoir computing.Rmd
@@
     base_format: rticles::jss_article
+  bookdown::word_document2:
+    default
+knit: (function(inputFile, encoding){
+  rmarkdown::render(inputFile, encoding = encoding,
+  output_dir = here::here("output"), output_format = "all") })
 ---
@@
 
-Reservoir Computing (RC) is a prominent machine learning methodology that has gained significant attention in recent years for its ability to effectively process information generated by dynamical systems.
-This innovative approach leverages the dynamics of a high-dimensional reservoir to perform complex computations and solve various tasks based on the response of this dynamical system to input signals.
-Reservoir Computing is characterized by its unique architecture, which includes a dynamic system, often referred to as a reservoir, that projects temporal input signals onto a high-dimensional feature space.
-This high-dimensional representation of input data enables Reservoir Computing to excel in tasks such as time series prediction, pattern recognition, and signal processing.
+Reservoir Computing (RC) is a prominent machine learning methodology
+that has gained significant attention in recent years for its ability to
+effectively process information generated by dynamical systems. This
+innovative approach leverages the dynamics of a high-dimensional
+reservoir to perform complex computations and solve various tasks based
+on the response of this dynamical system to input signals. Reservoir
+Computing is characterized by its unique architecture, which includes a
+dynamic system, often referred to as a reservoir, that projects temporal
+input signals onto a high-dimensional feature space. This
+high-dimensional representation of input data enables Reservoir
+Computing to excel in tasks such as time series prediction, pattern
+recognition, and signal processing.
 
@@
 
-RC exhibits several advantages over its counterparts.
-It excels in tasks that require the processing of temporal data, such as time series forecasting and speech recognition, thanks to its inherent memory capabilities.
-RC offers computational efficiency, reduced susceptibility to overfitting, and a higher degree of interpretability compared to deep learning models.
-Moreover, RC is able to capture intricate temporal patterns compared to statistical methods such as penalized regression.
+RC exhibits several advantages over its counterparts. It excels in tasks
+that require the processing of temporal data, such as time series
+forecasting and speech recognition, thanks to its inherent memory
+capabilities. RC offers computational efficiency, reduced susceptibility
+to overfitting, and a higher degree of interpretability compared to deep
+learning models. Moreover, RC is able to capture intricate temporal
+patterns compared to statistical methods such as penalized regression.
 
@@
 
-TODO:
-- Ghosh S, Senapati A, Mishra A, Chattopadhyay J, Dana SK, Hens C, et al. Reservoir computing on epidemic spreading: A case study on COVID-19 cases. Phys Rev E. 16 juill 2021;104(1):014308.
-- Existing review?
+TODO: - Ghosh S, Senapati A, Mishra A, Chattopadhyay J, Dana SK, Hens C,
+et al. Reservoir computing on epidemic spreading: A case study on
+COVID-19 cases. Phys Rev E. 16 juill 2021;104(1):014308. - Existing
+review?
 
@@
 
-RC use is stille challenging for several reasons.
-First, its stochastic nature due to the reservoir random part makes the results of a given RC unpredictable even if the hyperparameters are well chosen.
-Second, eventhough, RC relies on several hyperparameter to determine the connections between neurons.
-There exist few guidance about the way to set the hyperparameters and most often the user must rely on a wrapper approach were combinations of hyperparameters performance are evaluated on a train set before chosing the ideal combination one.
-Third, most published paper are based on a low dimensional setting and few guidance exist in the context of high dimensional data, especially considering the choice of hyperparameters.
-Fourth, the use of reservoir computing in an epidemic setting with highly non-stationary time serie is scarce whereas it is of tremendous interest for epidemiologists.
-Fifth, there is currently no implementation available on R making the use of this method challenging for user not familiar with Python.
+RC use is stille challenging for several reasons. First, its stochastic
+nature due to the reservoir random part makes the results of a given RC
+unpredictable even if the hyperparameters are well chosen. Second,
+eventhough, RC relies on several hyperparameter to determine the
+connections between neurons. There exist few guidance about the way to
+set the hyperparameters and most often the user must rely on a wrapper
+approach were combinations of hyperparameters performance are evaluated
+on a train set before chosing the ideal combination one. Third, most
+published paper are based on a low dimensional setting and few guidance
+exist in the context of high dimensional data, especially considering
+the choice of hyperparameters. Fourth, the use of reservoir computing in
+an epidemic setting with highly non-stationary time serie is scarce
+whereas it is of tremendous interest for epidemiologists. Fifth, there
+is currently no implementation available on R making the use of this
+method challenging for user not familiar with Python.
 
@@
 
-In this paper we propose to explore those challenges with extensive guidance in order to help new user to fully benefit from RC.
-First a general introduction to reservoir computing is provided with a tutorial for its use using reservoirnet, a R package based on reservoirPy Python module.
-Second, we explore the different challenges of the use of RC in a non-stationary high dimensional setting based on the use case of covid-19 hospitalisations forecast with extensive guidance on the modelling strategy, the choice of hyperparameters using a genetic algorithm and the implementation.
+In this paper we propose to explore those challenges with extensive
+guidance in order to help new user to fully benefit from RC. First a
+general introduction to reservoir computing is provided with a tutorial
+for its use using reservoirnet, a R package based on reservoirPy Python
+module. Second, we explore the different challenges of the use of RC in
+a non-stationary high dimensional setting based on the use case of
+covid-19 hospitalisations forecast with extensive guidance on the
+modelling strategy, the choice of hyperparameters using a genetic
+algorithm and the implementation.
 
@@
 
-RC is a machine learning algorithm composed of an input layer denoted $u(t)$ which is randomy transformed by the reservoir into an activation state denoted $x(t)$ which is fed to a ridge penalized linear regression trained to foreast the outcome denoted as $y(t)$ as depicted at figure \@ref(fig:rcpresentation).
+RC is a machine learning algorithm composed of an input layer denoted
+$u(t)$ which is randomy transformed by the reservoir into an activation
+state denoted $x(t)$ which is fed to a ridge penalized linear regression
+trained to foreast the outcome denoted as $y(t)$ as depicted at figure
+\@ref(fig:rcpresentation).
 
@@
 
-The input layer $u(t)$ is an $M$-dimension vector which corresponds to the values of the input time series at time $t$ where $t = 1, …, T$.
-The reservoir layer $x(t)$ is an $N_{res}$-dimensional vector where $N_{res}$ is the number of nodes in the reservoir.
-The value $x(t)$ is defined as follow:
+The input layer $u(t)$ is an $M$-dimension vector which corresponds to
+the values of the input time series at time $t$ where $t = 1, …, T$. The
+reservoir layer $x(t)$ is an $N_{res}$-dimensional vector where
+$N_{res}$ is the number of nodes in the reservoir. The value $x(t)$ is
+defined as follow:
 
@@
 
-The leaking rate alpha define the rate of update of the nodes.
-The closer $\alpha$ is to $1$, the higher the reservoir is sensitive to new inputs (i.e $u(t)$).
-Therefore, the reservoir state at time $t+1$ denoted $x(t+1)$ depends on the reservoir state at the previous time (i.e $x(t)$) and the new inputs (i.e $u(t+1)$).
-Both $W_{in}$ and $W$ are random matrices of size $Nres \times M$ and $Nres \times Nres$ respectively.
+The leaking rate alpha define the rate of update of the nodes. The
+closer $\alpha$ is to $1$, the higher the reservoir is sensitive to new
+inputs (i.e $u(t)$). Therefore, the reservoir state at time $t+1$
+denoted $x(t+1)$ depends on the reservoir state at the previous time
+(i.e $x(t)$) and the new inputs (i.e $u(t+1)$). Both $W_{in}$ and $W$
+are random matrices of size $Nres \times M$ and $Nres \times Nres$
+respectively.
 
-$W_{in}$ is a dense matrix generated using a bernouilli distribution where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an equal probability where $m = 1, …, M$ corresponds to a given feature in the input layer.
-The input scaling, denoted $I_{scale}$, is an hyperparameter coefficient which can be common to all features from the input layer or specific to each feature $m$.
-In that case, the more important the feature is, the greater should be its input scaling.
-$W$ is a sparse matrix where values are generated base on a gaussian distribution $\mathcal{N}(0,1)$.
-Then the matrix $W$ is rescaled according to the spectral radius, an hyperparameter defining the highest eigen value of $W$.
+$W_{in}$ is a dense matrix generated using a bernouilli distribution
+where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an
+equal probability where $m = 1, …, M$ corresponds to a given feature in
+the input layer. The input scaling, denoted $I_{scale}$, is an
+hyperparameter coefficient which can be common to all features from the
+input layer or specific to each feature $m$. In that case, the more
+important the feature is, the greater should be its input scaling. $W$
+is a sparse matrix where values are generated base on a gaussian
+distribution $\mathcal{N}(0,1)$. Then the matrix $W$ is rescaled
+according to the spectral radius, an hyperparameter defining the highest
+eigen value of $W$.
 
-The final layer is a linear regression with ridge penalization where the explanatory features are the reservoir state and the variable to be explain is the outcome to predict such that:
+The final layer is a linear regression with ridge penalization where the
+explanatory features are the reservoir state and the variable to be
+explain is the outcome to predict such that:
 
@@
 
-The parameter $\lambda$ is the ridge penalisation which aims to prevent overfitting.
+The parameter $\lambda$ is the ridge penalisation which aims to prevent
+overfitting.
 
-Overall, there are four main hyperparameters to be chosen by the user: the leaking rate wich defines the memory of the RC, the input scaling wich define the relative importance of the features, the spectral radius wich define the connections of the neurons inside the reservoir and the ridge penalization wich defines the degree of overfitting.
-The choice of hyperparameter is difficult and the use of a wrapper approach where the performance of the RC with different combinations of hyperparameters is evaluated on a train set, the best combination is chosen for the use on the test set.
+Overall, there are four main hyperparameters to be chosen by the user:
+the leaking rate wich defines the memory of the RC, the input scaling
+wich define the relative importance of the features, the spectral radius
+wich define the connections of the neurons inside the reservoir and the
+ridge penalization wich defines the degree of overfitting. The choice of
+hyperparameter is difficult and the use of a wrapper approach where the
+performance of the RC with different combinations of hyperparameters is
+evaluated on a train set, the best combination is chosen for the use on
+the test set.
 
@@
 
-The interest of RC in high dimensional setting is unclear for at least two reasons: (i) RC relies primarily on the projection of the low dimension input layer to the high dimension space of the reservoir neurons and (ii) the high number of features for which each can be associated with a different input scaling increases the hyperparameter search space making standard approaches such as random search or grid search inefficient.
+The interest of RC in high dimensional setting is unclear for at least
+two reasons: (i) RC relies primarily on the projection of the low
+dimension input layer to the high dimension space of the reservoir
+neurons and (ii) the high number of features for which each can be
+associated with a different input scaling increases the hyperparameter
+search space making standard approaches such as random search or grid
+search inefficient.
 
-To tackle both problems, we proposed a genetic algorithm for hyperparameter optimisation including feature selection.
+To tackle both problems, we proposed a genetic algorithm for
+hyperparameter optimisation including feature selection.
 
 1.  Initialize :
-  - Population parameters : Npop = 200, Ne = 100, Ngeneration = 30
-  - Cross-over parameters : Ntournament = 2 
-  - Mutation parameters : pmutQuant = .5, pmutCat = .25, sigma = 1 (0.1 for leaking rate as the hyperparameter range is shorter)
-  - Define the hyperparameter_list, a list containing each param.
-  - For each param in hyperparameter_list define hyperparameter search_space : define the lower and upper bound, the type (numeric, integer, categorical) and if it should be log transformed (True, False)
 
-2. Define tournamentSelection(Npop, Ntournament): 
-    - Get the best Npop individuals from previous generations 
-    - Randomly select Ntournament challengers among them
+-   Population parameters : $N_{pop} = 200$, $N_e = 100$,
+    $N_{generation} = 30$
+-   Cross-over parameters : $N_{tournament} = 2$
+-   Mutation parameters : $p_{mutQuant} = .5$, $p_{mutCat} = .25$,
+    $\sigma = 1$ ($\sigma = 0.1$ for leaking rate as the hyperparameter
+    range is shorter)
+-   Define the $hyperparameter\_list$, a list containing each
+    hyperparameter denoted as $param$.
+-   `for` each $param$ in $hyperparameter\_list$ define hyperparameter
+    $search\_space$ :
+    -   the $lower$ and $upper$ bounds,
+    -   the $type$ (numeric, integer, categorical)
+    -   if it should be $log$ transformed (True, False)
+
+2.  Define main functions
+    -   Define $tournamentSelection(N_{pop}, N_{tournament})$:
+        -   Get the best $N_{pop}$ individuals from previous generations
+        -   Randomly select $N_{tournament}$ challengers among them
         -   Get the best individuals from the challengers
         -   Repeat twice to get a father and a mother
-    - return father, mother 
-3. Define crossoverMutation(pere.param.value, mere.param.value, param.search_space): 
-    - sample alp and mut from a Uniform(0,1) distribution
-    - If search_space.log is True then  pere.param.value, mere.param.value = log10(pere.param.value), log10(mere.param.value)
-    - Cross-over, and define child.param.value as :
-        - Numerical : alp*pere.param.value + (1-alp)*mere.param.value
-        - Integer : round( alp*pere.param.value + (1-alp)*mere.param.value)
-        - Categorical : random choice (list = [pere.param.value, mere.param.value], prob = [alp, 1-alp]) 
-    - Mutation if mut < pmutQuant or mut < pmutCat for quantitative and categorical features respectively and update child.param.value
-        - Categorical : Randomly selected an other modality 
-        - Integer : Randomly add or substract 1 
-        - Numerical : Add sigma * N(0,1) 
-    - If search_space.log is True then child.param.value = 10^child.param.value
-    - Return child.param.value
+        -   `return` $father, \: mother$
+    -   Define
+        $crossoverMutation(pere\$param\$value, mere\$param\$value, param\$search\_space)$:
+        -   `if` $search\_space\$log$ $\newline$
+            $\text{} \qquad pere\$param\$value = log_{10}(pere\$param\$value)$
+            $\newline$
+            $\text{} \qquad mere\$param\$value = log_{10}(mere\$param\$value)$
 
-4. Define geneticSearch(Npop, Ntournament) :
-    - pere, mere = tournamentSelection(Npop, Ntournament)
-    - for param in  hyperparameter_list :
-        - child.param.value = crossoverMutation(pere.param.value, mere.param.value, param.search_space)
-    - return child.param
+        -   sample $alp$ and $mut$ probabilities from $\mathcal{U}(0,1)$
 
-Step 1 : Initialize genetic algorithm
+        -   Cross-over, and define $child\$param\$value$ as :
 
-- Sample Npop individuals (i.e set of hyperparameters) using random search
+            -   Numerical :
+                $alp*pere\$param\$value + (1-alp)*mere\$param\$value$
+            -   Integer :
+                $round( alp*pere\$param\$value + (1-alp)*mere\$param\$value)$
+            -   Categorical :
+                $sample(x = c(pere\$param\$value, mere\$param\$value), \newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad prob = c(alp, 1-alp))$
 
-Step 2 : Run genetic algorithm
-- generation = 1
-While generation <  Ngeneration :	
-    generation = generation + 1
-    sample Npop individuals using geneticSearch method
+        -   `if` $mut < p_{mutQuant}$ or $mut < p_{mutCat}$ for
+            quantitative and categorical features respectively, mutate :
 
-3) Basic package use
+            -   Categorical : sample another modality
+            -   Integer : add or substract 1 at random
+            -   Numerical : Add $\sigma \times \mathcal{N}(0,1)$
 
-In this section, we will cover the basics of reservoirnet use including installation, classification and regression.
-We will provide brief overview of the package use and more in depth description is provided in section 4 with the covid-19 forecast use case.
+        -   `if` $search\_space\$log$ :
+            $child\$param\$value = 10^{child\$param\$value}$
 
-3.1) Installation
+        -   `return` $child\$param\$value$
+    -   Define $geneticSearch(N_{pop}, N_{tournament})$ :
+        -   $pere, \: mere = tournamentSelection(N_{pop}, N_{tournament})$
+        -   `for` $param$ in $hyperparameter\_list$ : $\newline$
+            $\text{} \qquad child\$param\$value = crossoverMutation(\newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad pere\$param\$value, \newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad mere\$param\$value, \newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad param\$search\_space)$
+        -   `return` $child$
+3.  Run the algorithm
+    -   Step 1 : Initialize genetic algorithm
+        -   Sample $N_{pop}$ individuals (i.e set of hyperparameters)
+            using random search
+    -   Step 2 : Run genetic algorithm
+        -   $generation = 1$
+        -   `while` $generation < Ngeneration$ :
+            -   $generation = generation + 1$
+            -   sample $N_{pop}$ individuals using $geneticSearch$
+                function
 
-reservoirnet is an R package api making the python module reservoirPy easily callable from R. It is available on CRAN (see https://cran.r-project.org/package=reservoirnet) and can be installed using:
+# Basic package use
 
+In this section, we will cover the basics of reservoirnet use including
+installation, classification and regression. We will provide brief
+overview of the package use and more in depth description is provided in
+section 4 with the covid-19 forecast use case.
+
+## Installation
+
+reservoirnet is an R package api making the python module reservoirPy
+easily callable from R. It is available on CRAN (see
+<https://cran.r-project.org/package=reservoirnet>) and can be installed
+using:
+
+```{r eval=FALSE}
 install.packages("reservoirnet")
+```
 
-3.2) Classification
+## Classification
+
+Reservoir Computing (RC) is well suited to both regression and classification tasks.
+We will introduce a simple example of classification task.
+This chapter is largely inspired from reservoirnet vignette on
+[classification](https://cran.r-project.org/web/packages/reservoirnet/vignettes/Classification_with_RC.html).
+
+The first step is to load useful packages. For this example we will load both reservoirnet and ggplot2 (add ref) a useful package for datavisualisation.
+
+```{r}
+library(reservoirnet)
+library(ggplot2)
+```
+
+### The Japanese vowel dataset
+
+The Japanese vowel dataset is composed of 640 utterances of the Japanese vowel \ae, from 9 different male speakers.
+The goal of this task is to assign to each utterance the label of its speaker.
+Dataset is split between a 270 utterances training set and a 340 utterances testing set.
+Each spoken utterance is a timeseries of 7\~29 timesteps.
+Each timestep of signal is a 12 dimensional vector representing Linear Prediction Coefficient (LPC), which encode the audio signal into the cepstral domain (a variant of the frequency domain).
+
+References M. Kudo, J. Toyama and M. Shimbo. (1999). "Multidimensional
+Curve Classification Using Passing-Through Regions". Pattern Recognition
+Letters, Vol. 20, No. 11--13, pages 1103--1111.
+<https://archive.ics.uci.edu/ml/machine-learning-databases/JapaneseVowels-mld/>
+
+The data set is splitted into a train and a test set.
+
+```{r}
+japanese_vowels <- reservoirnet::generate_data(dataset = "japanese_vowels")$japanese_vowels
+
+X_train <- japanese_vowels$X_train
+Y_train <- japanese_vowels$Y_train
+X_test <- japanese_vowels$X_test
+Y_test <- japanese_vowels$Y_test
+```
+
+```{r fig.cap="Vowel dataset"}
+sample_per_speaker <- 30
+n_speaker <- 9
+X_train_per_speaker <- list()
+
+for (i in 0:8) {
+ X_speaker <- X_train[((i*sample_per_speaker)+1):((i+1)*sample_per_speaker)]
+ X_train_per_speaker[[i+1]]<-(as.numeric(unlist(sapply(X_speaker, t))))
+}
+
+d <- data.frame(LPC = unlist(X_train_per_speaker), 
+                Speaker = factor(rep(1:9,times = sapply(X_train_per_speaker,length))))
+
+# plot data
+ggplot(d,aes(x = Speaker, y = LPC)) + geom_boxplot()+theme_bw()
+```
+
+### Echo state network setup
+
+#### Transduction (sequence-to-sequence model)
+
+As ReservoirPy Nodes are built to work on sequences, the simplest setup
+to solve this task is sequence-to-sequence encoding, also called
+transduction. A model is trained on encoding each vector of input
+sequence into a new vector in the output space. Thus, a sequence of
+audio yields a sequence of label, one label per timestep.
+
+```{r}
+japanese_vowels <- reservoirnet::generate_data(
+    dataset = "japanese_vowels",
+    repeat_targets=TRUE)$japanese_vowels
+
+X_train <- japanese_vowels$X_train
+Y_train <- japanese_vowels$Y_train
+X_test <- japanese_vowels$X_test
+Y_test <- japanese_vowels$Y_test
+```
+
+#### Train a simple Echo State Network to solve this task:
+
+```{r}
+source <- createNode("Input")
+readout <- createNode("Ridge",ridge=1e-6)
+reservoir <- createNode("Reservoir",units = 500,lr=0.1, sr=0.9)
+
+#[source >> reservoir, source] >> readout
+model <- list(source %>>% reservoir, source) %>>% readout
+```
+
+#### Fit the model
+
+```{r}
+model_fit <- reservoirnet::reservoirR_fit(node = model,
+                                       X = X_train,
+                                       Y = Y_train,
+                                       stateful = FALSE,
+                                       warmup = 2)
+
+Y_pred <- reservoirnet::predict_seq(node = model_fit$fit,
+                                 X = X_test,
+                                 stateful = FALSE)
+```
+
+Get the scores:
+
+There are 9 speakers, hence the output space is 9-dimensional. The
+speaker label is the index of the output neuron with maximum activation.
+
+```{r}
+accuracy <- function(pred, truth){
+  mean(pred == truth)
+}
+```
+
+```{r}
+Y_pred_class <- sapply(Y_pred, FUN = function(x) apply(as.matrix(x),1,which.max))
+Y_test_class <- sapply(Y_test, FUN = function(x) apply(as.matrix(x),1,which.max))
+score <- accuracy(array(unlist(Y_pred_class)), array(unlist(Y_test_class)))
+
+print(paste0("Accuracy: ", round(score * 100,3) ,"%"))
+```
+
+    ## [1] "Accuracy: 91.296%"
+
+### Classification (sequence-to-vector model)
+
+We can create a more elaborated model where inference is performed only
+once on the whole input sequence. Indeed, we only need to assign one
+label to each input sequence. This new setup is known as a
+sequence-to-vector model, and this is usually the type of model we refer
+to when talking about classification of sequencial patterns.
+
+```{r}
+japanese_vowels <- reservoirnet::generate_data(
+    dataset = "japanese_vowels")$japanese_vowels
+
+X_train <- japanese_vowels$X_train
+Y_train <- japanese_vowels$Y_train
+X_test <- japanese_vowels$X_test
+Y_test <- japanese_vowels$Y_test
+```
+
+```{r}
+source <- reservoirnet::createNode("Input")
+readout <- reservoirnet::createNode("Ridge",ridge=1e-6)
+reservoir <- reservoirnet::createNode("Reservoir", units = 500, lr=0.1, sr=0.9)
+
+#source >> reservoir >> readout
+model <- source %>>% reservoir %>>% readout
+```
+
+We need to modify the training loop by hand a bit to perform this task:
+
+first, we compute all reservoir states over the input sequence using the
+reservoir.run method. then, we gather in a list only the last vector of
+the states sequence.
+
+```{r}
+states_train = list()
+k <- 1
+for (x in X_train) {
+  states <- reservoirnet::predict_seq(node = reservoir, X = x, reset=TRUE)
+  states_train[[k]] <- t(as.matrix(states[nrow(states),]))
+  k <- k+1
+}
+```
+
+We can now train the readout only on the last state vectors. Here,
+Y_train is an array storing a single label for each utterance.
+
+```{r}
+res <- reservoirnet::reservoirR_fit(readout,X = states_train, Y = Y_train)
+summary(res)
+```
+
+    ## Parametrs using to fit:
+    ##  warmup: 0 ; stateful: FALSE ; reset: FALSE 
+    ## results of fitting:
+    ## 'Ridge-5': Ridge(ridge=1e-06, input_bias=True, in=500, out=9)
+
+We also modify the inference code using the same method as above:
+
+```{r}
+Y_pred <- list()
+k <- 1
+for (x in X_test) {
+  states <- reservoirnet::predict_seq(node = reservoir, X = x, reset=TRUE)
+  y <- reservoirnet::predict_seq(node = readout, X = as.array(states[nrow(states),]))
+  Y_pred[[k]] <- y
+  k <- k+1
+}
+```
+
+```{r}
+Y_pred_class <- sapply(Y_pred, FUN = function(x) apply(as.matrix(x),1,which.max))
+Y_test_class <- sapply(Y_test, FUN = function(x) apply(as.matrix(x),1,which.max))
+
+score <- accuracy(pred = Y_test_class,
+                       truth = Y_pred_class)
+
+print(paste0("Accuracy: ", round(score * 100,3) ,"%"))
+```
+
+    ## [1] "Accuracy: 88.108%"
+
+## Regression
 
 Vignette:
-https://cran.r-project.org/web/packages/reservoirnet/vignettes/Classification_with_RC.html
+<https://cran.r-project.org/web/packages/reservoirnet/vignettes/basic_usage_01.html>
 
-3.3) Regression
-
-Vignette:
-https://cran.r-project.org/web/packages/reservoirnet/vignettes/basic_usage_01.html
-
-4) Study case: Prediction covid-19
+# Study case: Prediction covid-19
 
@@
 
-hospital is a central actor during pandemic, importance to anticipate high number of patients
+hospital is a central actor during pandemic, importance to anticipate
+high number of patients
 
-several models tried to forecast hospitalisations but task is difficult and many models struggled to perform well.
+several models tried to forecast hospitalisations but task is difficult
+and many models struggled to perform well.
 
-In this setting, reservoir computing seem to be an appropriate approach regarding the task.
+In this setting, reservoir computing seem to be an appropriate approach
+regarding the task.
 
@@
 
-This use case is based on data from Bordeaux University Hospital.
-During the pandemic a predictive model was elaborate and regularly use to forecast 14 days ahead the number of hospitalized patients at the hospital.
-In this work we will compare the first approach based on linear regression with elastic-net penalisation to reservoir computing approach.
+This use case is based on data from Bordeaux University Hospital. During
+the pandemic a predictive model was elaborate and regularly use to
+forecast 14 days ahead the number of hospitalized patients at the
+hospital. In this work we will compare the first approach based on
+linear regression with elastic-net penalisation to reservoir computing
+approach.
 
@@
 
-The task was to forecast 14 days ahead the number of hospitalised patients.
-The dataset was separated into two periods.
-First period from xx to xx served to identify relevant hyperparameters.
-Second period after xx was used to evaluate the model performance.
-As the time serie is not stationary and most appropriate hyperparameter value might change over time, we also investigated the effect of hyperparameter update every month.
+The task was to forecast 14 days ahead the number of hospitalised
+patients. The dataset was separated into two periods. First period from
+xx to xx served to identify relevant hyperparameters. Second period
+after xx was used to evaluate the model performance. As the time serie
+is not stationary and most appropriate hyperparameter value might change
+over time, we also investigated the effect of hyperparameter update
+every month.
 
-The performance of the model was evaluated according to several metrics: the median absolute error, the median relative error, the absolute error relative to baseline, the relative error relative to baseline.
-For each metric, the median was chosen over the mean to be robust to outliers.
-For instance, for relative error, when the outcome is low (e.g equal to 1 hospitalisation), an error of 5 hospitalisations could results to a 500% relative error.
+The performance of the model was evaluated according to several metrics:
+the median absolute error, the median relative error, the absolute error
+relative to baseline, the relative error relative to baseline. For each
+metric, the median was chosen over the mean to be robust to outliers.
+For instance, for relative error, when the outcome is low (e.g equal to
+1 hospitalisation), an error of 5 hospitalisations could results to a
+500% relative error.
 
-Median absolute error (AE): 
-Median relative error (RE):
+Median absolute error (AE): Median relative error (RE):
 
@@
 -   Recall hyperparameters
-- Present GA algorithm + explain why it is important here (large number of hp + assumption free compared to other algorithm, cons = expensive)
+-   Present GA algorithm + explain why it is important here (large
+    number of hp + assumption free compared to other algorithm, cons =
+    expensive)
 
@@
 <!-- * \proglang{Java} -->
+
 <!-- * \pkg{plyr} -->
@@
 <!-- ```{r} -->
+
 <!-- x <- 1:10 -->
+
 <!-- x -->
+
 <!-- ``` -->
@@
 -->
+
 <!-- * Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format. -->

### 2023-10-19 draft_jss/draft_jss.Rmd
clean tutorials parts

@@
 documentclass: jss
+bibliography: article.bib
 author:
@@
     orcid: 0000-0000-0000-0000
+    affiliation: 'Affiliation'
+    # To add another line, use \AND at the end of the previous one as above
+  - name: Third Author
+    orcid: 0000-0000-0000-0000
+    affiliation: 'Affiliation \AND'
+    # To add another line, use \AND at the end of the previous one as above
+  - name: Third Author
+    orcid: 0000-0000-0000-0000
+    affiliation: 'Affiliation'
+    # To add another line, use \AND at the end of the previous one as above
+  - name: Fourth Author
+    orcid: 0000-0000-0000-0000
     address: |
@@
 title:
-  formatted: "A Capitalized Title: Something about a Package \\pkg{foo}"
+  formatted: "Reservoir computing in R : a tutorial using \\pkg{reservoirnet}"
   # If you use tex in the formatted title, also supply version without
-  plain:     "A Capitalized Title: Something about a Package foo"
+  plain:     "Reservoir computing in R : a tutorial using reservoirnet"
   # For running headers, if needed
-  short:     "\\pkg{foo}: A Capitalized Title"
+  short:     "\\pkg{reservoirnet}: reservoir computing in R"
 abstract: >
@@
   rmarkdown::render(inputFile, encoding = encoding,
-  output_dir = here::here("output"), output_format = "all") })
+  output_dir = here::here("draft_jss"), output_format = "all") })
 ---
@@
 options(prompt = 'R> ', continue = '+ ')
+set.seed(1)
 ```
@@
 Only the output layer is optimized based on a ridge penalized linear regression."}
-knitr::include_graphics(here::here("reservoirnet : an R package for reservoir computing/images/schema.png/image1.png"))
+knitr::include_graphics("images/schema.png/image1.png")
 ```
@@
 
+Reservoir Computing (RC) is well suited to both regression and
+classification tasks. We will introduce a simple example for both task.
+
 ## Classification
 
-Reservoir Computing (RC) is well suited to both regression and classification tasks.
-We will introduce a simple example of classification task.
-This chapter is largely inspired from reservoirnet vignette on
-[classification](https://cran.r-project.org/web/packages/reservoirnet/vignettes/Classification_with_RC.html).
+This example is largely inspired from the [classification tutorial of reservoirpy](https://github.com/reservoirpy/reservoirpy/blob/master/tutorials/5-Classification-with-RC.ipynb).
+The first step is to load useful packages.
+For this example we will load both reservoirnet, ggplot2 [@wickham_ggplot2_2016] and dplyr
+[@wickham_dplyr_2023].
 
-The first step is to load useful packages. For this example we will load both reservoirnet and ggplot2 (add ref) a useful package for datavisualisation.
-
-```{r}
+```{r message=FALSE, warning=FALSE}
 library(reservoirnet)
 library(ggplot2)
+library(dplyr)
 ```
@@
 
-The Japanese vowel dataset is composed of 640 utterances of the Japanese vowel \ae, from 9 different male speakers.
-The goal of this task is to assign to each utterance the label of its speaker.
-Dataset is split between a 270 utterances training set and a 340 utterances testing set.
-Each spoken utterance is a timeseries of 7\~29 timesteps.
-Each timestep of signal is a 12 dimensional vector representing Linear Prediction Coefficient (LPC), which encode the audio signal into the cepstral domain (a variant of the frequency domain).
-
-References M. Kudo, J. Toyama and M. Shimbo. (1999). "Multidimensional
-Curve Classification Using Passing-Through Regions". Pattern Recognition
-Letters, Vol. 20, No. 11--13, pages 1103--1111.
-<https://archive.ics.uci.edu/ml/machine-learning-databases/JapaneseVowels-mld/>
-
-The data set is splitted into a train and a test set.
+To illustrate the classification task, we will use the Japanese vowel dataset (@kudo_multidimensional_1999). The data can be loaded from reservoirnet as follow :
 
 ```{r}
-japanese_vowels <- reservoirnet::generate_data(dataset = "japanese_vowels")$japanese_vowels
-
+japanese_vowels <- reservoirnet::generate_data(dataset = "japanese_vowels")[[1]]
 X_train <- japanese_vowels$X_train
@@
 
-```{r fig.cap="Vowel dataset"}
-sample_per_speaker <- 30
-n_speaker <- 9
-X_train_per_speaker <- list()
+The dataset is composed of 640 utterances of the Japanese vowel \ae, from 9 different speakers.
+The goal is to assign each uterance to the its speaker.
+Each utterances is a time series of 7 to 29 timesteps encoded by a 12 dimensional vector representing Linear Prediction Coefficient (LPC).
 
-for (i in 0:8) {
- X_speaker <- X_train[((i*sample_per_speaker)+1):((i+1)*sample_per_speaker)]
- X_train_per_speaker[[i+1]]<-(as.numeric(unlist(sapply(X_speaker, t))))
-}
+```{r fig.cap="Vowel dataset, sample with 3 speakers and 2 utterance each."}
+vec_sample <- c(1, 2, 41, 42, 71, 72)
+dfplot_vowel <- lapply(vec_sample,
+                 FUN = function(i){
+                   speaker <- which(Y_test[[i]] == 1)
+                   X_test[[i]] %>%
+                     as.data.frame() %>%
+                     tibble::rowid_to_column(var = "Time") %>%
+                     tidyr::pivot_longer(cols = -Time,
+                                         names_to = "component",
+                                         values_to = "LPC") %>%
+                     mutate(speaker = speaker, .before = 1,
+                            uterrance = i) %>%
+                     return()
+                 }) %>%
+  bind_rows()
 
-d <- data.frame(LPC = unlist(X_train_per_speaker), 
-                Speaker = factor(rep(1:9,times = sapply(X_train_per_speaker,length))))
+ggplot(dfplot_vowel, mapping = aes(x = Time, y = LPC, color = component)) +
+  geom_line() +
+  facet_wrap(uterrance ~ speaker,
+             labeller = label_bquote(cols = "speaker" : .(speaker)),
+             ncol = 2) +
+  theme_minimal() +
+  theme(legend.position = "none")
 
-# plot data
-ggplot(d,aes(x = Speaker, y = LPC)) + geom_boxplot()+theme_bw()
 ```
@@
 
-As ReservoirPy Nodes are built to work on sequences, the simplest setup
-to solve this task is sequence-to-sequence encoding, also called
-transduction. A model is trained on encoding each vector of input
-sequence into a new vector in the output space. Thus, a sequence of
-audio yields a sequence of label, one label per timestep.
+We can first solve this task using a sequence-to-sequence encoding approach. For this task, the goal is to predict the speaker for each time step of each utterance.
+
+The first step is to get the data where the label is repeated for each time step. This is easily done with the `repeat_targets` argument as follow :
 
@@
 
-#### Train a simple Echo State Network to solve this task:
+Then we can train a simple Echo State Network to solve this task. Here we will connect both the input layer and the reservoir layer to the readout layer.
 
@@
 
-#[source >> reservoir, source] >> readout
 model <- list(source %>>% reservoir, source) %>>% readout
@@
 
-#### Fit the model
+We can then fit the model and predict the labels for the test data :
 
@@
 
-Get the scores:
+From the `Y_pred` and `Y_test` we can represent graphically the forecast for some patients :
 
-There are 9 speakers, hence the output space is 9-dimensional. The
-speaker label is the index of the output neuron with maximum activation.
+```{r fig.cap="Prediction in a sequence-to-sequence approach 6 samples with 3 speakers and 2 utterance each. The higher the score of the speaker, the lighter the color.", fig.height=6}
+dfplotseqtoseq <- lapply(vec_sample,
+                 FUN = function(i){
+                   speaker <- which(Y_test[[i]][1,] == 1)
+                   Y_pred[[i]] %>%
+                     as.data.frame() %>%
+                     tibble::rowid_to_column(var = "Time") %>%
+                     tidyr::pivot_longer(cols = -Time,
+                                         names_to = "pred_speaker",
+                                         values_to = "prediction") %>%
+                     mutate(pred_speaker = gsub(x = pred_speaker,
+                                                pattern = "V", "")) %>%
+                     mutate(speaker = speaker, .before = 1,
+                            uterrance = i) %>%
+                     return()
+                 }) %>%
+  bind_rows()
+
+ggplot(dfplotseqtoseq, mapping = aes(x = Time,
+                                     y = pred_speaker,
+                                     fill = prediction)) +
+  geom_tile() +
+  facet_wrap(uterrance ~ speaker,
+             labeller = label_bquote(cols = "speaker" : .(speaker)),
+             ncol = 2) +
+  theme_minimal() +
+  labs(y = 'Predicted speaker',
+       fill = "Prediction score")
+```
+
+For those 6 utterances, the model correctly identify the speaker for most of the time steps. We can then evaluate the overall accuracy of the model :
 
 ```{r}
-accuracy <- function(pred, truth){
-  mean(pred == truth)
-}
-```
+accuracy <- function(pred, truth) mean(pred == truth)
 
-```{r}
-Y_pred_class <- sapply(Y_pred, FUN = function(x) apply(as.matrix(x),1,which.max))
-Y_test_class <- sapply(Y_test, FUN = function(x) apply(as.matrix(x),1,which.max))
+Y_pred_class <- sapply(Y_pred, FUN = function(x) apply(as.matrix(x),
+                                                       1,
+                                                       which.max))
+Y_test_class <- sapply(Y_test, FUN = function(x) apply(as.matrix(x),
+                                                       1,
+                                                       which.max))
 score <- accuracy(array(unlist(Y_pred_class)), array(unlist(Y_test_class)))
@@
 
-    ## [1] "Accuracy: 91.296%"
-
 ### Classification (sequence-to-vector model)
 
-We can create a more elaborated model where inference is performed only
-once on the whole input sequence. Indeed, we only need to assign one
-label to each input sequence. This new setup is known as a
-sequence-to-vector model, and this is usually the type of model we refer
-to when talking about classification of sequencial patterns.
+Another approach is the sequence-to-vector encoding.
+For this task we aim to predict the speaker of the whole utterance instead of the speaker of each time step (i.e the label is assign to the whole sequence).
 
@@
 
+In this example, we only connect the reservoir to the output layer.
+
 ```{r}
-source <- reservoirnet::createNode("Input")
+reservoir <- reservoirnet::createNode("Reservoir", units = 500, lr=0.1, sr=0.9)
 readout <- reservoirnet::createNode("Ridge",ridge=1e-6)
-reservoir <- reservoirnet::createNode("Reservoir", units = 500, lr=0.1, sr=0.9)
-
-#source >> reservoir >> readout
-model <- source %>>% reservoir %>>% readout
 ```
 
-We need to modify the training loop by hand a bit to perform this task:
-
-first, we compute all reservoir states over the input sequence using the
-reservoir.run method. then, we gather in a list only the last vector of
-the states sequence.
+To perform this task, we need to modify the training and testing process.
+Thanks to the echo property of RC, the information of previous time step is stored in the reservoir (i.e RC have a "memory").
+Therefore, the last vector of states accumulates information from all previous states.
+For this task of sequence-to-vector encoding we only use this last state.
+This performed as follow :
 
@@
 for (x in X_train) {
-  states <- reservoirnet::predict_seq(node = reservoir, X = x, reset=TRUE)
+  states <- reservoirnet::predict_seq(node = reservoir, X = x,
+                                      reset=TRUE)
   states_train[[k]] <- t(as.matrix(states[nrow(states),]))
@@
 
-We can now train the readout only on the last state vectors. Here,
-Y_train is an array storing a single label for each utterance.
+Then we can train the readout based on this last state vector. In that case, `Y_train`
+
+We can now train the readout only on the last state vectors. Here `Y_train` contains a single label for each utterance.
 
@@
 res <- reservoirnet::reservoirR_fit(readout,X = states_train, Y = Y_train)
-summary(res)
 ```
 
-    ## Parametrs using to fit:
-    ##  warmup: 0 ; stateful: FALSE ; reset: FALSE 
-    ## results of fitting:
-    ## 'Ridge-5': Ridge(ridge=1e-06, input_bias=True, in=500, out=9)
-
-We also modify the inference code using the same method as above:
+The prediction is also modified to use only the final state :
 
@@
 for (x in X_test) {
-  states <- reservoirnet::predict_seq(node = reservoir, X = x, reset=TRUE)
-  y <- reservoirnet::predict_seq(node = readout, X = as.array(states[nrow(states),]))
+  states <- reservoirnet::predict_seq(node = reservoir, X = x,
+                                      reset=TRUE)
+  y <- reservoirnet::predict_seq(node = readout,
+                                 X = as.array(states[nrow(states),]))
   Y_pred[[k]] <- y
@@
 
+```{r fig.cap="Prediction in a sequence-to-sequence approach 6 samples with 3 speakers and 2 utterance each. The higher the score of the speaker, the lighter the color."}
+dfplotseqtovec <- lapply(vec_sample,
+                 FUN = function(i){
+                   speaker <- which(Y_test[[i]][1,] == 1)
+                   Y_pred[[i]] %>%
+                     as.data.frame() %>%
+                     tidyr::pivot_longer(cols = everything(),
+                                         names_to = "pred_speaker",
+                                         values_to = "prediction") %>%
+                     mutate(pred_speaker = gsub(x = pred_speaker,
+                                                pattern = "V", "")) %>%
+                     mutate(speaker = speaker, .before = 1,
+                            uterrance = i,
+                            target = speaker == pred_speaker) %>%
+                     return()
+                 }) %>%
+  bind_rows()
+
+ggplot(dfplotseqtovec,
+       mapping = aes(x = pred_speaker,
+                     y = prediction,
+                     color = target)) +
+  geom_point() +
+  facet_wrap(uterrance ~ speaker,
+             labeller = label_bquote(cols = "speaker" : .(speaker)),
+             ncol = 2) +
+  scale_color_manual(values = c("darkgrey", "red")) +
+  theme_minimal() +
+  theme(legend.position = "none") +
+  labs(y = 'Score',
+       fill = "Speaker")
+```
+
+We observe, for those 6 uterrances that the model is correct for each of them. We can also compute the overall accuracy :
+
 ```{r}
-Y_pred_class <- sapply(Y_pred, FUN = function(x) apply(as.matrix(x),1,which.max))
-Y_test_class <- sapply(Y_test, FUN = function(x) apply(as.matrix(x),1,which.max))
+Y_pred_class <- sapply(Y_pred,
+                       FUN = function(x) apply(as.matrix(x),1,which.max))
+Y_test_class <- sapply(Y_test,
+                       FUN = function(x) apply(as.matrix(x),1,which.max))
 
-score <- accuracy(pred = Y_test_class,
-                       truth = Y_pred_class)
+score <- accuracy(pred = Y_test_class, truth = Y_pred_class)
 
@@
 
-    ## [1] "Accuracy: 88.108%"
-
 ## Regression
 
-Vignette:
-<https://cran.r-project.org/web/packages/reservoirnet/vignettes/basic_usage_01.html>
+### Goal
 
-# Study case: Prediction covid-19
+Here you will learn basic use of reservoirnet package. This will be
+illustrated with the covid-19 dataset included in the pakage. Those data
+contain the hospitalisations, positive RT-PCR and overall RT-PCR from
+Santé Publique France available on data.gouv.fr (see `help(dfCovid)` for
+more informations). Our goal will be to forecast the number of
+hospitalised patients 14 days ahead. To do so, we will first learn on
+data prior to 2022-01-01 and forecast on the following data.
+
+We can load the data and set the given task :
+
+```{r}
+data("dfCovid")
+dist_forecast = 14
+traintest_date = as.Date("2022-01-01")
+```
+
+### Covid-19 data
+
+Because both RT-PCR have strong variations, we will first proceed by
+taking the moving average on the last 7 days of those features. Also we
+add an outcome and an `outcomeDate` colum which will be useful to train
+the model. Furthermore we will compute the
+`outcome_deriv = outcome - hosp` which corresponds to the variation of
+hospitalisation compared to the current number of hospitalized patients.
+
+```{r}
+dfOutcome <- dfCovid %>%
+  # outcome at 14 days
+  mutate(outcome = lead(x = hosp, n = dist_forecast),
+         outcomeDate = date + dist_forecast,
+         outcome_deriv = outcome - hosp) %>%
+  # rolling average for tested and positive_pcr
+  mutate_at(.vars = c("Positive", "Tested"),
+            .funs = function(x) slider::slide_dbl(.x = x,
+                                                  .before = 6,
+                                                  .f = mean))
+```
+
+We can now plot the data :
+
+```{r fig.cap="Hospitalisations, IPTCC and positive PCR of Bordeaux University Hospital."}
+dfOutcome %>%
+  tidyr::pivot_longer(cols = c("hosp", "Positive", "Tested")) %>%
+  ggplot2::ggplot(mapping = aes(x = date, y = value)) +
+  geom_line() +
+  facet_grid(name ~ ., scales = "free_y") +
+  theme_bw() +
+  geom_vline(mapping = aes(color = "train-test sets",
+                           xintercept = traintest_date)) +
+  labs(color = "") +
+  theme(legend.position = "bottom")
+```
+
+### First reservoir
+
+Setting a reservoir is easily done with the `createNode()` function. The
+important hyperparameters are the following :
+
+-   Number of nodes (`units`) : it corresponds to the number of nodes
+    inside the reservoir. Usually, the more the better but more nodes
+    increase the computation time.
+-   Leaking rate (`lr`) : the leaking rate corresponds to the balance
+    between the new inputs and the previous state. A leaking rate of 1
+    only consider informations from the new inputs.
+-   Spectral radius (`sr`): the spectral radius is the maximum absolute
+    eigenvalue of the reservoir connectivity matrix. A small spectral
+    radius induces stable dynamics inside the reservoir, a high spectral
+    radius induces chaotic regimen inside the reservoir.
+-   Input scaling (`input_scaling`): the input scaling is a gain applied
+    to the input features of the reservoir.
+
+For this part of the tutorial, we will set the hyperparameter at a given
+value. Hyperparameter opitmisation will be detailed at section [Study
+case](#studycase).
+
+```{r}
+reservoir <- reservoirnet::createNode(nodeType = "Reservoir",
+                                      seed = 1,
+                                      units = 500,
+                                      lr = 0.7,
+                                      sr = 1,
+                                      input_scaling = 1)
+```
+
+Then we can feed the data to the reservoir and see the activation
+depending on time of the reservoir. To do so, we first prepare the data
+and transform it to an array.
+
+```{r}
+## select explanatory and transform it to an array
+X <- dfOutcome %>%
+  filter(outcomeDate < traintest_date) %>%
+  select(hosp, Positive, Tested) %>%
+  as.matrix() %>%
+  as.array()
+```
+
+Then we run the `predict_seq` function. The parameter reset is set to
+TRUE to be sure that the node is clear from data (it is optional here).
+
+```{r}
+reservoir_state <- predict_seq(node = reservoir, X = X, reset = TRUE)
+```
+
+Now we can visualise node activation using the `plot` function :
+
+```{r fig.cap="Node acivation over time."}
+plot(reservoir_state)
+```
+
+We see a lot of nodes with a steady state. It will be really difficult
+for the ridge output layer to learn on those nodes because they present
+no information. The problem is caused by the different scales of the
+features, to take this into account, we can divide them by their maximum
+value to scale each of them between `-1` and `1` :
+
+```{r}
+stand_max <- function(x) return(x/max(x))
+# scaled features
+Xstand <- dfOutcome %>%
+  filter(date < traintest_date) %>%
+  select(hosp, Positive, Tested) %>%
+  mutate_all(.funs = stand_max) %>%
+  as.matrix() %>%
+  as.array()
+```
+
+We then feed them to the reservoir and plot the node activation :
+
+```{r fig.cap="Node acivation over time. Scaled features"}
+# feed them to the reservoir
+reservoir_state_stand <- predict_seq(node = reservoir,
+                                     X = Xstand,
+                                     reset = TRUE)
+# plot the output
+plot(reservoir_state_stand)
+```
+
+The reservoir dynamics now look much better as no node seem to get
+saturated.
+
+### Forecast
+
+In order to train the reservoir, we should train the last layer which
+linearly combines the neuron's output.
+
+#### Set the ESN
+
+First we define this last layer with a ridge penalty of `1e3`. This
+hyperparameter can also be optimised which will be discussed at section
+[Study case](#studycase).
+
+```{r}
+readout <- reservoirnet::createNode(nodeType = "Ridge",
+                                    ridge = 1e3)
+```
+
+And we connect it to the previously defined reservoir :
+
+```{r}
+model <- reservoirnet::link(reservoir, readout)
+```
+
+The model is now ready we must now train the last layer.
+
+#### Set the data
+
+First we separate the train set on which we will learn the ridge
+coefficients and the test set on which we will make the forecast. We
+define the train set to be all the data before 2022-01-01 and the test
+data to be all the data to have forecast both on train and test set.
+
+```{r}
+# train set
+yTrain <- dfOutcome %>%
+  filter(outcomeDate <= traintest_date) %>%
+  select(outcome)
+yTrain_variation <- dfOutcome %>%
+  filter(outcomeDate <= traintest_date) %>%
+  select(outcome_deriv)
+xTrain <- dfOutcome %>%
+  filter(outcomeDate <= traintest_date) %>%
+  select(hosp, Positive, Tested)
+# test set
+xTest <- dfOutcome %>%
+  select(hosp, Positive, Tested)
+```
+
+We now standardise with the same formula as seen before. We learn the
+standardisation on the training set :
+
+```{r}
+# standardise based on training set values
+ls_fct_stand <- apply(xTrain,
+                      MARGIN = 2,
+                      FUN = function(x) function(feature) return(feature/(max(x))))
+```
+
+And we apply it to both the train and test sets :
+
+```{r results=FALSE}
+xTrainstand <- xTrain
+xTeststand <- xTest
+lapply(X = names(ls_fct_stand),
+       FUN = function(x){
+         xTrainstand[,x] <<- ls_fct_stand[[x]](feature = xTrain[,x])
+         xTeststand[,x] <<- ls_fct_stand[[x]](feature = xTest[,x])
+         return()
+       })
+```
+
+Finally we convert all those data to array :
+
+```{r}
+# convert to array
+lsdf <- lapply(list(yTrain = yTrain,
+                    yTrain_variation = yTrain_variation,
+                    xTrain = xTrainstand,
+                    xTest = xTeststand),
+               function(x) as.array(as.matrix(x)))
+```
+
+#### Train the model and predict
+
+Now, the easy part. We are going to train the reservoir with the train
+set. To do so, we set a warmup of 30 days during which the data are
+propagating into the reservoir but not used to fit the ridge layer.
+
+```{r}
+### train the reservoir ridge output
+fit <- reservoirnet::reservoirR_fit(node = model,
+                                    X = lsdf$xTrain,
+                                    Y = lsdf$yTrain,
+                                    warmup = 30,
+                                    reset = TRUE)
+```
+
+Now that the ridge layer is trained, we can perform predictions as seen
+before. We set the parameter `reset` to TRUE in order to clean the
+reservoir from the data used by the training set.
+
+```{r}
+### predict with the reservoir
+vec_pred <- reservoirnet::predict_seq(node = fit$fit,
+                                      X = lsdf$xTest,
+                                      reset = TRUE)
+```
+
+```{r fig.cap="Forecast"}
+dfOutcome %>%
+  mutate(pred = vec_pred) %>%
+  na.omit() %>%
+  ggplot(mapping = aes(x = outcomeDate)) +
+  geom_line(mapping = aes(y = outcome,
+                          color = "observed")) +
+  geom_line(mapping = aes(y = pred,
+                          color = "forecast")) +
+  geom_vline(mapping = aes(color = "train-test sets",
+                           xintercept = traintest_date)) +
+  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
+  theme_bw() +
+  labs(color = "", x = "Date", y = "Hospitalisations")
+```
+
+We observe that the model forecast is not fully accurate, both on the
+test set and the train set. In that case, one option could be to reduce
+ridge penalisation to fit more closely the data, the optimisation of
+ridge hyperparameter will be discussed at section [Study
+case](#studycase). Another possibility is to ease the learning of the
+algorithm by forecasting the variation of the hospitalisation instead of
+the number of hospitalised patients. For that step, we will learn on the
+`outcome_deriv` contained in `yTrain_variation` data which is defined
+outcome as `outcome_deriv = outcome - hosp`.
+
+```{r fig.cap="Covid-19 hospitalisations forecast.
+The model is either trained to forecast the number of hospitalisations (denoted Raw) or the variation of the hospitalisations compared to current level of hospitalisation (denoted Variation)"}
+## Fit reservoir on outcome variation instead of raw outcome
+fit2 <- reservoirnet::reservoirR_fit(node = model,
+                                     X = lsdf$xTrain,
+                                     Y = lsdf$yTrain_variation,
+                                     warmup = 30,
+                                     reset = TRUE)
+## Get the forecast on the test set
+vec_pred2_variation <- reservoirnet::predict_seq(node = fit2$fit,
+                                                 X = lsdf$xTest,
+                                                 reset = TRUE)
+## Transform the outome variation forecast into hospitalisation forecast
+vec_pred2 <- vec_pred2_variation + xTest$hosp
+## Plot the results
+dfOutcome %>%
+  mutate(Raw = vec_pred,
+         Variation = vec_pred2) %>%
+  tidyr::pivot_longer(cols = c(Raw, Variation),
+                      names_to = "Outcome_type",
+                      values_to = "Forecast") %>%
+  na.omit() %>%
+  ggplot(mapping = aes(x = outcomeDate)) +
+  geom_line(mapping = aes(y = outcome,
+                          color = "observed")) +
+  geom_line(mapping = aes(y = Forecast,
+                          color = "Forecast")) +
+  geom_vline(mapping = aes(color = "train-test sets",
+                           xintercept = traintest_date)) +
+  facet_wrap(Outcome_type ~ .,
+             labeller = label_bquote(cols = "Outcome" : .(Outcome_type))) +
+  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
+  theme_minimal() +
+  labs(color = "", x = "Date", y = "Hospitalisations")
+```
+
+We observe an improvement of the forecast compared to the forecast of
+hospitalisation previously discussed. From there, many improvement can
+be implemented including : frequent update of the model, leverage
+additional features and hyperparameter optimisation. This notions will
+be discussed in the next [section](#studycase).
+
+# Study case: Prediction covid-19 {#studycase}
 
@@
 
-4.1) Study case
+## Study case
 
@@
 
-4.2) Evaluation framework
+## Evaluation framework
 
--   Train and test set
+### Train and test set
 
@@
 
-4.3) Hyperparameter optimisation using GA
+## Hyperparameter optimisation using GA
 
@@
 
-4.4) Present results from GA
+## Present results from GA
 
@@
 
-5)  Discussion and conclusion
+## Present forecast performance
 
-<!-- ## Code formatting -->
+## Discussion on use case
 
-<!-- In general, don't use Markdown, but use the more precise LaTeX commands instead: -->
+# Previous work
 
-<!-- * \proglang{Java} -->
+Those work were done previously, the setting of train/test and
+hyperparameter optimisation were different and probably they should be
+done again in a similar setting for relevant comparison.
 
-<!-- * \pkg{plyr} -->
+## Raw outcome, first derivative or second derivative ?
 
-<!-- One exception is inline code, which can be written inside a pair of backticks (i.e., using the Markdown syntax). -->
+See
+<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/explore_esn.md>
 
-<!-- If you want to use LaTeX commands in headers, you need to provide a `short-title` attribute.
-You can also provide a custom identifier if necessary.
-See the header of Section \ref{r-code} for example.
--->
+## Raw feature or scaled feature ?
 
-<!-- # \proglang{R} code {short-title="R code" #r-code} -->
+See
+<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/explore_esn.md>
 
-<!-- Can be inserted in regular R markdown blocks. -->
+## Smoothed or not smoothed
 
-<!-- ```{r} -->
+See
+<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/explore_esn.md>
 
-<!-- x <- 1:10 -->
+## Number of nodes
 
-<!-- x -->
+See
+<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/explore_esn_r_pipeline.md>
 
-<!-- ``` -->
+## Use of second derivative among the interesting features
 
-<!-- ## Features specific to \pkg{rticles} {short-title="Features specific to rticles"} -->
+See
+<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/tune_esn2.md>
 
-<!-- * Adding short titles to section headers is a feature specific to \pkg{rticles} (implemented via a Pandoc Lua filter).
-This feature is currently not supported by Pandoc and we will update this template if [it is officially supported in the future](https://github.com/jgm/pandoc/issues/4409).
--->
+## Results of random sampling with R
 
-<!-- * Using the `\AND` syntax in the `author` field to add authors on a new line. This is a specific to the `rticles::jss_article` format. -->
+See
+<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/meeting_1801/master_report.md>
+
+# Discussion and conclusion
+
+# Bibliography

### 2023-10-20 draft_jss/draft_jss.Rmd
update results

@@
 
+# Basic package use
+
+In this section, we will cover the basics of reservoirnet use including
+installation, classification and regression. We will provide brief
+overview of the package use and more in depth description is provided in
+section 4 with the covid-19 forecast use case.
+
+## Installation
+
+reservoirnet is an R package api making the python module reservoirPy
+easily callable from R. It is available on CRAN (see
+<https://cran.r-project.org/package=reservoirnet>) and can be installed
+using:
+
+```{r eval=FALSE}
+install.packages("reservoirnet")
+```
+
+Reservoir Computing (RC) is well suited to both regression and
+classification tasks. We will introduce a simple example for both task.
+
+## Regression
+
+### Covid-19 data
+
+In this first use case, we will introduce the fundamental usage of the `reservoirnet` package.
+This demonstration will be conducted using the COVID-19 dataset that is included within the package.
+These data encompass hospitalization figures, positive RT-PCR results, and overall RT-PCR data sourced from Santé Publique France, which are publicly available on data.gouv.fr (for further details, refer to the `help(dfCovid)` function).
+Our primary objective is to predict the number of hospitalized patients 14 days into the future.
+To accomplish this, we will initially train our model on data preceding the date of January 1, 2022, and subsequently apply it to forecast values using the subsequent dataset.
+
+We can proceed by loading useful packages (i.e ggplot2 [@wickham_ggplot2_2016] and dplyr [@wickham_dplyr_2023]), data and define the task:
+
+```{r message=FALSE, warning=FALSE}
+library(dplyr)
+library(ggplot2)
+library(reservoirnet)
+
+data("dfCovid")
+dist_forecast = 14
+traintest_date = as.Date("2022-01-01")
+```
+
+Due to the substantial fluctuations observed in both RT-PCR metrics, our initial step involves applying a moving average computation over the most recent 7-day periods for these features.
+Additionally, we augment the dataset by introducing an `outcome` column and an `outcomeDate` column, which will serve as valuable inputs for model training.
+Moreover, we calculate the `outcome_deriv` as the difference between the outcome and the number of hospitalized patients (`hosp`), representing the variation in hospitalization in relation to the current count of hospitalized individuals.
+The resulting smoothed data is visualized in Figure @ref(fig:covidintro).
+
+```{r}
+dfOutcome <- dfCovid %>%
+  # outcome at 14 days
+  mutate(outcome = lead(x = hosp, n = dist_forecast),
+         outcomeDate = date + dist_forecast,
+         outcome_deriv = outcome - hosp) %>%
+  # rolling average for tested and positive_pcr
+  mutate_at(.vars = c("Positive", "Tested"),
+            .funs = function(x) slider::slide_dbl(.x = x,
+                                                  .before = 6,
+                                                  .f = mean))
+```
+
+We can now plot the data :
+
+```{r covidintro, fig.cap="Hospitalisations, IPTCC and positive PCR of Bordeaux University Hospital.", echo = FALSE}
+dfOutcome %>%
+  tidyr::pivot_longer(cols = c("hosp", "Positive", "Tested")) %>%
+  ggplot2::ggplot(mapping = aes(x = date, y = value)) +
+  geom_line() +
+  facet_grid(name ~ ., scales = "free_y") +
+  theme_bw() +
+  geom_vline(mapping = aes(color = "train-test sets",
+                           xintercept = traintest_date)) +
+  labs(color = "") +
+  theme(legend.position = "bottom")
+```
+
+### First reservoir
+
+Setting a reservoir is done with the `createNode()` function. The important hyperparameters are the following :
+
+-   Number of nodes (`units`) : it corresponds to the number of nodes inside the reservoir. Usually, the more the better but more nodes increase the computation time.
+-   Leaking rate (`lr`) : the leaking rate corresponds to the balance between the new inputs and the previous state. A leaking rate of 1 only consider informations from the new inputs.
+-   Spectral radius (`sr`): the spectral radius is the maximum absolute eigenvalue of the reservoir connectivity matrix.
+A small spectral radius induces stable dynamics inside the reservoir, a high spectral radius induces chaotic regimen inside the reservoir.
+-   Input scaling (`input_scaling`): the input scaling is a gain applied to the input features of the reservoir.
+-   Seed (`seed`): because the reservoir connections are set at random, setting the seed is a good approach to ensure   reproducibility.
+Another approach would be to use several different reservoir but it increases computation time.
+
+For this part of the tutorial, we will set the hyperparameter at a given value. Hyperparameter opitmisation will be detailed at section [Study case](#studycase).
+
+```{r}
+reservoir <- reservoirnet::createNode(nodeType = "Reservoir",
+                                      seed = 1,
+                                      units = 500,
+                                      lr = 0.7,
+                                      sr = 1,
+                                      input_scaling = 1)
+```
+
+Then we can feed the data to the reservoir and see the activation depending on time of the reservoir. To do so, we first prepare the data and transform it to a matrix :
+
+```{r}
+## select explanatory and transform it to an array
+X <- dfOutcome %>%
+  filter(outcomeDate < traintest_date) %>%
+  select(hosp, Positive, Tested) %>%
+  as.matrix()
+```
+
+Then we run the `predict_seq` function. It takes as input a node (i.e a reservoir or a reservoir associated with an output layer) and the feature matrix.
+
+```{r}
+reservoir_state <- predict_seq(node = reservoir, X = X)
+```
+
+Now we can visualise node activation using the `plot` function presented at figure \@ref(fig:nodeactivationbad).
+
+```{r nodeactivationbad, fig.cap="Node acivation over time."}
+plot(reservoir_state)
+```
+
+Numerous nodes within the system exhibit a consistent equilibrium state.
+The challenge arises when the ridge output layer attempts to acquire knowledge from these nodes, as they do not convey meaningful information.
+This issue can be attributed to the disparate scales of the features.
+To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby rescaling them within the range of `-1` to `1`.
+
+```{r}
+stand_max <- function(x) return(x/max(x))
+# scaled features
+Xstand <- dfOutcome %>%
+  filter(date < traintest_date) %>%
+  select(hosp, Positive, Tested) %>%
+  mutate_all(.funs = stand_max) %>%
+  as.matrix() %>%
+  as.array()
+```
+
+We then feed them to the reservoir and plot the node activation again.
+Compared to \@ref(fig:nodeactivationbad), the obtained node activation \@ref(fig:nodeactivationgood) shows interesting trend outputs as no node seems saturated.
+
+```{r nodeactivationgood, fig.cap="Node acivation over time. Scaled features"}
+# feed them to the reservoir
+reservoir_state_stand <- predict_seq(node = reservoir,
+                                     X = Xstand,
+                                     reset = TRUE)
+# plot the output
+plot(reservoir_state_stand)
+```
+
+### Forecast
+
+In order to train the reservoir, we should train the last layer which
+linearly combines the neuron's output.
+
+#### Set the ESN
+
+Initially, we establish the output layer, incorporating a ridge penalty set at `1e3`.
+It's important to note that this hyperparameter can be subject to optimization, a topic that will be explored in the forthcoming [Study Case](#studycase) section.
+This parameter plays a pivotal role in fine-tuning the model's conformity to the dataset.
+When set excessively high, the risk of underfitting arises, whereas setting it too low can lead to overfitting.
+Then we connect the output layer to the reservoir making the model ready to be trained.
+
+```{r}
+readout <- reservoirnet::createNode(nodeType = "Ridge",
+                                    ridge = 1e3)
+model <- reservoirnet::link(reservoir, readout)
+```
+
+#### Set the data
+
+First we separate the train set on which we will learn the ridge coefficients and the test set on which we will make the forecast.
+We define the train set to be all the data before 2022-01-01 and the test data to be all the data to have forecast both on train and test sets.
+
+```{r}
+# train set
+dftrain <- dfOutcome %>% filter(outcomeDate <= traintest_date)
+yTrain <- dftrain %>% select(outcome)
+yTrain_variation <- dftrain %>% select(outcome_deriv)
+xTrain <- dftrain %>% select(hosp, Positive, Tested)
+# test set
+xTest <- dfOutcome %>% select(hosp, Positive, Tested)
+```
+
+We now standardise with the same formula as seen before. We learn the standardisation on the training set and apply it on the test set. Then we convert the dataframe to matrix.
+
+```{r results=FALSE}
+# copy train and test sets
+xTrainstand <- xTrain
+xTeststand <- xTest
+# standardise based on training set values
+ls_fct_stand <- apply(xTrain,
+                      MARGIN = 2,
+                      FUN = function(x) function(feature) return(feature/(max(x))))
+lapply(X = names(ls_fct_stand),
+       FUN = function(x){
+         xTrainstand[,x] <<- ls_fct_stand[[x]](feature = xTrain[,x])
+         xTeststand[,x] <<- ls_fct_stand[[x]](feature = xTest[,x])
+         return()
+       })
+# convert to array
+lsdf <- lapply(list(yTrain = yTrain,
+                    yTrain_variation = yTrain_variation,
+                    xTrain = xTrainstand,
+                    xTest = xTeststand),
+               function(x) as.matrix(x))
+```
+
+#### Train the model and predict
+
+We then feed the reservoir with the train set. To do so, we set a `warmup` of `30` days during which the data are propagating into the reservoir but not used to fit the output layer.
+
+```{r}
+### train the reservoir ridge output
+fit <- reservoirnet::reservoirR_fit(node = model,
+                                    X = lsdf$xTrain,
+                                    Y = lsdf$yTrain,
+                                    warmup = 30,
+                                    reset = TRUE)
+```
+
+Now that the ridge layer is trained, we can forecast. We set the parameter `reset` to `TRUE` in order to clean the reservoir from the data used by the training set.
+
+```{r}
+vec_pred <- reservoirnet::predict_seq(node = fit$fit,
+                                      X = lsdf$xTest,
+                                      reset = TRUE)
+```
+
+```{r fig.cap="Forecast"}
+dfOutcome %>%
+  mutate(pred = vec_pred) %>%
+  na.omit() %>%
+  ggplot(mapping = aes(x = outcomeDate)) +
+  geom_line(mapping = aes(y = outcome,
+                          color = "observed")) +
+  geom_line(mapping = aes(y = pred,
+                          color = "forecast")) +
+  geom_vline(mapping = aes(color = "train-test sets",
+                           xintercept = traintest_date)) +
+  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
+  theme_bw() +
+  labs(color = "", x = "Date", y = "Hospitalisations")
+```
+
+We observe that the model forecast is not fully accurate, both on the test set and the train set.
+In that case, one option could be to reduce ridge penalisation to fit more closely the data, the optimisation of ridge hyperparameter will be discussed at section [Study case](#studycase).
+Another possibility is to ease the learning of the algorithm by forecasting the variation of the hospitalisation instead of the number of hospitalised patients.
+For that step, we will learn on the `outcome_deriv` contained in `yTrain_variation` data which is defined outcome as `outcome_deriv = outcome - hosp`.
+
+```{r fig.cap="Covid-19 hospitalisations forecast.
+The model is either trained to forecast the number of hospitalisations (denoted Raw) or the variation of the hospitalisations compared to current level of hospitalisation (denoted Variation)"}
+## Fit reservoir on outcome variation instead of raw outcome
+fit2 <- reservoirnet::reservoirR_fit(node = model,
+                                     X = lsdf$xTrain,
+                                     Y = lsdf$yTrain_variation,
+                                     warmup = 30,
+                                     reset = TRUE)
+## Get the forecast on the test set
+vec_pred2_variation <- reservoirnet::predict_seq(node = fit2$fit,
+                                                 X = lsdf$xTest,
+                                                 reset = TRUE)
+## Transform the outome variation forecast into hospitalisation forecast
+vec_pred2 <- vec_pred2_variation + xTest$hosp
+## Plot the results
+dfOutcome %>%
+  mutate(Raw = vec_pred,
+         Variation = vec_pred2) %>%
+  tidyr::pivot_longer(cols = c(Raw, Variation),
+                      names_to = "Outcome_type",
+                      values_to = "Forecast") %>%
+  na.omit() %>%
+  ggplot(mapping = aes(x = outcomeDate)) +
+  geom_line(mapping = aes(y = outcome,
+                          color = "observed")) +
+  geom_line(mapping = aes(y = Forecast,
+                          color = "Forecast")) +
+  geom_vline(mapping = aes(color = "train-test sets",
+                           xintercept = traintest_date)) +
+  facet_wrap(Outcome_type ~ .,
+             labeller = label_bquote(cols = "Outcome" : .(Outcome_type))) +
+  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
+  theme_minimal() +
+  labs(color = "", x = "Date", y = "Hospitalisations")
+```
+
+We observe an improvement of the forecast compared to the forecast of hospitalisation previously discussed.
+From there, many improvement can be implemented including : frequent update of the model, leverage additional features and hyperparameter optimisation.
+This notions will be discussed in the [Study case section](#studycase).
+
+## Classification
+
+### The Japanese vowel dataset
+
+This example is largely inspired from the [classification tutorial of reservoirpy](https://github.com/reservoirpy/reservoirpy/blob/master/tutorials/5-Classification-with-RC.ipynb).
+To illustrate the classification task, we will use the Japanese vowel dataset (@kudo_multidimensional_1999).
+The data can be loaded from reservoirnet as follow :
+
+```{r}
+japanese_vowels <- reservoirnet::generate_data(dataset = "japanese_vowels")[[1]]
+X_train <- japanese_vowels$X_train
+Y_train <- japanese_vowels$Y_train
+X_test <- japanese_vowels$X_test
+Y_test <- japanese_vowels$Y_test
+```
+
+The dataset comprises 640 vocalizations of the Japanese vowel \ae, contributed by nine distinct speakers.
+Each vocalization represents a time series spanning between 7 and 29 timesteps, encoded as a 12-dimensional vector denoting the Linear Prediction Coefficients (LPC).
+A visual representation of six distinct utterances from the test set, originating from three different speakers, is depicted in Figure \@ref(fig:vowelpresentation).
+
+```{r vowelpresentation, fig.cap="Vowel dataset, sample with 3 speakers and 2 utterance each.", echo = FALSE}
+vec_sample <- c(1, 2, 41, 42, 71, 72)
+dfplot_vowel <- lapply(vec_sample,
+                 FUN = function(i){
+                   speaker <- which(Y_test[[i]] == 1)
+                   X_test[[i]] %>%
+                     as.data.frame() %>%
+                     tibble::rowid_to_column(var = "Time") %>%
+                     tidyr::pivot_longer(cols = -Time,
+                                         names_to = "component",
+                                         values_to = "LPC") %>%
+                     mutate(speaker = speaker, .before = 1,
+                            uterrance = i) %>%
+                     return()
+                 }) %>%
+  bind_rows()
+
+ggplot(dfplot_vowel, mapping = aes(x = Time, y = LPC, color = component)) +
+  geom_line() +
+  facet_wrap(uterrance ~ speaker,
+             labeller = label_bquote(cols = "speaker" : .(speaker)),
+             ncol = 2) +
+  theme_minimal() +
+  theme(legend.position = "none")
+
+```
+
+The primary objective involves the attribution of each utterance to its respective speaker, this is denoted as classification or sequence-to-vector encoding.
+The secondary objective involves the attribution of each time step of each utterance to its speaker, this is denoted as transduction or sequence-to-sequence encoding.
+
+### Classification (sequence-to-vector model)
+
+The first approach is the sequence-to-vector encoding.
+For this task we aim to predict the speaker of the whole utterance (i.e the label is assigned to the whole sequence).
+We first start by creating the reservoir and the output layer.
+
+```{r}
+reservoir <- reservoirnet::createNode("Reservoir", units = 500,
+                                      lr=0.1, sr=0.9,
+                                      seed = 1)
+readout <- reservoirnet::createNode("Ridge",ridge=1e-6)
+```
+
+To perform this task, we need to modify the training and testing process.
+Leveraging the inherent echo property of the RC, information from preceding time steps is preserved within the reservoir, effectively endowing the RC with a form of memory.
+Consequently, the final state vector encapsulates insights gathered from all antecedent states.
+In the context of the sequence-to-vector encoding task, only this ultimate state is employed.
+This process is executed as follows:
+
+```{r}
+states_train = list()
+k <- 1
+for (x in X_train) {
+  states <- reservoirnet::predict_seq(node = reservoir, X = x,
+                                      reset=TRUE)
+  states_train[[k]] <- t(as.matrix(states[nrow(states),]))
+  k <- k+1
+}
+```
+
+Then we can train the readout based on this last state vector. In that case, `Y_train` contains a single label for each utterance.
+
+```{r}
+res <- reservoirnet::reservoirR_fit(readout,X = states_train, Y = Y_train)
+```
+
+The prediction is also modified using only the final state :
+
+```{r}
+Y_pred <- list()
+k <- 1
+for (x in X_test) {
+  states <- reservoirnet::predict_seq(node = reservoir, X = x,
+                                      reset=TRUE)
+  y <- reservoirnet::predict_seq(node = readout,
+                                 X = as.array(states[nrow(states),]))
+  Y_pred[[k]] <- y
+  k <- k+1
+}
+```
+
+Figure \@ref(fig:seqtovec) shows the prediction for the 6 uterrances depicted at figure \@ref(fig:vowelpresentation) where the model correctly identifies the speaker.
+
+```{r seqtovec, fig.cap="Prediction in a sequence-to-sequence approach 6 samples with 3 speakers and 2 utterance each.
+The speaker to predict is depicted in red.
+For each of the 6 uterrance, the model correctly identifies the speaker."}
+dfplotseqtovec <- lapply(vec_sample,
+                 FUN = function(i){
+                   speaker <- which(Y_test[[i]][1,] == 1)
+                   Y_pred[[i]] %>%
+                     as.data.frame() %>%
+                     tidyr::pivot_longer(cols = everything(),
+                                         names_to = "pred_speaker",
+                                         values_to = "prediction") %>%
+                     mutate(pred_speaker = gsub(x = pred_speaker,
+                                                pattern = "V", "")) %>%
+                     mutate(speaker = speaker, .before = 1,
+                            uterrance = i,
+                            target = speaker == pred_speaker) %>%
+                     return()
+                 }) %>%
+  bind_rows()
+
+ggplot(dfplotseqtovec,
+       mapping = aes(x = pred_speaker,
+                     y = prediction,
+                     color = target)) +
+  geom_point() +
+  facet_wrap(uterrance ~ speaker,
+             labeller = label_bquote(cols = "speaker" : .(speaker)),
+             ncol = 2) +
+  scale_color_manual(values = c("darkgrey", "red")) +
+  theme_minimal() +
+  theme(legend.position = "none") +
+  labs(y = 'Score',
+       x = "Speaker")
+```
+
+Then, we can also compute the overall accuracy :
+
+```{r}
+accuracy <- function(pred, truth) mean(pred == truth)
+
+Y_pred_class <- sapply(Y_pred,
+                       FUN = function(x) apply(as.matrix(x),1,which.max))
+Y_test_class <- sapply(Y_test,
+                       FUN = function(x) apply(as.matrix(x),1,which.max))
+
+score <- accuracy(pred = Y_test_class, truth = Y_pred_class)
+
+print(paste0("Accuracy: ", round(score * 100,3) ,"%"))
+```
+
+### Transduction (sequence-to-sequence model)
+
+For this task, the goal is to predict the speaker for each time step of each utterance.
+The first step is to get the data where the label is repeated for each time step.
+This is easily done with the `repeat_targets` argument as follow :
+
+```{r}
+japanese_vowels <- reservoirnet::generate_data(
+    dataset = "japanese_vowels",
+    repeat_targets=TRUE)$japanese_vowels
+X_train <- japanese_vowels$X_train
+Y_train <- japanese_vowels$Y_train
+X_test <- japanese_vowels$X_test
+Y_test <- japanese_vowels$Y_test
+```
+
+Then we can train a simple Echo State Network to solve this task.
+For this example we will connect both the input layer and the reservoir layer to the readout layer which is performed by the `%>>%` operator :
+
+```{r}
+source <- createNode("Input")
+readout <- createNode("Ridge",ridge=1e-6)
+reservoir <- createNode("Reservoir",units = 500,lr=0.1, sr=0.9, seed = 1)
+
+model <- list(source %>>% reservoir, source) %>>% readout
+```
+
+We can then fit the model and predict the labels for the test data. The `reset` parameter is set to `TRUE` to remove information from the reservoir from the training process.
+
+```{r}
+model_fit <- reservoirnet::reservoirR_fit(node = model,
+                                          X = X_train,
+                                          Y = Y_train,
+                                          warmup = 2)
+
+Y_pred <- reservoirnet::predict_seq(node = model_fit$fit,
+                                    X = X_test,
+                                    reset = TRUE)
+```
+
+From the `Y_pred` and `Y_test` we represent at figure \@ref(fig:figseqtoseq) the predictions for the same patients as in figure \@ref(fig:vowelpresentation).
+
+```{r figseqtoseq, fig.cap="Prediction in a sequence-to-sequence approach 6 samples with 3 speakers and 2 utterance each. The higher the score of the speaker, the lighter the color.", fig.height=6}
+dfplotseqtoseq <- lapply(vec_sample,
+                 FUN = function(i){
+                   speaker <- which(Y_test[[i]][1,] == 1)
+                   Y_pred[[i]] %>%
+                     as.data.frame() %>%
+                     tibble::rowid_to_column(var = "Time") %>%
+                     tidyr::pivot_longer(cols = -Time,
+                                         names_to = "pred_speaker",
+                                         values_to = "prediction") %>%
+                     mutate(pred_speaker = gsub(x = pred_speaker,
+                                                pattern = "V", ""),
+                            speaker = speaker,
+                            uterrance = i,
+                            .before = 1) %>%
+                     return()
+                 }) %>%
+  bind_rows()
+
+ggplot(dfplotseqtoseq, mapping = aes(x = Time,
+                                     y = pred_speaker,
+                                     fill = prediction)) +
+  geom_tile() +
+  facet_wrap(uterrance ~ speaker,
+             labeller = label_bquote(cols = "speaker" : .(speaker)),
+             ncol = 2) +
+  theme_minimal() +
+  labs(y = 'Predicted speaker',
+       fill = "Prediction score")
+```
+
+For those 6 utterances, the model correctly identify the speaker for most of the time steps. We can then evaluate the overall accuracy of the model :
+
+```{r}
+Y_pred_class <- sapply(Y_pred, FUN = function(x) apply(as.matrix(x),
+                                                       1,
+                                                       which.max))
+Y_test_class <- sapply(Y_test, FUN = function(x) apply(as.matrix(x),
+                                                       1,
+                                                       which.max))
+score <- accuracy(array(unlist(Y_pred_class)), array(unlist(Y_test_class)))
+
+print(paste0("Accuracy: ", round(score * 100,3) ,"%"))
+```
+
+# Study case: Covid-19 hospitalisations forecast {#studycase}
+
+## Introduction
+
+Since late 2020, millions of cases of SARS-CoV-2 infection have been documented across the globe [@carrat_evidence_2021, @world_health_organisation_who_2020, @noauthor_estimating_2022].
+This ongoing pandemic has exerted significant strain on healthcare systems, resulting in a surge in hospitalizations.
+This surge, in turn, necessitated modifications to the healthcare infrastructure and gave rise to unprecedented population-wide lockdown measures aimed at preventing the saturation of healthcare facilities [@kim_health_2020, @simoes_organisation_2021, @hubner_surgery_2020].
+The capacity to predict the trajectory of the epidemic on a regional scale is of paramount importance for effective healthcare system management.
+
+Numerous COVID-19 forecasting algorithms have been proposed using different methods (e.g ensemble, deep learning, compartmental), yet none has proven entirely satisfactory [@rahimi_review_2021, @cramer_evaluation_2022].
+In France, short-term forecasts  with different methods have been evaluated with similar results [@pottier_forecast_2021, @paireau_ensemble_2022, @carvalho_analysis_2021, @mohimont_convolutional_2021].
+In this context a machine learning algorithm based on linear regression with elastic-net penalisation, leveraging both EHRs and public data, was implemented at Bordeaux University Hospital [@ferte_benefit_2022].
+This model, which aimed at forecasting the number of hospitalised patients at 14 days, showed good performance but struggled to accurately anticipate dynamic shifts of the epidemic.
+
+RC has been used in the context of covid-19 epidemic forecast [@ghosh_reservoir_2021, @kmet_bezier_2019, @liu_nanophotonic_2023, @ray_optimized_2021, @zhang_sentiment_2023].
+Among them, @ghosh_reservoir_2021, @liu_nanophotonic_2023 and @ray_optimized_2021 used it to forecast epidemic (@zhang_sentiment_2023 performed sentiment analysis and @kmet_bezier_2019 used it to solve optimal control related to vaccine).
+The evaluation of RC for epidemic forecast showed promising results in all approaches, being competitive with Long-Short Term Memory (LSTM) and Feed-Forward Neural Network (FFNN) in @ray_optimized_2021.
+However, the test period was short for @ghosh_reservoir_2021 (21 and 14 days) and @ray_optimized_2021 (86 days) making it difficult to evaluate the behavior of the methods during epidemic dynamic shift.
+This was not the case for @liu_nanophotonic_2023 (6 months) but they implemented daily ahead forecast which would be difficult to use to manage a hospital.
+Finally, all three implementations used only one time series as input whereas it has been shown that using different data sources could improve forecast [@ferte_benefit_2022].
+Therefore, it is still difficult to assess the usefulness of RC over a large period and using many time series as inputs.
+
+`TODO :` add paragraph about RC in high dimension
+
+`TODO :` add paragraph about hp optimisation for RC (focus on GA)
+
+RC can be approached as a complexification of penalized linear regression where inputs are processed by a reservoir which allows memory and non-linear combinations.
+As penalised linear regression was the best approach for covid-19 forecast in @ferte_benefit_2022 and that RC showed promising results for epidemic forecast in @ghosh_reservoir_2021, @liu_nanophotonic_2023 and @ray_optimized_2021, we proposed to use RC to forecast hospitalisations at 14 days at the University Hospital of Bordeaux.
+The main objective of this study is to evaluate the performance of RC for this task.
+Secondary objective will be (i) to assess the performance of RC depending on the input dimension, (ii) to assess the relevance of Genetic Algorithm (GA) for hyperparameter optimisation and feature selection in this context.
+
+## Methods
+
+### Data source
+
+Aggregated data from May 16, 2020, to January 17, 2022, regarding French COVID-19 epidemic were included. In order to improve forecasting, several data sources were used.
+
+#### Open data
+
+Open data included both epidemiologic data from Santé Publique France and weather data from National Oceanic and Atmospheric Administration (NOAA) Integrated Surface Database [@etalab_les_2020, @smith_integrated_2011].
+Both provide department aggregated data and are daily updated.
+
+Santé Publique France data included hospitalizations, number of RT-PCR, positive RT-PCR, proportion of positive RT-PCR, dominant variant, and number of first dose vaccinated.
+RT-PCR data were available by age and were grouped as 0–19, 20–59, and 60 and more years old categories.
+Variant identification data before February 18, 2021 were not available, and majority variant before that date was assigned to wild type.
+
+NOAA data, including temperature, wind speed, humidity, and dew point, were extracted and the Predict Index for COVID-19 Climate Transmissibility—Index PREDICT de Transmissivité Climatique de la COVID-19—(IPTCC) was computed [].
+Missing weather data were imputed using a 2-step procedure: (1) the mean value of the adjacent department was imputed; (2) remaining missing values were imputed using last observation carried forward.
+
+#### EHR data
+
+The Bordeaux Hospital is a large structure comprising 3 hospital structures taking care of nearly 250 000 hospitalized patients and 100 000 emergency consultations during 2020.16 A data warehouse based on i2b2 structure was built in 2017.13 This star architecture is based on a central fact table where each row represents a diagnosis, a laboratory result, a procedure, a medical observation, etc.
+Each fact is related to other tables with information about the patient, the visit, the provider, or type of fact.17 This structure allows for quick data queries compared to the usual siloed organizations.
+
+To perform those queries, ontology alignment have been performed on laboratory results, and ad hoc natural language processing tools have been developed, including ROMEDI (ie, a French drug terminology to extract drug information from text), IAMsystem (ie, a dictionary-based approach for name entity recognition), and SmartCRF (ie, a software to visualize and annotate EHR).18–20 Several applied projects were performed using those tools including automatic detection of surgical site infection and transfusion associated circulatory overload.21,22
+
+Thanks to those previous experiences, the Bordeaux hospital data warehouse was used, during the pandemic, to describe the current state of the epidemic at the hospital level on a daily basis.
+Those data were then used in the forecast model including: hospitalizations, hospital and ICU admission and discharge, ambulance service notes, and emergency unit notes.
+Concepts related to COVID-19 were extracted from notes by dictionary-based approaches (eg, cough, dyspnea, COVID-19).
+Dictionaries were manually created based on manual chart review to identify terms used by practitioners.
+Then, the number and proportion of ambulance service calls or hospitalization in emergency units mentioning concepts related to COVID-19 were extracted.
+Detail of features is available in Supplementary Table S1.
+
+Due to different data acquisition mechanisms, there was a delay between the occurrence of events and the data acquisition.
+It was of 1 day for EHR data, 5 days for department hospitalizations and RT-PCR, 4 days for weather, 2 days for variants, and 4 days for vaccination.
+For the training and evaluation of the model, the chosen date was the date of data availability to mimic a real-time streaming forecast.
+
+### Evaluation framework
+
+### Train and test set
+
+The task was to forecast 14 days ahead the number of hospitalised
+patients. The dataset was separated into two periods. First period from
+xx to xx served to identify relevant hyperparameters. Second period
+after xx was used to evaluate the model performance. As the time serie
+is not stationary and most appropriate hyperparameter value might change
+over time, we also investigated the effect of hyperparameter update
+every month.
+
+The performance of the model was evaluated according to several metrics:
+the median absolute error, the median relative error, the absolute error
+relative to baseline, the relative error relative to baseline. For each
+metric, the median was chosen over the mean to be robust to outliers.
+For instance, for relative error, when the outcome is low (e.g equal to
+1 hospitalisation), an error of 5 hospitalisations could results to a
+500% relative error.
+
+Median absolute error (AE): Median relative error (RE):
+
+-   Baseline model (forecast at T+14 = observed at T)
+-   Metrics
+
+### Hyperparameter optimisation using GA
+
+-   Recall hyperparameters
+-   Present GA algorithm + explain why it is important here (large
+    number of hp + assumption free compared to other algorithm, cons =
+    expensive)
+
 To tackle both problems, we proposed a genetic algorithm for
@@
 
-# Basic package use
+## Results
 
-In this section, we will cover the basics of reservoirnet use including
-installation, classification and regression. We will provide brief
-overview of the package use and more in depth description is provided in
-section 4 with the covid-19 forecast use case.
-
-## Installation
-
-reservoirnet is an R package api making the python module reservoirPy
-easily callable from R. It is available on CRAN (see
-<https://cran.r-project.org/package=reservoirnet>) and can be installed
-using:
-
-```{r eval=FALSE}
-install.packages("reservoirnet")
-```
-
-Reservoir Computing (RC) is well suited to both regression and
-classification tasks. We will introduce a simple example for both task.
-
-## Classification
-
-This example is largely inspired from the [classification tutorial of reservoirpy](https://github.com/reservoirpy/reservoirpy/blob/master/tutorials/5-Classification-with-RC.ipynb).
-The first step is to load useful packages.
-For this example we will load both reservoirnet, ggplot2 [@wickham_ggplot2_2016] and dplyr
-[@wickham_dplyr_2023].
-
-```{r message=FALSE, warning=FALSE}
-library(reservoirnet)
-library(ggplot2)
-library(dplyr)
-```
-
-### The Japanese vowel dataset
-
-To illustrate the classification task, we will use the Japanese vowel dataset (@kudo_multidimensional_1999). The data can be loaded from reservoirnet as follow :
-
-```{r}
-japanese_vowels <- reservoirnet::generate_data(dataset = "japanese_vowels")[[1]]
-X_train <- japanese_vowels$X_train
-Y_train <- japanese_vowels$Y_train
-X_test <- japanese_vowels$X_test
-Y_test <- japanese_vowels$Y_test
-```
-
-The dataset is composed of 640 utterances of the Japanese vowel \ae, from 9 different speakers.
-The goal is to assign each uterance to the its speaker.
-Each utterances is a time series of 7 to 29 timesteps encoded by a 12 dimensional vector representing Linear Prediction Coefficient (LPC).
-
-```{r fig.cap="Vowel dataset, sample with 3 speakers and 2 utterance each."}
-vec_sample <- c(1, 2, 41, 42, 71, 72)
-dfplot_vowel <- lapply(vec_sample,
-                 FUN = function(i){
-                   speaker <- which(Y_test[[i]] == 1)
-                   X_test[[i]] %>%
-                     as.data.frame() %>%
-                     tibble::rowid_to_column(var = "Time") %>%
-                     tidyr::pivot_longer(cols = -Time,
-                                         names_to = "component",
-                                         values_to = "LPC") %>%
-                     mutate(speaker = speaker, .before = 1,
-                            uterrance = i) %>%
-                     return()
-                 }) %>%
-  bind_rows()
-
-ggplot(dfplot_vowel, mapping = aes(x = Time, y = LPC, color = component)) +
-  geom_line() +
-  facet_wrap(uterrance ~ speaker,
-             labeller = label_bquote(cols = "speaker" : .(speaker)),
-             ncol = 2) +
-  theme_minimal() +
-  theme(legend.position = "none")
-
-```
-
-### Echo state network setup
-
-#### Transduction (sequence-to-sequence model)
-
-We can first solve this task using a sequence-to-sequence encoding approach. For this task, the goal is to predict the speaker for each time step of each utterance.
-
-The first step is to get the data where the label is repeated for each time step. This is easily done with the `repeat_targets` argument as follow :
-
-```{r}
-japanese_vowels <- reservoirnet::generate_data(
-    dataset = "japanese_vowels",
-    repeat_targets=TRUE)$japanese_vowels
-
-X_train <- japanese_vowels$X_train
-Y_train <- japanese_vowels$Y_train
-X_test <- japanese_vowels$X_test
-Y_test <- japanese_vowels$Y_test
-```
-
-Then we can train a simple Echo State Network to solve this task. Here we will connect both the input layer and the reservoir layer to the readout layer.
-
-```{r}
-source <- createNode("Input")
-readout <- createNode("Ridge",ridge=1e-6)
-reservoir <- createNode("Reservoir",units = 500,lr=0.1, sr=0.9)
-
-model <- list(source %>>% reservoir, source) %>>% readout
-```
-
-We can then fit the model and predict the labels for the test data :
-
-```{r}
-model_fit <- reservoirnet::reservoirR_fit(node = model,
-                                          X = X_train,
-                                          Y = Y_train,
-                                          stateful = FALSE,
-                                          warmup = 2)
-
-Y_pred <- reservoirnet::predict_seq(node = model_fit$fit,
-                                    X = X_test,
-                                    stateful = FALSE)
-```
-
-From the `Y_pred` and `Y_test` we can represent graphically the forecast for some patients :
-
-```{r fig.cap="Prediction in a sequence-to-sequence approach 6 samples with 3 speakers and 2 utterance each. The higher the score of the speaker, the lighter the color.", fig.height=6}
-dfplotseqtoseq <- lapply(vec_sample,
-                 FUN = function(i){
-                   speaker <- which(Y_test[[i]][1,] == 1)
-                   Y_pred[[i]] %>%
-                     as.data.frame() %>%
-                     tibble::rowid_to_column(var = "Time") %>%
-                     tidyr::pivot_longer(cols = -Time,
-                                         names_to = "pred_speaker",
-                                         values_to = "prediction") %>%
-                     mutate(pred_speaker = gsub(x = pred_speaker,
-                                                pattern = "V", "")) %>%
-                     mutate(speaker = speaker, .before = 1,
-                            uterrance = i) %>%
-                     return()
-                 }) %>%
-  bind_rows()
-
-ggplot(dfplotseqtoseq, mapping = aes(x = Time,
-                                     y = pred_speaker,
-                                     fill = prediction)) +
-  geom_tile() +
-  facet_wrap(uterrance ~ speaker,
-             labeller = label_bquote(cols = "speaker" : .(speaker)),
-             ncol = 2) +
-  theme_minimal() +
-  labs(y = 'Predicted speaker',
-       fill = "Prediction score")
-```
-
-For those 6 utterances, the model correctly identify the speaker for most of the time steps. We can then evaluate the overall accuracy of the model :
-
-```{r}
-accuracy <- function(pred, truth) mean(pred == truth)
-
-Y_pred_class <- sapply(Y_pred, FUN = function(x) apply(as.matrix(x),
-                                                       1,
-                                                       which.max))
-Y_test_class <- sapply(Y_test, FUN = function(x) apply(as.matrix(x),
-                                                       1,
-                                                       which.max))
-score <- accuracy(array(unlist(Y_pred_class)), array(unlist(Y_test_class)))
-
-print(paste0("Accuracy: ", round(score * 100,3) ,"%"))
-```
-
-### Classification (sequence-to-vector model)
-
-Another approach is the sequence-to-vector encoding.
-For this task we aim to predict the speaker of the whole utterance instead of the speaker of each time step (i.e the label is assign to the whole sequence).
-
-```{r}
-japanese_vowels <- reservoirnet::generate_data(
-    dataset = "japanese_vowels")$japanese_vowels
-
-X_train <- japanese_vowels$X_train
-Y_train <- japanese_vowels$Y_train
-X_test <- japanese_vowels$X_test
-Y_test <- japanese_vowels$Y_test
-```
-
-In this example, we only connect the reservoir to the output layer.
-
-```{r}
-reservoir <- reservoirnet::createNode("Reservoir", units = 500, lr=0.1, sr=0.9)
-readout <- reservoirnet::createNode("Ridge",ridge=1e-6)
-```
-
-To perform this task, we need to modify the training and testing process.
-Thanks to the echo property of RC, the information of previous time step is stored in the reservoir (i.e RC have a "memory").
-Therefore, the last vector of states accumulates information from all previous states.
-For this task of sequence-to-vector encoding we only use this last state.
-This performed as follow :
-
-```{r}
-states_train = list()
-k <- 1
-for (x in X_train) {
-  states <- reservoirnet::predict_seq(node = reservoir, X = x,
-                                      reset=TRUE)
-  states_train[[k]] <- t(as.matrix(states[nrow(states),]))
-  k <- k+1
-}
-```
-
-Then we can train the readout based on this last state vector. In that case, `Y_train`
-
-We can now train the readout only on the last state vectors. Here `Y_train` contains a single label for each utterance.
-
-```{r}
-res <- reservoirnet::reservoirR_fit(readout,X = states_train, Y = Y_train)
-```
-
-The prediction is also modified to use only the final state :
-
-```{r}
-Y_pred <- list()
-k <- 1
-for (x in X_test) {
-  states <- reservoirnet::predict_seq(node = reservoir, X = x,
-                                      reset=TRUE)
-  y <- reservoirnet::predict_seq(node = readout,
-                                 X = as.array(states[nrow(states),]))
-  Y_pred[[k]] <- y
-  k <- k+1
-}
-```
-
-```{r fig.cap="Prediction in a sequence-to-sequence approach 6 samples with 3 speakers and 2 utterance each. The higher the score of the speaker, the lighter the color."}
-dfplotseqtovec <- lapply(vec_sample,
-                 FUN = function(i){
-                   speaker <- which(Y_test[[i]][1,] == 1)
-                   Y_pred[[i]] %>%
-                     as.data.frame() %>%
-                     tidyr::pivot_longer(cols = everything(),
-                                         names_to = "pred_speaker",
-                                         values_to = "prediction") %>%
-                     mutate(pred_speaker = gsub(x = pred_speaker,
-                                                pattern = "V", "")) %>%
-                     mutate(speaker = speaker, .before = 1,
-                            uterrance = i,
-                            target = speaker == pred_speaker) %>%
-                     return()
-                 }) %>%
-  bind_rows()
-
-ggplot(dfplotseqtovec,
-       mapping = aes(x = pred_speaker,
-                     y = prediction,
-                     color = target)) +
-  geom_point() +
-  facet_wrap(uterrance ~ speaker,
-             labeller = label_bquote(cols = "speaker" : .(speaker)),
-             ncol = 2) +
-  scale_color_manual(values = c("darkgrey", "red")) +
-  theme_minimal() +
-  theme(legend.position = "none") +
-  labs(y = 'Score',
-       fill = "Speaker")
-```
-
-We observe, for those 6 uterrances that the model is correct for each of them. We can also compute the overall accuracy :
-
-```{r}
-Y_pred_class <- sapply(Y_pred,
-                       FUN = function(x) apply(as.matrix(x),1,which.max))
-Y_test_class <- sapply(Y_test,
-                       FUN = function(x) apply(as.matrix(x),1,which.max))
-
-score <- accuracy(pred = Y_test_class, truth = Y_pred_class)
-
-print(paste0("Accuracy: ", round(score * 100,3) ,"%"))
-```
-
-## Regression
-
-### Goal
-
-Here you will learn basic use of reservoirnet package. This will be
-illustrated with the covid-19 dataset included in the pakage. Those data
-contain the hospitalisations, positive RT-PCR and overall RT-PCR from
-Santé Publique France available on data.gouv.fr (see `help(dfCovid)` for
-more informations). Our goal will be to forecast the number of
-hospitalised patients 14 days ahead. To do so, we will first learn on
-data prior to 2022-01-01 and forecast on the following data.
-
-We can load the data and set the given task :
-
-```{r}
-data("dfCovid")
-dist_forecast = 14
-traintest_date = as.Date("2022-01-01")
-```
-
-### Covid-19 data
-
-Because both RT-PCR have strong variations, we will first proceed by
-taking the moving average on the last 7 days of those features. Also we
-add an outcome and an `outcomeDate` colum which will be useful to train
-the model. Furthermore we will compute the
-`outcome_deriv = outcome - hosp` which corresponds to the variation of
-hospitalisation compared to the current number of hospitalized patients.
-
-```{r}
-dfOutcome <- dfCovid %>%
-  # outcome at 14 days
-  mutate(outcome = lead(x = hosp, n = dist_forecast),
-         outcomeDate = date + dist_forecast,
-         outcome_deriv = outcome - hosp) %>%
-  # rolling average for tested and positive_pcr
-  mutate_at(.vars = c("Positive", "Tested"),
-            .funs = function(x) slider::slide_dbl(.x = x,
-                                                  .before = 6,
-                                                  .f = mean))
-```
-
-We can now plot the data :
-
-```{r fig.cap="Hospitalisations, IPTCC and positive PCR of Bordeaux University Hospital."}
-dfOutcome %>%
-  tidyr::pivot_longer(cols = c("hosp", "Positive", "Tested")) %>%
-  ggplot2::ggplot(mapping = aes(x = date, y = value)) +
-  geom_line() +
-  facet_grid(name ~ ., scales = "free_y") +
-  theme_bw() +
-  geom_vline(mapping = aes(color = "train-test sets",
-                           xintercept = traintest_date)) +
-  labs(color = "") +
-  theme(legend.position = "bottom")
-```
-
-### First reservoir
-
-Setting a reservoir is easily done with the `createNode()` function. The
-important hyperparameters are the following :
-
--   Number of nodes (`units`) : it corresponds to the number of nodes
-    inside the reservoir. Usually, the more the better but more nodes
-    increase the computation time.
--   Leaking rate (`lr`) : the leaking rate corresponds to the balance
-    between the new inputs and the previous state. A leaking rate of 1
-    only consider informations from the new inputs.
--   Spectral radius (`sr`): the spectral radius is the maximum absolute
-    eigenvalue of the reservoir connectivity matrix. A small spectral
-    radius induces stable dynamics inside the reservoir, a high spectral
-    radius induces chaotic regimen inside the reservoir.
--   Input scaling (`input_scaling`): the input scaling is a gain applied
-    to the input features of the reservoir.
-
-For this part of the tutorial, we will set the hyperparameter at a given
-value. Hyperparameter opitmisation will be detailed at section [Study
-case](#studycase).
-
-```{r}
-reservoir <- reservoirnet::createNode(nodeType = "Reservoir",
-                                      seed = 1,
-                                      units = 500,
-                                      lr = 0.7,
-                                      sr = 1,
-                                      input_scaling = 1)
-```
-
-Then we can feed the data to the reservoir and see the activation
-depending on time of the reservoir. To do so, we first prepare the data
-and transform it to an array.
-
-```{r}
-## select explanatory and transform it to an array
-X <- dfOutcome %>%
-  filter(outcomeDate < traintest_date) %>%
-  select(hosp, Positive, Tested) %>%
-  as.matrix() %>%
-  as.array()
-```
-
-Then we run the `predict_seq` function. The parameter reset is set to
-TRUE to be sure that the node is clear from data (it is optional here).
-
-```{r}
-reservoir_state <- predict_seq(node = reservoir, X = X, reset = TRUE)
-```
-
-Now we can visualise node activation using the `plot` function :
-
-```{r fig.cap="Node acivation over time."}
-plot(reservoir_state)
-```
-
-We see a lot of nodes with a steady state. It will be really difficult
-for the ridge output layer to learn on those nodes because they present
-no information. The problem is caused by the different scales of the
-features, to take this into account, we can divide them by their maximum
-value to scale each of them between `-1` and `1` :
-
-```{r}
-stand_max <- function(x) return(x/max(x))
-# scaled features
-Xstand <- dfOutcome %>%
-  filter(date < traintest_date) %>%
-  select(hosp, Positive, Tested) %>%
-  mutate_all(.funs = stand_max) %>%
-  as.matrix() %>%
-  as.array()
-```
-
-We then feed them to the reservoir and plot the node activation :
-
-```{r fig.cap="Node acivation over time. Scaled features"}
-# feed them to the reservoir
-reservoir_state_stand <- predict_seq(node = reservoir,
-                                     X = Xstand,
-                                     reset = TRUE)
-# plot the output
-plot(reservoir_state_stand)
-```
-
-The reservoir dynamics now look much better as no node seem to get
-saturated.
-
-### Forecast
-
-In order to train the reservoir, we should train the last layer which
-linearly combines the neuron's output.
-
-#### Set the ESN
-
-First we define this last layer with a ridge penalty of `1e3`. This
-hyperparameter can also be optimised which will be discussed at section
-[Study case](#studycase).
-
-```{r}
-readout <- reservoirnet::createNode(nodeType = "Ridge",
-                                    ridge = 1e3)
-```
-
-And we connect it to the previously defined reservoir :
-
-```{r}
-model <- reservoirnet::link(reservoir, readout)
-```
-
-The model is now ready we must now train the last layer.
-
-#### Set the data
-
-First we separate the train set on which we will learn the ridge
-coefficients and the test set on which we will make the forecast. We
-define the train set to be all the data before 2022-01-01 and the test
-data to be all the data to have forecast both on train and test set.
-
-```{r}
-# train set
-yTrain <- dfOutcome %>%
-  filter(outcomeDate <= traintest_date) %>%
-  select(outcome)
-yTrain_variation <- dfOutcome %>%
-  filter(outcomeDate <= traintest_date) %>%
-  select(outcome_deriv)
-xTrain <- dfOutcome %>%
-  filter(outcomeDate <= traintest_date) %>%
-  select(hosp, Positive, Tested)
-# test set
-xTest <- dfOutcome %>%
-  select(hosp, Positive, Tested)
-```
-
-We now standardise with the same formula as seen before. We learn the
-standardisation on the training set :
-
-```{r}
-# standardise based on training set values
-ls_fct_stand <- apply(xTrain,
-                      MARGIN = 2,
-                      FUN = function(x) function(feature) return(feature/(max(x))))
-```
-
-And we apply it to both the train and test sets :
-
-```{r results=FALSE}
-xTrainstand <- xTrain
-xTeststand <- xTest
-lapply(X = names(ls_fct_stand),
-       FUN = function(x){
-         xTrainstand[,x] <<- ls_fct_stand[[x]](feature = xTrain[,x])
-         xTeststand[,x] <<- ls_fct_stand[[x]](feature = xTest[,x])
-         return()
-       })
-```
-
-Finally we convert all those data to array :
-
-```{r}
-# convert to array
-lsdf <- lapply(list(yTrain = yTrain,
-                    yTrain_variation = yTrain_variation,
-                    xTrain = xTrainstand,
-                    xTest = xTeststand),
-               function(x) as.array(as.matrix(x)))
-```
-
-#### Train the model and predict
-
-Now, the easy part. We are going to train the reservoir with the train
-set. To do so, we set a warmup of 30 days during which the data are
-propagating into the reservoir but not used to fit the ridge layer.
-
-```{r}
-### train the reservoir ridge output
-fit <- reservoirnet::reservoirR_fit(node = model,
-                                    X = lsdf$xTrain,
-                                    Y = lsdf$yTrain,
-                                    warmup = 30,
-                                    reset = TRUE)
-```
-
-Now that the ridge layer is trained, we can perform predictions as seen
-before. We set the parameter `reset` to TRUE in order to clean the
-reservoir from the data used by the training set.
-
-```{r}
-### predict with the reservoir
-vec_pred <- reservoirnet::predict_seq(node = fit$fit,
-                                      X = lsdf$xTest,
-                                      reset = TRUE)
-```
-
-```{r fig.cap="Forecast"}
-dfOutcome %>%
-  mutate(pred = vec_pred) %>%
-  na.omit() %>%
-  ggplot(mapping = aes(x = outcomeDate)) +
-  geom_line(mapping = aes(y = outcome,
-                          color = "observed")) +
-  geom_line(mapping = aes(y = pred,
-                          color = "forecast")) +
-  geom_vline(mapping = aes(color = "train-test sets",
-                           xintercept = traintest_date)) +
-  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
-  theme_bw() +
-  labs(color = "", x = "Date", y = "Hospitalisations")
-```
-
-We observe that the model forecast is not fully accurate, both on the
-test set and the train set. In that case, one option could be to reduce
-ridge penalisation to fit more closely the data, the optimisation of
-ridge hyperparameter will be discussed at section [Study
-case](#studycase). Another possibility is to ease the learning of the
-algorithm by forecasting the variation of the hospitalisation instead of
-the number of hospitalised patients. For that step, we will learn on the
-`outcome_deriv` contained in `yTrain_variation` data which is defined
-outcome as `outcome_deriv = outcome - hosp`.
-
-```{r fig.cap="Covid-19 hospitalisations forecast.
-The model is either trained to forecast the number of hospitalisations (denoted Raw) or the variation of the hospitalisations compared to current level of hospitalisation (denoted Variation)"}
-## Fit reservoir on outcome variation instead of raw outcome
-fit2 <- reservoirnet::reservoirR_fit(node = model,
-                                     X = lsdf$xTrain,
-                                     Y = lsdf$yTrain_variation,
-                                     warmup = 30,
-                                     reset = TRUE)
-## Get the forecast on the test set
-vec_pred2_variation <- reservoirnet::predict_seq(node = fit2$fit,
-                                                 X = lsdf$xTest,
-                                                 reset = TRUE)
-## Transform the outome variation forecast into hospitalisation forecast
-vec_pred2 <- vec_pred2_variation + xTest$hosp
-## Plot the results
-dfOutcome %>%
-  mutate(Raw = vec_pred,
-         Variation = vec_pred2) %>%
-  tidyr::pivot_longer(cols = c(Raw, Variation),
-                      names_to = "Outcome_type",
-                      values_to = "Forecast") %>%
-  na.omit() %>%
-  ggplot(mapping = aes(x = outcomeDate)) +
-  geom_line(mapping = aes(y = outcome,
-                          color = "observed")) +
-  geom_line(mapping = aes(y = Forecast,
-                          color = "Forecast")) +
-  geom_vline(mapping = aes(color = "train-test sets",
-                           xintercept = traintest_date)) +
-  facet_wrap(Outcome_type ~ .,
-             labeller = label_bquote(cols = "Outcome" : .(Outcome_type))) +
-  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
-  theme_minimal() +
-  labs(color = "", x = "Date", y = "Hospitalisations")
-```
-
-We observe an improvement of the forecast compared to the forecast of
-hospitalisation previously discussed. From there, many improvement can
-be implemented including : frequent update of the model, leverage
-additional features and hyperparameter optimisation. This notions will
-be discussed in the next [section](#studycase).
-
-# Study case: Prediction covid-19 {#studycase}
-
-what is covid-19 and covid-19 pandemic
-
-hospital is a central actor during pandemic, importance to anticipate
-high number of patients
-
-several models tried to forecast hospitalisations but task is difficult
-and many models struggled to perform well.
-
-In this setting, reservoir computing seem to be an appropriate approach
-regarding the task.
-
-## Study case
-
-This use case is based on data from Bordeaux University Hospital. During
-the pandemic a predictive model was elaborate and regularly use to
-forecast 14 days ahead the number of hospitalized patients at the
-hospital. In this work we will compare the first approach based on
-linear regression with elastic-net penalisation to reservoir computing
-approach.
-
-## Evaluation framework
-
-### Train and test set
-
-The task was to forecast 14 days ahead the number of hospitalised
-patients. The dataset was separated into two periods. First period from
-xx to xx served to identify relevant hyperparameters. Second period
-after xx was used to evaluate the model performance. As the time serie
-is not stationary and most appropriate hyperparameter value might change
-over time, we also investigated the effect of hyperparameter update
-every month.
-
-The performance of the model was evaluated according to several metrics:
-the median absolute error, the median relative error, the absolute error
-relative to baseline, the relative error relative to baseline. For each
-metric, the median was chosen over the mean to be robust to outliers.
-For instance, for relative error, when the outcome is low (e.g equal to
-1 hospitalisation), an error of 5 hospitalisations could results to a
-500% relative error.
-
-Median absolute error (AE): Median relative error (RE):
-
--   Baseline model (forecast at T+14 = observed at T)
--   Metrics
-
-## Hyperparameter optimisation using GA
-
--   Recall hyperparameters
--   Present GA algorithm + explain why it is important here (large
-    number of hp + assumption free compared to other algorithm, cons =
-    expensive)
-
-## Present results from GA
+### Present results from GA
 
@@
 
-## Present forecast performance
+### Present forecast performance
 
-## Discussion on use case
+## Discussion
 

### 2023-10-23 draft_jss/draft_jss.Rmd
update jss

@@
 
-$$x( t+1 ) = ( 1 - \alpha )  x ( t) + \alpha \: tanh( W x(t) + W_{in} u(t+1) ) \text{ where } \alpha \in [0, 1 ]$$
+$$x( t+1 ) = ( 1 - \alpha )  x ( t) + \alpha \: tanh( W x(t) + W_{in} u(t+1) )", "ext{ where } \alpha \in [0, 1 ]$$
 
@@
 (i.e $x(t)$) and the new inputs (i.e $u(t+1)$). Both $W_{in}$ and $W$
-are random matrices of size $Nres \times M$ and $Nres \times Nres$
+are random matrices of size $Nres", "imes M$ and $Nres", "imes Nres$
 respectively.
@@
 
-## Regression
+## Regression {#basicregression}
 
@@
                      y = prediction,
-                     color = target)) +
-  geom_point() +
+                     fill = target)) +
+  geom_bar(stat = "identity") +
   facet_wrap(uterrance ~ speaker,
@@
              ncol = 2) +
-  scale_color_manual(values = c("darkgrey", "red")) +
+  scale_fill_manual(values = c("darkgrey", "darkred")) +
   theme_minimal() +
@@
 
-### Data source
+### Data
 
-Aggregated data from May 16, 2020, to January 17, 2022, regarding French COVID-19 epidemic were included. In order to improve forecasting, several data sources were used.
+The study utilized aggregated data spanning from May 16, 2020, to January 17, 2022, regarding the COVID-19 epidemic in France, drawing from various sources to enhance forecasting accuracy.
+These sources encompassed epidemiological statistics from Santé Publique France, weather data from the National Oceanic and Atmospheric Administration (NOAA), both providing department-level data [@etalab_les_2020, @smith_integrated_2011] and Electronic Health Record (EHR) data from the Bordeaux Hospital providing hospital-level data.
+All data were daily updated.
+Santé Publique France data included information on hospitalizations, RT-PCR tests, positive RT-PCR results, variant prevalence, and vaccination data, categorized by age groups.
+NOAA data contributed temperature, wind speed, humidity, and dew point data, allowing for the computation of the COVID-19 Climate Transmissibility Predict Index [@roumagnac_etude_2021].
+EHRs data included hospitalizations, ICU admissions, ambulance service records, and emergency unit notes, with relevant COVID-19-related concepts extracted from the notes.
+Data are discussed more in depth in @ferte_benefit_2022.
 
-#### Open data
-
-Open data included both epidemiologic data from Santé Publique France and weather data from National Oceanic and Atmospheric Administration (NOAA) Integrated Surface Database [@etalab_les_2020, @smith_integrated_2011].
-Both provide department aggregated data and are daily updated.
-
-Santé Publique France data included hospitalizations, number of RT-PCR, positive RT-PCR, proportion of positive RT-PCR, dominant variant, and number of first dose vaccinated.
-RT-PCR data were available by age and were grouped as 0–19, 20–59, and 60 and more years old categories.
-Variant identification data before February 18, 2021 were not available, and majority variant before that date was assigned to wild type.
-
-NOAA data, including temperature, wind speed, humidity, and dew point, were extracted and the Predict Index for COVID-19 Climate Transmissibility—Index PREDICT de Transmissivité Climatique de la COVID-19—(IPTCC) was computed [].
-Missing weather data were imputed using a 2-step procedure: (1) the mean value of the adjacent department was imputed; (2) remaining missing values were imputed using last observation carried forward.
-
-#### EHR data
-
-The Bordeaux Hospital is a large structure comprising 3 hospital structures taking care of nearly 250 000 hospitalized patients and 100 000 emergency consultations during 2020.16 A data warehouse based on i2b2 structure was built in 2017.13 This star architecture is based on a central fact table where each row represents a diagnosis, a laboratory result, a procedure, a medical observation, etc.
-Each fact is related to other tables with information about the patient, the visit, the provider, or type of fact.17 This structure allows for quick data queries compared to the usual siloed organizations.
-
-To perform those queries, ontology alignment have been performed on laboratory results, and ad hoc natural language processing tools have been developed, including ROMEDI (ie, a French drug terminology to extract drug information from text), IAMsystem (ie, a dictionary-based approach for name entity recognition), and SmartCRF (ie, a software to visualize and annotate EHR).18–20 Several applied projects were performed using those tools including automatic detection of surgical site infection and transfusion associated circulatory overload.21,22
-
-Thanks to those previous experiences, the Bordeaux hospital data warehouse was used, during the pandemic, to describe the current state of the epidemic at the hospital level on a daily basis.
-Those data were then used in the forecast model including: hospitalizations, hospital and ICU admission and discharge, ambulance service notes, and emergency unit notes.
-Concepts related to COVID-19 were extracted from notes by dictionary-based approaches (eg, cough, dyspnea, COVID-19).
-Dictionaries were manually created based on manual chart review to identify terms used by practitioners.
-Then, the number and proportion of ambulance service calls or hospitalization in emergency units mentioning concepts related to COVID-19 were extracted.
-Detail of features is available in Supplementary Table S1.
-
-Due to different data acquisition mechanisms, there was a delay between the occurrence of events and the data acquisition.
-It was of 1 day for EHR data, 5 days for department hospitalizations and RT-PCR, 4 days for weather, 2 days for variants, and 4 days for vaccination.
-For the training and evaluation of the model, the chosen date was the date of data availability to mimic a real-time streaming forecast.
+First and second derivative over the last $7$ days were computed to enrich model information.
+To take into account measurement error and daily noise variation, data were smoothed using a local polynomial regression with a span of 21 days.
 
@@
 
-### Train and test set
+The task was to forecast 14 days ahead the number of hospitalised patients.
+As seen at section [Regression](#basicregression), we will train the model to predict the variation of hospitalisation defined as $outcome_{t+14} = hosp_{t+14} - hosp_{t}$.
+Metrics computation and visualisations will be performed on the predicted number of hospitalizations denoted as $\hat{y_{t+14}} = \widehat{outcome_{t+14}} + hospitalizations_{t}$ and the observed number of hospitalisations at time $t+14$ denoted as $y_{t+14} = hospitalizations_{t+14}$
 
-The task was to forecast 14 days ahead the number of hospitalised
-patients. The dataset was separated into two periods. First period from
-xx to xx served to identify relevant hyperparameters. Second period
-after xx was used to evaluate the model performance. As the time serie
-is not stationary and most appropriate hyperparameter value might change
-over time, we also investigated the effect of hyperparameter update
-every month.
+The dataset was separated into two periods.
+First period from May 16, 2020 to March 1, 2021 served to identify relevant hyperparameters.
+Remaining data was used to evaluate the model performance.
+As the time serie is not stationary and most appropriate hyperparameter value might change over time, we also investigated the effect of hyperparameter update every month.
 
-The performance of the model was evaluated according to several metrics:
-the median absolute error, the median relative error, the absolute error
-relative to baseline, the relative error relative to baseline. For each
-metric, the median was chosen over the mean to be robust to outliers.
-For instance, for relative error, when the outcome is low (e.g equal to
-1 hospitalisation), an error of 5 hospitalisations could results to a
-500% relative error.
-
-Median absolute error (AE): Median relative error (RE):
-
--   Baseline model (forecast at T+14 = observed at T)
--   Metrics
+The performance of the model was evaluated according to several metrics: the median absolute error ($MAE = median(|\hat{y_t}-y_t|)$) , the median relative error ($MRE = median(|\frac{\hat{y_t}-y_t}{y_t}|)$), the median absolute error to baseline ($MAEB =  median(|\hat{y_t}-y_t| - |y_{t-14} - y_t|)$), the median relative error to baseline ($MREB = median(\frac{|\hat{y_t}-y_t|}{|y_{t-14} - y_t|})$).
+Such that the model denoted as baseline was the model defined as $\hat{y_t} = y_{t-14}$.
+All those metrics reflects different information.
+$MAE$ and $MAEB$ are highly sensitive to forecast error when the number of hospitalizations is high.
+$MAEB$ encapsulates information about the current dynamic of the hospitalizations : for the same absolute error, absolute error to baseline will be lower if the number of hospitalization is not constant.
+**MAEB = redundant with MAE ?**.
+$MRE$ is highly sensitive to forecast error when the number of hospitalizations is low because the denominator will be small.
+$MREB$ is highly sensitive to small forecast error when the number of hospitalized patients is constant over time.
+For each metric, the median was chosen over the mean to be robust to outliers.
+For instance, for relative error, when the outcome is low (e.g equal to 1 hospitalization), an error of 5 hospitalizations could results to a 400% relative error.
+**But that might favor models with chaotic metric because only half of them will be used to compute the error ?
+=> keep median for relative error and mean of absolute error and optimize on MAE instead of MREB ?
+Also, because MAE is a difference it is similar if we compute it on the number of hospitalization or on the variation of the number of hospitalization.
+Also we might discuss to penalize more the error when the number of hospitalizations is high (already done implicitly by AE) and when there is a dynamic shift => ponderate AE by absolute value of second derivative ?**.
 
@@
 
--   Recall hyperparameters
--   Present GA algorithm + explain why it is important here (large
-    number of hp + assumption free compared to other algorithm, cons =
-    expensive)
+RC relies mainly on 4 hyperparameters including the leaking rate (i.e "memory" parameter), spectral radius (i.e "chaoticity" parameter), input scaling (i.e "feature gain" parameter) and ridge (i.e penalization parameter).
+Input scaling can be either common to all features or specific to each feature which increases the number of hyperparameter by the number of features.
+As the relevancy of feature for forecast evolves over time, we explored (i) filter based feature selection based on elastic-net before RC which has a penalization hyperparameter denoted as $enet$ and (ii) wrapper feature selection for which each feature has an additional binary hyperparameter defining if it should be selected or not.
 
-To tackle both problems, we proposed a genetic algorithm for
-hyperparameter optimisation including feature selection.
+Regarding the number of hyperparameter, an automatic approach is necessary the relevant hyperparameters.
+We chose to use a genetic algorithm (GA) which has the benefit of making few assumptions on the link function between the vector of hyperparameter value and the objective function but might need many trials which are possible due to the fast of RC.
+The objective function of GA was to minimize the $MREB$ metric **(change ?)**.
+The pseudo-code of GA is defined as follow :
 
@@
             -   Integer : add or substract 1 at random
-            -   Numerical : Add $\sigma \times \mathcal{N}(0,1)$
+            -   Numerical : Add $\sigma", "imes \mathcal{N}(0,1)$
 
@@
 
+### Feature selection with RC
+
+We investigated the performance of RC using various feature selection approach and input scaling configuration. The following setting were evaluated :
+
+- Feature selection based on expert inputs with common input scaling
+- Feature selection based on expert inputs with specific input scaling for each feature
+- Genetic algorithm feature selection with common input scaling
+- Genetic algorithm feature selection with specific input scaling for each feature
+- Elastic-net feature selection with common input scaling
+
+Experts inputs were a priori defined features that were judge important for epidemic forecast : hospitalizations, positive RT-PCR, positive RT-PCR among the 60
+
+In addition, we investigated if monthly update of hyperparameter was improving the performance and if optimizing the random seed (which defines the connections inside the reservoir) was improving the results.
+
 ## Results
+
+### Data
+
+```{r casestudyhospitalisations, echo = FALSE, fig.cap = "Hospitalisations at the university hospital of Bordeaux", out.width="75%"}
+knitr::include_graphics(path = here::here("draft_jss/images/monthly update/fig-eval-setting-1.png"))
+```
 
@@
 
--   Present results and detail them
+### Present forecast performance
 
-### Present forecast performance
+```{r echo = FALSE}
+tibble::tribble(
+  ~"Update", ~"Feature", ~"Nb. Features", ~"MAE", ~"MRE", ~"MAEB", ~"MREB ",
+  "No", "Epi - Single IS", 14.0, 13.79, 0.34, -2.92, 0.77 ,
+  "Yes", "Epi - Single IS", 14.0, 10.16, 0.32, -5.09, 0.67 ,
+  "No", "Epi - Multiple IS", 14.0, 9.35, 0.28, -5.31, 0.66 ,
+  "Yes", "Epi - Multiple IS", 14.0, 11.83, 0.31, -2.03, 0.80 ,
+  "No", "GA - Single IS", 231.0, 10.62, 0.28, -5.05, 0.72 ,
+  "Yes", "GA - Single IS", 225.0, 9.33, 0.28, -4.76, 0.72 ,
+  "No", "GA - Multiple IS", 227.0, 9.67, 0.33, -3.64, 0.71 ,
+  "Yes", "GA - Multiple IS", 227.5, 9.45, 0.29, -3.90, 0.69 ,
+  "No", "GA - Multiple IS (seed)", 225.0, 10.68, 0.27, -4.37, 0.72 ,
+  "Yes", "GA - Multiple IS (seed)", 227.5, 10.20, 0.28, -5.33, 0.68 ,
+  "No", "Enet - Single IS", 172.0, 11.72, 0.31, -3.29, 0.79 ,
+  "Yes", "Enet - Single IS", 208.0, 13.68, 0.35, -2.58, 0.81) %>%
+  knitr::kable(caption = "Model performance with or without monthly update (Update), feature selection (Feature).
+For each setting, 40 reservoir are computed and the forecast is the median of the 40 forecast.
+Results show the mean number of features and the performance metrics.
+MAE = Median Absolute Error, MRE = Median Relative Error, MAEB = Median Absolute Error to Baseline, MREB = Median Relative Error to Baseline.",
+               booktabs = TRUE)
+```
+
+```{r echo = FALSE, fig.cap = "Reservoir computing forecast depending on the setting without monthly update.
+Red line is the median forecast of 40 reservoirs.
+Grey lines are individual forecast of each of the 40 reservoir.", out.width="75%"}
+knitr::include_graphics(path = here::here("draft_jss/images/monthly update/fig-eval-prediction-noupdate-1.png"))
+```
+
+```{r echo = FALSE, fig.cap = "Reservoir computing forecast depending on the setting with monthly update.
+Red line is the median forecast of 40 reservoirs.
+Grey lines are individual forecast of each of the 40 reservoir.", out.width="75%"}
+knitr::include_graphics(path = here::here("draft_jss/images/monthly update/fig-eval-prediction-update-1.png"))
+```
 

### 2023-10-30 draft_jss/draft_jss.Rmd
update v0.0.1

@@
 
+**TODO : cite : Jaeger, Herbert.
+“The “echo state” approach to analysing and training recurrent neural networks-with an erratum note.” Bonn, Germany: German National Research Center for Information Technology GMD Technical Report 148.34 (2001): 13.**
+**Maass W, Natschläger T, Markram H. Real-time computing without stable states: a new framework for neural computation based on perturbations. Neural Comput. 2002;14(11):2531‐2560.**
+
 ## Why reservoir are interesting compared to standard approaches (penalised regression) and deep learning approaches?
@@
 forecasting and speech recognition, thanks to its inherent memory
-capabilities. RC offers computational efficiency, reduced susceptibility
-to overfitting, and a higher degree of interpretability compared to deep
-learning models. Moreover, RC is able to capture intricate temporal
+capabilities. RC offers computational efficiency and reduced susceptibility
+to overfitting compared to deep learning. Moreover, RC is able to capture intricate temporal
 patterns compared to statistical methods such as penalized regression.
 
+**TODO : find ref to support it.**
+
 ## examples of previous use
 
-TODO: - Ghosh S, Senapati A, Mishra A, Chattopadhyay J, Dana SK, Hens C,
-et al. Reservoir computing on epidemic spreading: A case study on
-COVID-19 cases. Phys Rev E. 16 juill 2021;104(1):014308. - Existing
-review?
+**TODO: add ref from use case intro on use in epidemiology + find review**
+**Table 1 contains a list of use case of RC : https://www.sciencedirect.com/science/article/pii/S0893608019300784**
 
@@
 
-RC use is stille challenging for several reasons. First, its stochastic
+RC use is still challenging for several reasons. First, its stochastic
 nature due to the reservoir random part makes the results of a given RC
 unpredictable even if the hyperparameters are well chosen. Second,
-eventhough, RC relies on several hyperparameter to determine the
+even though, RC relies on several hyperparameter to determine the
 connections between neurons. There exist few guidance about the way to
@@
 approach were combinations of hyperparameters performance are evaluated
-on a train set before chosing the ideal combination one. Third, most
+on a train set before choosing the ideal combination one. Third, most
 published paper are based on a low dimensional setting and few guidance
@@
 the choice of hyperparameters. Fourth, the use of reservoir computing in
-an epidemic setting with highly non-stationary time serie is scarce
+an epidemic setting with highly non-stationary time series is scarce
 whereas it is of tremendous interest for epidemiologists. Fifth, there
@@
 
-## what we do
-
-In this paper we propose to explore those challenges with extensive
-guidance in order to help new user to fully benefit from RC. First a
-general introduction to reservoir computing is provided with a tutorial
-for its use using reservoirnet, a R package based on reservoirPy Python
-module. Second, we explore the different challenges of the use of RC in
-a non-stationary high dimensional setting based on the use case of
-covid-19 hospitalisations forecast with extensive guidance on the
-modelling strategy, the choice of hyperparameters using a genetic
-algorithm and the implementation.
-
-# RC presentation
-
-RC is a machine learning algorithm composed of an input layer denoted
-$u(t)$ which is randomy transformed by the reservoir into an activation
-state denoted $x(t)$ which is fed to a ridge penalized linear regression
-trained to foreast the outcome denoted as $y(t)$ as depicted at figure
-\@ref(fig:rcpresentation).
-
-```{r rcpresentation, echo=FALSE, fig.cap="my caption", out.width = '90%', fig.cap="Reservoir computing is composed of an input layer, a reservoir and an output layer.
-Connection between input layer and reservoir and inside reservoir are random.
-Only the output layer is optimized based on a ridge penalized linear regression."}
-knitr::include_graphics("images/schema.png/image1.png")
-```
-
-The input layer $u(t)$ is an $M$-dimension vector which corresponds to
-the values of the input time series at time $t$ where $t = 1, …, T$. The
-reservoir layer $x(t)$ is an $N_{res}$-dimensional vector where
-$N_{res}$ is the number of nodes in the reservoir. The value $x(t)$ is
-defined as follow:
-
-$$x( t+1 ) = ( 1 - \alpha )  x ( t) + \alpha \: tanh( W x(t) + W_{in} u(t+1) )", "ext{ where } \alpha \in [0, 1 ]$$
-
-The leaking rate alpha define the rate of update of the nodes. The
-closer $\alpha$ is to $1$, the higher the reservoir is sensitive to new
-inputs (i.e $u(t)$). Therefore, the reservoir state at time $t+1$
-denoted $x(t+1)$ depends on the reservoir state at the previous time
-(i.e $x(t)$) and the new inputs (i.e $u(t+1)$). Both $W_{in}$ and $W$
-are random matrices of size $Nres", "imes M$ and $Nres", "imes Nres$
-respectively.
-
-$W_{in}$ is a dense matrix generated using a bernouilli distribution
-where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an
-equal probability where $m = 1, …, M$ corresponds to a given feature in
-the input layer. The input scaling, denoted $I_{scale}$, is an
-hyperparameter coefficient which can be common to all features from the
-input layer or specific to each feature $m$. In that case, the more
-important the feature is, the greater should be its input scaling. $W$
-is a sparse matrix where values are generated base on a gaussian
-distribution $\mathcal{N}(0,1)$. Then the matrix $W$ is rescaled
-according to the spectral radius, an hyperparameter defining the highest
-eigen value of $W$.
-
-The final layer is a linear regression with ridge penalization where the
-explanatory features are the reservoir state and the variable to be
-explain is the outcome to predict such that:
-
-$$W_{out} = YX^T ( XX^T + \lambda  I)^{ -1 }$$
-
-Where x(t) and y(t) are accumulated in X and Y respectively such that:
-
-$$X = \begin{bmatrix} x(1) \\ x(2) \\ ... \\ x(T) \end{bmatrix}
-\text{ and } Y = \begin{bmatrix} y(1) \\ y(2) \\ ... \\ y(T) \end{bmatrix}$$
-
-The parameter $\lambda$ is the ridge penalisation which aims to prevent
-overfitting.
-
-Overall, there are four main hyperparameters to be chosen by the user:
-the leaking rate wich defines the memory of the RC, the input scaling
-wich define the relative importance of the features, the spectral radius
-wich define the connections of the neurons inside the reservoir and the
-ridge penalization wich defines the degree of overfitting. The choice of
-hyperparameter is difficult and the use of a wrapper approach where the
-performance of the RC with different combinations of hyperparameters is
-evaluated on a train set, the best combination is chosen for the use on
-the test set.
-
+**TODO : extend this part and link to previous work**
+**Incorporate the following :**
 RC in high dimension
@@
 
+## what we do
+
+In this paper we propose to explore those challenges with extensive
+guidance in order to help new user to fully benefit from RC. First a
+general introduction to reservoir computing is provided with a tutorial
+for its use using `reservoirnet`, a R package based on `reservoirPy` Python
+module. Second, we explore the different challenges of the use of RC in
+a non-stationary high dimensional setting based on the use case of
+covid-19 hospitalizations forecast with extensive guidance on the
+modelling strategy, the choice of hyperparameters using a genetic
+algorithm and the implementation.
+
+# RC presentation
+
+RC is a machine learning algorithm composed of an input layer denoted
+$u(t)$ which is randomly transformed by the reservoir into an activation
+state denoted $x(t)$ which is fed to a ridge penalized linear regression
+trained to forecast the outcome denoted as $y(t)$ as depicted at figure
+\@ref(fig:rcpresentation).
+
+```{r rcpresentation, echo=FALSE, fig.cap="my caption", out.width = '90%', fig.cap="Reservoir computing is composed of an input layer, a reservoir and an output layer.
+Connection between input layer and reservoir and inside reservoir are random.
+Only the output layer is optimized based on a ridge penalized linear regression."}
+knitr::include_graphics("images/schema.png/image1.png")
+```
+
+The input layer $u(t)$ is an $M$-dimension vector, where $M$ is the number of input time series, which corresponds to the values of the input time series at time $t$ where $t = 1, …, T$.
+The reservoir layer $x(t)$ is an $N_{res}$-dimensional vector where $N_{res}$ is the number of nodes in the reservoir.
+The value $x(t)$ is defined as follow:
+
+$$x( t+1 ) = ( 1 - \alpha )  x ( t) + \alpha \: tanh( W x(t) + W_{in} u(t+1) ) \text{ , where } \alpha \in [0, 1 ]$$
+
+The leaking rate alpha define the rate of update of the nodes. The
+closer $\alpha$ is to $1$, the higher the reservoir is sensitive to new
+inputs (i.e $u(t)$). Therefore, the reservoir state at time $t+1$
+denoted $x(t+1)$ depends on the reservoir state at the previous time
+(i.e $x(t)$) and the new inputs (i.e $u(t+1)$). Both $W_{in}$ and $W$
+are random matrices of size $Nres \times M$ and $Nres \times Nres$
+respectively.
+
+$W_{in}$ is a dense matrix generated using a Bernoulli distribution
+where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an
+equal probability where $m = 1, …, M$ corresponds to a given feature in
+the input layer. The input scaling, denoted $I_{scale}$, is an
+hyperparameter coefficient which can be common to all features from the
+input layer or specific to each feature $m$. In that case, the more
+important the feature is, the greater should be its input scaling. $W$
+is a sparse matrix where values are generated from a Gaussian
+distribution $\mathcal{N}(0,1)$. Then the matrix $W$ is scaled
+according to the spectral radius, an hyperparameter defining the highest
+eigen value of $W$.
+
+The final layer is a linear regression with ridge penalization where the
+explanatory features are the reservoir state and the variable to be
+explain is the outcome to predict such that:
+
+$$W_{out} = YX^T ( XX^T + \lambda  I)^{ -1 }$$
+
+Where x(t) and y(t) are accumulated in X and Y respectively such that:
+
+$$X = \begin{bmatrix} x(1) \\ x(2) \\ ... \\ x(T) \end{bmatrix}
+\text{ and } Y = \begin{bmatrix} y(1) \\ y(2) \\ ... \\ y(T) \end{bmatrix}$$
+
+The parameter $\lambda$ is the ridge penalization which aims to prevent
+overfitting.
+
+Overall, there are four main hyperparameters to be chosen by the user: the leaking rate which defines the memory of the RC, the input scaling which define the relative importance of the features, the spectral radius which define the connections of the neurons inside the reservoir which in turn define the degree of non-linear combination of features and the ridge penalization which controls the degree of overfitting.
+The choice of hyperparameter is difficult and the use of a wrapper approach where the performance of the RC with different combinations of hyperparameters is evaluated on a train set, the best combination is chosen for the use on the test set.
+
 # Basic package use
@@
 This demonstration will be conducted using the COVID-19 dataset that is included within the package.
-These data encompass hospitalization figures, positive RT-PCR results, and overall RT-PCR data sourced from Santé Publique France, which are publicly available on data.gouv.fr (for further details, refer to the `help(dfCovid)` function).
+These data encompass hospitalization, positive RT-PCR results, and overall RT-PCR data sourced from Santé Publique France, which are publicly available on data.gouv.fr (for further details, refer to `help(dfCovid)`).
 Our primary objective is to predict the number of hospitalized patients 14 days into the future.
@@
 
-We can proceed by loading useful packages (i.e ggplot2 [@wickham_ggplot2_2016] and dplyr [@wickham_dplyr_2023]), data and define the task:
+We can proceed by loading useful packages (i.e ggplot2 [@wickham_ggplot2_2016] and dplyr [@wickham_dplyr_2023], **TODO : add cowplot**), data and define the task:
 
@@
 library(reservoirnet)
+library(cowplot)
 
@@
 Moreover, we calculate the `outcome_deriv` as the difference between the outcome and the number of hospitalized patients (`hosp`), representing the variation in hospitalization in relation to the current count of hospitalized individuals.
-The resulting smoothed data is visualized in Figure @ref(fig:covidintro).
+The resulting smoothed data is visualized in Figure \@ref(fig:covidintro).
 
@@
 
-We can now plot the data :
-
-```{r covidintro, fig.cap="Hospitalisations, IPTCC and positive PCR of Bordeaux University Hospital.", echo = FALSE}
+```{r covidintro, fig.cap="Hospitalizations, IPTCC and positive PCR of Bordeaux University Hospital.", echo = FALSE}
 dfOutcome %>%
@@
 
--   Number of nodes (`units`) : it corresponds to the number of nodes inside the reservoir. Usually, the more the better but more nodes increase the computation time.
--   Leaking rate (`lr`) : the leaking rate corresponds to the balance between the new inputs and the previous state. A leaking rate of 1 only consider informations from the new inputs.
+-   Number of nodes (`units`) : it corresponds to the number of nodes inside the reservoir. Usually, the more the better but more nodes increases the computation time.
+-   Leaking rate (`lr`) : the leaking rate corresponds to the balance between the new inputs and the previous state. A leaking rate of 1 only consider information from new inputs.
 -   Spectral radius (`sr`): the spectral radius is the maximum absolute eigenvalue of the reservoir connectivity matrix.
@@
 
-For this part of the tutorial, we will set the hyperparameter at a given value. Hyperparameter opitmisation will be detailed at section [Study case](#studycase).
+For this part of the tutorial, we will set the hyperparameter at a given value. Hyperparameter optimization will be detailed at section [Study case](#studycase).
 
@@
 
-Now we can visualise node activation using the `plot` function presented at figure \@ref(fig:nodeactivationbad).
-
-```{r nodeactivationbad, fig.cap="Node acivation over time."}
+Now we can visualize node activation using the `plot` function presented at figure \@ref(fig:nodeactivationbad).
+
+```{r nodeactivationbad, fig.cap="Node activation over time."}
 plot(reservoir_state)
@@
 This issue can be attributed to the disparate scales of the features.
-To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby rescaling them within the range of `-1` to `1`.
+To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby scaling them within the range of `-1` to `1`.
 
@@
 It's important to note that this hyperparameter can be subject to optimization, a topic that will be explored in the forthcoming [Study Case](#studycase) section.
-This parameter plays a pivotal role in fine-tuning the model's conformity to the dataset.
+This parameter plays a pivotal role in fine-tuning the model's conformity to the data.
 When set excessively high, the risk of underfitting arises, whereas setting it too low can lead to overfitting.
-Then we connect the output layer to the reservoir making the model ready to be trained.
+We connect the output layer to the reservoir making the model ready to be trained.
 
@@
 
-We now standardise with the same formula as seen before. We learn the standardisation on the training set and apply it on the test set. Then we convert the dataframe to matrix.
+We standardize with the same formula as seen before. We learn the standardization on the training set and apply it on the test set. Then we convert the dataframe to matrix.
 
@@
   theme_bw() +
-  labs(color = "", x = "Date", y = "Hospitalisations")
+  labs(color = "", x = "Date", y = "Hospitalizations")
 ```
@@
 We observe that the model forecast is not fully accurate, both on the test set and the train set.
-In that case, one option could be to reduce ridge penalisation to fit more closely the data, the optimisation of ridge hyperparameter will be discussed at section [Study case](#studycase).
-Another possibility is to ease the learning of the algorithm by forecasting the variation of the hospitalisation instead of the number of hospitalised patients.
+In that case, one option could be to reduce ridge penalization to fit more closely the data, the optimization of ridge hyperparameter will be discussed at section [Study case](#studycase).
+Another possibility is to ease the learning of the algorithm by forecasting the variation of the hospitalization instead of the number of hospitalized patients.
 For that step, we will learn on the `outcome_deriv` contained in `yTrain_variation` data which is defined outcome as `outcome_deriv = outcome - hosp`.
 
-```{r fig.cap="Covid-19 hospitalisations forecast.
-The model is either trained to forecast the number of hospitalisations (denoted Raw) or the variation of the hospitalisations compared to current level of hospitalisation (denoted Variation)"}
+```{r fig.cap="Covid-19 hospitalizations forecast.
+The model is either trained to forecast the number of hospitalizations (denoted Raw) or the variation of the hospitalizations compared to current level of hospitalisation (denoted Variation)"}
 ## Fit reservoir on outcome variation instead of raw outcome
@@
   theme_minimal() +
-  labs(color = "", x = "Date", y = "Hospitalisations")
-```
-
-We observe an improvement of the forecast compared to the forecast of hospitalisation previously discussed.
-From there, many improvement can be implemented including : frequent update of the model, leverage additional features and hyperparameter optimisation.
+  labs(color = "", x = "Date", y = "Hospitalizations")
+```
+
+We observe an improvement of the forecast compared to the forecast of hospitalization previously discussed.
+From there, many improvement can be implemented including : frequent update of the model, leverage additional features and hyperparameter optimization.
 This notions will be discussed in the [Study case section](#studycase).
@@
 To illustrate the classification task, we will use the Japanese vowel dataset (@kudo_multidimensional_1999).
-The data can be loaded from reservoirnet as follow :
+The data can be loaded from `reservoirnet` as follow :
 
@@
 The dataset comprises 640 vocalizations of the Japanese vowel \ae, contributed by nine distinct speakers.
-Each vocalization represents a time series spanning between 7 and 29 timesteps, encoded as a 12-dimensional vector denoting the Linear Prediction Coefficients (LPC).
+Each vocalization represents a time series spanning between 7 and 29 time steps, encoded as a 12-dimensional vector denoting the Linear Prediction Coefficients (LPC).
 A visual representation of six distinct utterances from the test set, originating from three different speakers, is depicted in Figure \@ref(fig:vowelpresentation).
@@
 
-Figure \@ref(fig:seqtovec) shows the prediction for the 6 uterrances depicted at figure \@ref(fig:vowelpresentation) where the model correctly identifies the speaker.
+Figure \@ref(fig:seqtovec) shows the prediction for the 6 utterances depicted at figure \@ref(fig:vowelpresentation) where the model correctly identifies the speaker.
 
 ```{r seqtovec, fig.cap="Prediction in a sequence-to-sequence approach 6 samples with 3 speakers and 2 utterance each.
-The speaker to predict is depicted in red.
-For each of the 6 uterrance, the model correctly identifies the speaker."}
+The speaker to predict is depicted in blue.
+For each of the 6 utterance, the model correctly identifies the speaker."}
 dfplotseqtovec <- lapply(vec_sample,
@@
              ncol = 2) +
-  scale_fill_manual(values = c("darkgrey", "darkred")) +
+  scale_fill_manual(values = c("#BDBDBD", "#A3CEF1")) +
   theme_minimal() +
@@
 
-# Study case: Covid-19 hospitalisations forecast {#studycase}
+# Study case: Covid-19 hospitalizations forecast {#studycase}
 
@@
 This ongoing pandemic has exerted significant strain on healthcare systems, resulting in a surge in hospitalizations.
-This surge, in turn, necessitated modifications to the healthcare infrastructure and gave rise to unprecedented population-wide lockdown measures aimed at preventing the saturation of healthcare facilities [@kim_health_2020, @simoes_organisation_2021, @hubner_surgery_2020].
+This surge, in turn, necessitated modifications to the healthcare infrastructure and gave rise to population-wide lockdown measures aimed at preventing the saturation of healthcare facilities [@kim_health_2020, @simoes_organisation_2021, @hubner_surgery_2020].
 The capacity to predict the trajectory of the epidemic on a regional scale is of paramount importance for effective healthcare system management.
@@
 In France, short-term forecasts  with different methods have been evaluated with similar results [@pottier_forecast_2021, @paireau_ensemble_2022, @carvalho_analysis_2021, @mohimont_convolutional_2021].
-In this context a machine learning algorithm based on linear regression with elastic-net penalisation, leveraging both EHRs and public data, was implemented at Bordeaux University Hospital [@ferte_benefit_2022].
-This model, which aimed at forecasting the number of hospitalised patients at 14 days, showed good performance but struggled to accurately anticipate dynamic shifts of the epidemic.
+In this context a machine learning algorithm based on linear regression with elastic-net penalization, leveraging both Electronic Health Records (EHRs) and public data, was implemented at Bordeaux University Hospital [@ferte_benefit_2022].
+This model, which aimed at forecasting the number of hospitalized patients at 14 days, showed good performance but struggled to accurately anticipate dynamic shifts of the epidemic.
 
 RC has been used in the context of covid-19 epidemic forecast [@ghosh_reservoir_2021, @kmet_bezier_2019, @liu_nanophotonic_2023, @ray_optimized_2021, @zhang_sentiment_2023].
-Among them, @ghosh_reservoir_2021, @liu_nanophotonic_2023 and @ray_optimized_2021 used it to forecast epidemic (@zhang_sentiment_2023 performed sentiment analysis and @kmet_bezier_2019 used it to solve optimal control related to vaccine).
+Among them, @ghosh_reservoir_2021, @liu_nanophotonic_2023 and @ray_optimized_2021 used it to forecast epidemic, @zhang_sentiment_2023 performed sentiment analysis and @kmet_bezier_2019 used it to solve optimal control related to vaccine.
 The evaluation of RC for epidemic forecast showed promising results in all approaches, being competitive with Long-Short Term Memory (LSTM) and Feed-Forward Neural Network (FFNN) in @ray_optimized_2021.
@@
 
+`TODO :` check for new published forecast models
+
 `TODO :` add paragraph about RC in high dimension
 
-`TODO :` add paragraph about hp optimisation for RC (focus on GA)
-
-RC can be approached as a complexification of penalized linear regression where inputs are processed by a reservoir which allows memory and non-linear combinations.
-As penalised linear regression was the best approach for covid-19 forecast in @ferte_benefit_2022 and that RC showed promising results for epidemic forecast in @ghosh_reservoir_2021, @liu_nanophotonic_2023 and @ray_optimized_2021, we proposed to use RC to forecast hospitalisations at 14 days at the University Hospital of Bordeaux.
+`TODO :` add paragraph about hp optimisation for RC (focus on GA). Discussed in https://ieeexplore.ieee.org/abstract/document/8481347
+
+RC can be approached as an extension of penalized linear regression where inputs are processed by a reservoir which allows memory and non-linear combinations.
+As penalized linear regression was the best approach for covid-19 forecast in @ferte_benefit_2022 and that RC showed promising results for epidemic forecast in @ghosh_reservoir_2021, @liu_nanophotonic_2023 and @ray_optimized_2021, we proposed to use RC to forecast hospitalizations at 14 days at the University Hospital of Bordeaux.
 The main objective of this study is to evaluate the performance of RC for this task.
-Secondary objective will be (i) to assess the performance of RC depending on the input dimension, (ii) to assess the relevance of Genetic Algorithm (GA) for hyperparameter optimisation and feature selection in this context.
+Secondary objective will be (i) to assess the performance of RC depending on the input dimension, (ii) to assess the relevance of Genetic Algorithm (GA) for hyperparameter optimization and feature selection in this context.
 
@@
 
-The task was to forecast 14 days ahead the number of hospitalised patients.
-As seen at section [Regression](#basicregression), we will train the model to predict the variation of hospitalisation defined as $outcome_{t+14} = hosp_{t+14} - hosp_{t}$.
-Metrics computation and visualisations will be performed on the predicted number of hospitalizations denoted as $\hat{y_{t+14}} = \widehat{outcome_{t+14}} + hospitalizations_{t}$ and the observed number of hospitalisations at time $t+14$ denoted as $y_{t+14} = hospitalizations_{t+14}$
+The task was to forecast 14 days ahead the number of hospitalized patients.
+As seen at section [Regression](#basicregression), we will train the model to predict the variation of hospitalization defined as $outcome_{t+14} = hosp_{t+14} - hosp_{t}$.
+Metrics computation and visualizations will be performed on the predicted number of hospitalizations denoted as $\hat{y_{t+14}} = \widehat{outcome_{t+14}} + hospitalizations_{t}$ and the observed number of hospitalizations at time $t+14$ denoted as $y_{t+14} = hospitalizations_{t+14}$
 
@@
 Remaining data was used to evaluate the model performance.
-As the time serie is not stationary and most appropriate hyperparameter value might change over time, we also investigated the effect of hyperparameter update every month.
+As the time series is not stationary and most appropriate hyperparameter value might change over time, we also investigated the effect of hyperparameter update every month.
 
@@
 For instance, for relative error, when the outcome is low (e.g equal to 1 hospitalization), an error of 5 hospitalizations could results to a 400% relative error.
-**But that might favor models with chaotic metric because only half of them will be used to compute the error ?
+**But that might favor models with chaotic metric because extreme values won't be used to compute the error ?
 => keep median for relative error and mean of absolute error and optimize on MAE instead of MREB ?
@@
 
-### Hyperparameter optimisation using GA
+### Feature selection with RC
+
+We investigated the performance of RC using various feature selection approach and input scaling configuration. The following setting were evaluated :
+
+- Feature selection based on expert inputs with common input scaling
+- Feature selection based on expert inputs with specific input scaling for each feature
+- Genetic algorithm feature selection with common input scaling
+- Genetic algorithm feature selection with specific input scaling for each feature
+- Elastic-net feature selection with common input scaling
+
+Experts inputs were a priori defined features that were judge important for epidemic forecast : hospitalizations, positive RT-PCR, positive RT-PCR among the 60 years old and more, fraction of positive RT-PCR, fraction of positive RT-PCR among the 60 years old and more, IPTCC, number of patient EHR mentioning "covid-19" in emergency units department, number of people having received at least one dose of vaccine in Gironde.
+Additionally we provided the first derivative at 7 days for each feature.
+
+We used feature selection using elastic-net because it was the best model in previous experiment compared to other machine learning (@ferte_benefit_2022).
+This feature selection impose to select the penalty hyperparameter of elastic-net.
+This hyperparameter was selected using genetic algorithm as depicted [previously](#geneticalgo).
+
+Genetic algorithm feature selection was performed using a binary hyperparameter for each feature determining if it was to be selected or not.
+
+In addition, we investigated if monthly update of hyperparameter was improving the performance and if optimizing the random seed (which defines the connections inside the reservoir) was improving the results.
+
+### Hyperparameter optimisation using GA {#geneticalgo}
 
 RC relies mainly on 4 hyperparameters including the leaking rate (i.e "memory" parameter), spectral radius (i.e "chaoticity" parameter), input scaling (i.e "feature gain" parameter) and ridge (i.e penalization parameter).
-Input scaling can be either common to all features or specific to each feature which increases the number of hyperparameter by the number of features.
+Input scaling can be either, common to all features or specific to each feature which increases the number of hyperparameter by the number of features.
 As the relevancy of feature for forecast evolves over time, we explored (i) filter based feature selection based on elastic-net before RC which has a penalization hyperparameter denoted as $enet$ and (ii) wrapper feature selection for which each feature has an additional binary hyperparameter defining if it should be selected or not.
 
-Regarding the number of hyperparameter, an automatic approach is necessary the relevant hyperparameters.
-We chose to use a genetic algorithm (GA) which has the benefit of making few assumptions on the link function between the vector of hyperparameter value and the objective function but might need many trials which are possible due to the fast of RC.
+Regarding the number of hyperparameter, an automatic approach is necessary.
+We chose to use a genetic algorithm (GA) which has the benefit of making few assumptions on the link function between the vector of hyperparameter value and the objective function but might need many trials which are possible due to the fast of RC (**TODO : add ref**).
 The objective function of GA was to minimize the $MREB$ metric **(change ?)**.
@@
 
-### Feature selection with RC
-
-We investigated the performance of RC using various feature selection approach and input scaling configuration. The following setting were evaluated :
-
-- Feature selection based on expert inputs with common input scaling
-- Feature selection based on expert inputs with specific input scaling for each feature
-- Genetic algorithm feature selection with common input scaling
-- Genetic algorithm feature selection with specific input scaling for each feature
-- Elastic-net feature selection with common input scaling
-
-Experts inputs were a priori defined features that were judge important for epidemic forecast : hospitalizations, positive RT-PCR, positive RT-PCR among the 60
-
-In addition, we investigated if monthly update of hyperparameter was improving the performance and if optimizing the random seed (which defines the connections inside the reservoir) was improving the results.
+For each hyperparameter, we must define the boundaries of the search space.
+We chose a large search space because we had few hypothesis to restrict it.
+Spectral radius and input scaling were defined on $[10^{-5} ; 10^5]$, leaking rate on $[10^{-5} ; 1]$, ridge on $[10^{-10} ; 10^5]$.
+
+The penalization for elastic-net feature selection was initially defined on $[10^{-10} ; 10^5]$.
+However, first experiments showed that, after few generations, GA was stuck in a area were no feature selection was performed and without the ability to sample new individuals with feature selection.
+This phenomena is likely to append for evolutionary algorithm where the sampling variance is small compared to the search space.
+This phenomena is depicted at figure \@ref(fig:enetNbFeatures) were there is no feature selection performed when penalization is below $10^{-4}$.
+To mitigate this problem, we chose to restrict the search space on $[10^{-4} ; 10^5]$.
+
+```{r enetNbFeatures, echo = FALSE, fig.cap = "Number of features depending elastic-net penalization.
+Data from the 100 best enet hp sets.
+Variability represented by boxplot are due to the penalization applied at different dates.", out.width="75%"}
+knitr::include_graphics(path = here::here("draft_jss/images/GA/fig-plot_enet_nb_features-1.png"))
+```
 
@@
 
-### Data
-
-```{r casestudyhospitalisations, echo = FALSE, fig.cap = "Hospitalisations at the university hospital of Bordeaux", out.width="75%"}
+**TODO : change optimisation metric = mean absolute error ?**
+**TODO : optimise the seed ?**
+**TODO : compare with elastic-net alone ?**
+**TODO : compare with LSTM ?**
+**TODO : improve optimisation figures (too small)**
+**TODO : do everything with the R API ?**
+**TODO : provide clean replication code as a supplementary material (too long to be run inside a rmarkdown chunk)**
+
+The goal of this task is to predict 14 days ahead the hospitalization curve depicted at figure \@ref(fig:casestudyhospitalizations).
+It shows both the training set (i.e before 2021-03-01) and the test set encompass several slope shifts which are difficult to forecast.
+We will first introduce the results of the genetic algorithm hyperparameter optimization and then present the forecast performance.
+
+```{r casestudyhospitalizations, echo = FALSE, fig.cap = "Hospitalizations at the university hospital of Bordeaux", out.width="75%"}
 knitr::include_graphics(path = here::here("draft_jss/images/monthly update/fig-eval-setting-1.png"))
@@
 
-### Present results from GA
+### Hyperparameter optimisation
+
+**TODO : add final choice of hyperparameter value**
+
+In this section we will first introduce the results for the features selected based on expert knowledge with multiple input scaling and for the features selected by genetic algorithm with multiple input scaling and seed optimization.
+
+#### Epidemiology subset with multiple input scaling
+
+```{r GAMultipleisEpidemioOptim, echo = FALSE, fig.cap = "Hyperparameter optimisation on epidemiological subet with multiple input scaling.
+2021-03-01 correspond to the result of approximately 3200 trials, subsequent dates corresponds to the result of approximately 1200 trials.
+For each optimisation date, only values of the 200 best individuals are shown.", fig.height=8}
+p1 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleis_epidemio/unnamed-chunk-19-1.png"))
+p2 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleis_epidemio/unnamed-chunk-20-1.png"))
+p3 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleis_epidemio/unnamed-chunk-21-1.png"))
+p4 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleis_epidemio/unnamed-chunk-22-1.png"))
+
+cowplot::plot_grid(p1, p2, p3, p4, nrow = 2, ncol = 2,
+                   labels = c("(a) Convergence",
+                              "(b) Marginal",
+                              "(c) Conditional",
+                              "(d) Feature importance (input scaling)"),
+                   label_size = 8,
+                   label_y = 1.01)
+```
+
+Figures \@ref(fig:GAMultipleisEpidemioOptim) show the marginal value of hyperparameters selected by GA over the different optimization time at panels (a) and (b).
+We note that both leaking rate and ridge values selected on the train set were not well adapted to the test set, they increased rapidly and then stabilized indicating a need leverage more information from new inputs compared to old inputs (leaking rate) and to decrease overfitting (ridge).
+The spectral radius progressively decreased over time which might indicate less importance for non-linear feature projection and a need for less chaotic reservoir.
+
+Panel (c) shows conditional values of hyperparameters.
+We observe that the outlined hyperparameters are independent one from another except for the linear association between leaking rate and ridge but which was observed only for the first updates.
+**I remember Pierrick told us that independance was a good property (probably because if two hp are correlated you can just focus on the optimisation of 1 and deduce the other) but we might need a reference here.**
+
+Panel (d) shows feature importance, depicted by the feature input scaling (i.e the higher the more important), over the different time period.
+It is interesting to note that feature importance vary over time and that important features before 2021-03-01 (IPTCC, derivative of hospitalizations) have an importance decrease of contrary to features related to fraction of positive RT-PCR in the elderly which remained important.
+**Need to change the features label of the figure.**
+
+#### GA feature selection with multiple input scaling and seed control
+
+```{r GAMultipleisGAOptim, echo = FALSE, fig.cap = "Hyperparameter optimisation with GA feature selection with multiple input scaling and seed optimisation.
+2021-03-01 correspond to the result of approximately 3200 trials, subsequent dates corresponds to the result of approximately 1200 trials.
+For each optimisation date, only values of the 200 best individuals are shown.", fig.height=8}
+p1 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleISseed_GA/unnamed-chunk-39-1.png"))
+p2 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleISseed_GA/unnamed-chunk-40-1.png"))
+p3 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleISseed_GA/unnamed-chunk-41-1.png"))
+p4 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleISseed_GA/unnamed-chunk-42-1.png"))
+
+cowplot::plot_grid(p1, p2, p3, p4, nrow = 2, ncol = 2,
+                   labels = c("(a) Convergence",
+                              "(b) Marginal",
+                              "(c) Conditional",
+                              "(d) Feature selection frequency"),
+                   label_size = 8,
+                   label_y = 1.01)
+```
+
+Figure \@ref(fig:GAMultipleisGAOptim) show the marginal value of hyperparameters selected by GA over the different optimization time at panel (a) and (b).
+Leaking rate remained stable over time but again, spectral radius decreased and ridge increased indicating a lesser need for non-linear feature recombination and higher need to control for overfitting.
+Off note we outlined the seed hyperparameter here.
+Seed controls the randomness of RC connections and its value do not have epidemiological interpretation.
+Figure Panel (c) shows hyperparameter independence.
+
+Panel (d) shows feature selection over the different time period.
+It outlines that most important features (i.e the most often selected ones) are the ones related to RT-PCR, specifically for the elderly.
+We also note the importance of features related to emergency and ambulance departments.
+In addition, we observe that feature importance did not remained constant over time justifying the need for an adaptive feature selection algorithm.
+
+### Anteriority
+
+**TODO : if we keep this section, we should update with same hyperparameter (feature selection with only 30 features for GA multiple input scaling) and same aggregation mechanism as in the present forecast performance to have something clean.**
+
+Because epidemic mechanism change over time, it could of interest not to provide all the time series to the model but only its most recent data.
+We investigate whether the data anteriority we used to train and test the model had an impact on its performance.
+Because the hyperparameter optimization process is long, the hyperparameter used were optimized on all the time series (i.e 542 days).
+
+Results of the experiment are outlined at figure \@ref(fig:anteriorityPerformance).
+We show that at least 1 year of data seems necessary to leverage enough information.
+This is what will be used in the remaining sections.
+**Note that epidemio with one input scaling had very good performance on this evaluation.**
+
+```{r anteriorityPerformance, echo = FALSE, fig.cap = "Performance depending on number of days used to train the model.
+10 iterations with different seed for each of 10 best hyperparameters sets.
+Boxplots corresponds to individual models performance (no aggregation) and diamond is the median forecast of the 100 models.", out.width="75%"}
+knitr::include_graphics(path = here::here("draft_jss/images/anteriority/fig-plot_anteriority-test-set-1.png"))
+```
+
+### Number of model to aggregate
+
+**TODO : if we keep this section, we should update with same hyperparameter (feature selection with only 30 features for GA multiple input scaling) and same aggregation mechanism as in the present forecast performance to have something clean.**
+**Note : if we optimize the seed, then the repetition of a given set of hyperparameter makes no sense (i.e will provide same results each time).**
+
+```{r modelBudget, echo = FALSE, fig.cap = "Performance dependending on number of hyperparameter sets (model) and the scenario used.
+Panel (a) explores two scenario : one where the best hyperparameter set is replicated x times with different seed, one where the x best hyperparameter sets are used with one seed each.
+Each scenario is evaluated with the aggregation of 1, 3, 5, 10, 20, 30, 40, 60 or 100 models with 250 bootstrap 95\\%CI Panel (b) explores the optimal balance between both approaches with a budget of 40 models.", fig.height=8}
+p1 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/budget/fig-plot_aggregation_convergence-1.png"), scale = 0.9)
+p2 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/budget/fig-plot_aggregation_convergence_budget-1.png"), scale = 0.9)
+
+cowplot::plot_grid(p1, p2, nrow = 2, ncol = 1,
+                   labels = c("(a) Number of models needed",
+                              "(b) Optimal budget repartition"),
+                   label_size = 8,
+                   label_y = 1.01)
+```
+
+```{r best40epidemio, echo = FALSE, fig.cap = "Performance of each of the best 40 hp sets with 40 replicate each. 250 Bootstrap 95\\%CI. Epidemiology subset with one input scaling.", out.width="75%"}
+knitr::include_graphics(path = here::here("draft_jss/images/budget/fig-plot_aggregation_best40epidemio-1.png"))
+```
+
+Figure \@ref(fig:modelBudget) shows the model performance depending on the model and the aggregation process.
+We observe that for all models, the model performance seems achieved at 30 models, therefore, for remaining results, 40 models will be used.
+All models showed that the best aggregation mechanism was to use different hyperparameter sets instead of using the best hyperparameter set with different seeds.
+The exception was the epidemiology subset with single input scaling whose best performance was achieved using only the first two models.
+This is explained by the low performance of the fourth best hyperparameter set on the test set as depicted at figure \@ref(fig:best40epidemio).
 
@@
 
-```{r echo = FALSE}
+```{r perfMonthlyUpdate, echo = FALSE}
 tibble::tribble(
@@
 MAE = Median Absolute Error, MRE = Median Relative Error, MAEB = Median Absolute Error to Baseline, MREB = Median Relative Error to Baseline.",
-               booktabs = TRUE)
-```
-
-```{r echo = FALSE, fig.cap = "Reservoir computing forecast depending on the setting without monthly update.
+               booktabs = TRUE,
+               linesep = c('', '\\addlinespace'))
+```
+
+Table \@ref(tab:perfMonthlyUpdate) shows the performance on the test set.
+The number of selected features by elastic-net and by genetic algorithm were in the same order of magnitude.
+Updating the hyperparameter monthly did not always improve forecast accuracy (e.g epidemiological subset with multiple input scaling).
+Best model depend on the metric but epidemiological subset with multiple input scaling and no update was the best model or second to best model for all metrics.
+
+```{r forecastwithorwithoutUpdate, echo = FALSE, fig.cap = "Reservoir computing forecast depending on the setting with and without monthly update.
 Red line is the median forecast of 40 reservoirs.
-Grey lines are individual forecast of each of the 40 reservoir.", out.width="75%"}
-knitr::include_graphics(path = here::here("draft_jss/images/monthly update/fig-eval-prediction-noupdate-1.png"))
-```
-
-```{r echo = FALSE, fig.cap = "Reservoir computing forecast depending on the setting with monthly update.
-Red line is the median forecast of 40 reservoirs.
-Grey lines are individual forecast of each of the 40 reservoir.", out.width="75%"}
-knitr::include_graphics(path = here::here("draft_jss/images/monthly update/fig-eval-prediction-update-1.png"))
-```
+Grey lines are individual forecast of each of the 40 reservoir.", fig.height=8}
+p1 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/monthly update/fig-eval-prediction-noupdate-1.png"))
+p2 <- cowplot::ggdraw() +
+  cowplot::draw_image(here::here("draft_jss/images/monthly update/fig-eval-prediction-update-1.png"))
+
+cowplot::plot_grid(p1, p2, nrow = 2, ncol = 1,
+                   labels = c("(a) No update",
+                              "(b) Monthly update"),
+                   label_size = 8,
+                   label_y = 1.01)
+```
+
+Figure \@ref(fig:forecastwithorwithoutUpdate) shows the forecast with and without monthly hyperparameter update.
+Graphically we observe that the forecast of individual models (in grey) are chaotic and that forecast aggregation (in red) is necessary.
+This is less concerning for epidemiological subset with multiple input scaling or when the seed is controlled.
+In terms of performance, we observe similar conclusion as the one depicted for table \@ref(tab:perfMonthlyUpdate).
+However, the model errors did not occur for the same period for each model.
+For instance, the epidemiological with multiple scaling and monthly update error is located mainly during the winter 2022 increase.
+Model based on all features tend to overestimate the length of the summer 2021 wave.
+
+```{r perfPreviousExpNoFeatureSelection, echo = FALSE}
+tibble::tribble(
+  ~"Update", ~"Feature", ~"Nb. Features", ~"MAE", ~"MRE", ~"MAEB", ~"MREB ",
+  "No", "Epi - Single IS", 14.0, 12.08, 0.26, -3.04, 0.76,
+  "No", "Epi - Multiple IS", 14.0, 11.83, 0.28, -4.96, 0.67,
+  "No", "Epi - Multiple IS*", 14.0, 10.73, 0.28, -5.47, 0.65,
+  "No", "No FS - Single IS", 456.0, 9.38, 0.24, -5.23, 0.64,
+  "No", "No FS - Single IS*", 456.0, 12.16, 0.28, -4.38, 0.71,
+  "No", "Enet - Single IS", 153.5, 12.51, 0.30, -4.76, 0.76,
+  "No", "GA - Multiple IS limited to 30 features", 20.5, 10.47, 0.27, -3.03, 0.75,
+  "No", "CMAES - Multiple IS", 374.0, 13.91, 0.26, -2.62, 0.83
+  ) %>%
+  knitr::kable(caption = "Model performance depending on feature selection (Feature).
+For each setting, several reservoir are computed and the forecast is the median of the 40 forecast.
+Reservoir aggregation could be either to use the 10 best hyperparameter set once resulting in 10 models or to replicate each of the 10 best hyperparameter set 10 times (those models are denoted by a *) resulting in 100 models.
+Feature selection based on GA and CMAES use top k features with the highest input scaling instead of an indicator for each feature as previously depicted and GA was limited to 30 features.
+Models depicted as 'No FS' are models without feature selection.
+Results show the mean number of features and the performance metrics.
+MAE = Median Absolute Error, MRE = Median Relative Error, MAEB = Median Absolute Error to Baseline, MREB = Median Relative Error to Baseline.",
+               booktabs = TRUE,
+               linesep = c('\\addlinespace'))
+```
+
+Table \@ref(tab:perfPreviousExpNoFeatureSelection) shows additional results from previous experiences without monthly update of hyperparameter.
+The performance without feature selection are good compared to other results.
+Maybe we should give it a try and add it to table \@ref(tab:perfMonthlyUpdate) ?
 
@@
 
-Those work were done previously, the setting of train/test and
-hyperparameter optimisation were different and probably they should be
-done again in a similar setting for relevant comparison.
+Those work were done previously, the setting of train/test and hyperparameter optimisation were different and probably they should be done again in a similar setting for relevant comparison.
 
@@
 
-See
 <https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/explore_esn.md>
@@
 
-See
 <https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/explore_esn_r_pipeline.md>
@@
 
-See
 <https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/tune_esn2.md>
@@
 
-See
 <https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/meeting_1801/master_report.md>

### 2023-11-16 script/find_hyperparameter/esn_find_hp.R
update hp finder

@@ -1,3 +1,5 @@
+timestart = Sys.time()
+print(timestart)
 ##### LOAD PACKAGES
 library(dplyr)
 library(reservoirnet)
@@ -11,7 +13,6 @@ invisible(lapply(list.files(here::here("functions/"), full.names = TRUE), source
 ##### LOAD DATA (only data before 2021-03-01 for learning hp)
 lsdataDates <- fct_load_data_find_hp()
 ##### META PARAMETERS
-strsplit(slar_job_name, split = c("iter", "link"))
 nb_iter <- regexpr("\\d+", slar_job_name) %>%
   regmatches(slar_job_name, .) %>%
   as.numeric()
@@ -53,6 +54,7 @@ dfHyperparam <- cbind(df_fct_hp, df_fct_is)
 ##### COMPUTE MEAN ABSOLUTE ERROR
 dfres <- lapply(seq_len(n_samples),
                 function(id_sample){
+                  print(paste0("---------------- id_sample : ", id_sample))
                   time_start <- Sys.time()
                   mean_absolute_error <- fct_objective(lsdataDates$data_covid,
                                                        lsdataDates$vecDates,
@@ -94,6 +96,7 @@ dfHyperparam <- reservoirnet::random_search_hyperparam(n = n_samples,
 ##### COMPUTE MEAN ABSOLUTE ERROR
 dfres <- lapply(seq_len(n_samples),
                 function(id_sample){
+                  print(paste0("----------------- id_sample : ", id_sample))
                   time_start <- Sys.time()
                   mean_absolute_error <- fct_objective(data_covid = lsdataDates$data_covid,
                                                        vecDates = lsdataDates$vecDates,
@@ -120,3 +123,8 @@ dfres <- lapply(seq_len(n_samples),
 fct_save_results(subDir = paste0("data/common_input_scaling_", slar_jobid),
                  slar_taskid = slar_taskid,
                  object = dfres)
+
+timeend = Sys.time()
+print(timeend)
+print(timestart, timeend, units = "hour")
+

### 2023-11-16 script/forecast/forecast_test.R
update test script

@@ -14,10 +14,10 @@ warmup = 30
 units = 500
 nb_hp_set = 40
 ##### LOAD BEST HP SETS
-hp_sets = list(common_input_scaling = "data/common_input_scaling_11374318/",
-                   common_input_scaling_linked_source = "data/common_input_scaling_linked_source11374319/",
-                   multiple_input_scaling = "data/multiple_input_scaling_11374318/",
-                   multiple_input_scaling_linked_source = "data/multiple_input_scaling_linked_source11374319/") %>%
+hp_sets = list(common_input_scaling = "data/common_input_scaling_11423760/",
+               common_input_scaling_linked_source = "data/common_input_scaling_11426140/",
+               multiple_input_scaling = "data/multiple_input_scaling_11423760/",
+               multiple_input_scaling_linked_source = "data/multiple_input_scaling_11426140/") %>%
   lapply(function(path_i){
     list.files(path_i, full.names = TRUE) %>%
       lapply(readRDS) %>%
@@ -26,7 +26,7 @@ hp_sets = list(common_input_scaling = "data/common_input_scaling_11374318/",
   })
 
 best_hp_set = lapply(hp_sets,
-       function(x) x %>% slice_min(mean_absolute_error, n = nb_hp_set))
+                     function(x) x %>% slice_min(mean_absolute_error, n = nb_hp_set))
 ##### DEFINE EVALUATION SET
 vecDates <- data_covid %>%
   filter(START_DATE >= as.Date("2021-03-01") + forecast_days) %>%
@@ -48,7 +48,10 @@ forecast_common_is <- lapply(X = 1:nrow(best_hp_set$common_input_scaling),
                                                               sr = as.numeric(row[["spectral_radius"]]),
                                                               ridge = as.numeric(row[["ridge"]]),
                                                               input_scaling = as.numeric(row[["input_scaling"]]),
-                                                              seed = as.integer(row[["seed"]])) %>%
+                                                              seed = as.integer(row[["seed"]]),
+                                                              link_source = as.logical(row[["link_source"]]),
+                                                              model = "esn",
+                                                              nb_iter = 1) %>%
                                  mutate(hp_set = row[["hp_set"]])
                              }) %>%
   bind_rows() %>%
@@ -67,7 +70,10 @@ forecast_common_is_linked_source <- lapply(X = 1:nrow(best_hp_set$common_input_s
                                                                             sr = as.numeric(row[["spectral_radius"]]),
                                                                             ridge = as.numeric(row[["ridge"]]),
[…]

### 2023-11-16 script/forecast/forecast_test.R
update take only best hp set

@@ -12,7 +12,8 @@ data_covid <- readRDS(file = "data/df_obfuscated_epidemio.rds")
 forecast_days = 14
 warmup = 30
 units = 500
-nb_hp_set = 40
+nb_hp_set = 1
+nb_iter = 40
 ##### LOAD BEST HP SETS
 hp_sets = list(common_input_scaling = "data/common_input_scaling_11423760/",
                common_input_scaling_linked_source = "data/common_input_scaling_11426140/",
@@ -51,7 +52,7 @@ forecast_common_is <- lapply(X = 1:nrow(best_hp_set$common_input_scaling),
                                                               seed = as.integer(row[["seed"]]),
                                                               link_source = as.logical(row[["link_source"]]),
                                                               model = "esn",
-                                                              nb_iter = 1) %>%
+                                                              nb_iter = nb_iter) %>%
                                  mutate(hp_set = row[["hp_set"]])
                              }) %>%
   bind_rows() %>%
@@ -73,7 +74,7 @@ forecast_common_is_linked_source <- lapply(X = 1:nrow(best_hp_set$common_input_s
                                                                             seed = as.integer(row[["seed"]]),
                                                                             link_source = as.logical(row[["link_source"]]),
                                                                             model = "esn",
-                                                                            nb_iter = 1) %>%
+                                                                            nb_iter = nb_iter) %>%
                                                mutate(hp_set = row[["hp_set"]])
                                            }) %>%
   bind_rows() %>%
@@ -104,7 +105,7 @@ forecast_multiple_is <- lapply(X = 1:nrow(best_hp_set$multiple_input_scaling),
                                                                 seed = as.integer(row[["seed"]]),
                                                                 link_source = as.logical(row[["link_source"]]),
                                                                 model = "esn",
-                                                                nb_iter = 1) %>%
+                                                                nb_iter = nb_iter) %>%
                                    mutate(hp_set = row[["hp_set"]])
                                }) %>%
   bind_rows() %>%
@@ -135,7 +136,7 @@ forecast_multiple_is_linked_source <- lapply(X = 1:nrow(best_hp_set$multiple_inp
                                                                               seed = as.integer(row[["seed"]]),
                                                                               link_source = as.logical(row[["link_source"]]),
                                                                               model = "esn",
-                                                                              nb_iter = 1) %>%
+                                                                              nb_iter = nb_iter) %>%
                                                  mutate(hp_set = row[["hp_set"]])
                                              }) %>%
   bind_rows() %>%

### 2023-11-16 script/forecast/forecast_test_eval.R
add forecast evaluation

@@ -1,0 +1,177 @@
+library(dplyr)
+library(ggplot2)
+
+ls_files_results <- list.files("data/results_forecast_testset_11463695//", full.names = TRUE)
+ls_files_results_res <- grep(ls_files_results, pattern = "hp_sets", value = TRUE, invert = TRUE)
+hp_file <- grep(ls_files_results, pattern = "hp_sets", value = TRUE)
+
+##### hyperparameters plots
+hp_features_reservoir <- c("input_scaling",
+                           "spectral_radius",
+                           "leaking_rate",
+                           "ridge")
+hp_features <- c("hosp",
+                 "P_TOUS_AGES",
+                 "P_60_90_PLUS_ANS",
+                 "FRACP_TOUS_AGES",
+                 "FRACP_60_90_PLUS_ANS",
+                 "URG_covid_19_COUNT",
+                 "IPTCC.mean",
+                 "Vaccin_1dose",
+                 "hosp_rolDeriv7",
+                 "P_TOUS_AGES_rolDeriv7",
+                 "P_60_90_PLUS_ANS_rolDeriv7",
+                 "FRACP_TOUS_AGES_rolDeriv7",
+                 "FRACP_60_90_PLUS_ANS_rolDeriv7",
+                 "URG_covid_19_COUNT_rolDeriv7",
+                 "IPTCC.mean_rolDeriv7",
+                 "Vaccin_1dose_rolDeriv7")
+
+feature_labels <- conservation_status <- c(
+  "input_scaling" = "Input scaling",
+  "leaking_rate" = "Leaking rate",
+  "ridge" = "Ridge",
+  "spectral_radius" = "Spectral radius",
+  "hosp" = "Hospitalizations",
+  "P_TOUS_AGES" = "Positive RT-PCR",
+  "P_60_90_PLUS_ANS" = "Positive RT-PCR 60 yo+",
+  "FRACP_TOUS_AGES" = "% Positive RT-PCR",
+  "FRACP_60_90_PLUS_ANS" = "% Positive RT-PCR 60 yo+",
+  "URG_covid_19_COUNT" = "Emergency sojourn with covid-19",
+  "IPTCC.mean" = "IPTCC",
+  "Vaccin_1dose" = "Vaccine",
+  "hosp_rolDeriv7" = "Hospitalizations (1st deriv)",
[…]

### 2023-11-17 functions/fct_fit_enet.R
add enet hp finder

@@ -5,13 +5,24 @@
 #' @param X The training features
 #' @param Xtest The test features
 #' @param Y The training outcome
+#' @param lambda The lambda elastic-net (use cv.glmnet if not supplied)
+#' @param alpha The alpha parameter of elastic-net
 #'
 #' @return A vector with elastic-net predictions
-fct_fit_enet <- function(X, Xtest, Y){
-  enet <- glmnet::cv.glmnet(x = X,
-                            y = Y,
-                            alpha = 0.5,
-                            family = "gaussian")
-  res <- predict(enet, newx = Xtest, s = "lambda.1se")
+fct_fit_enet <- function(X, Xtest, Y, lambda = NULL, alpha = 0.5){
+  if(is.null(lambda)){
+    enet <- glmnet::cv.glmnet(x = X,
+                              y = Y,
+                              alpha = alpha,
+                              family = "gaussian")
+    res <- predict(enet, newx = Xtest, s = "lambda.1se")
+  } else {
+    enet <- glmnet::glmnet(x = X,
+                           y = Y,
+                           lambda = lambda,
+                           alpha = alpha,
+                           family = "gaussian")
+    res <- predict(enet, newx = Xtest)
+  }
   return(as.numeric(res))
 }

### 2023-11-17 functions/fct_iterative_forecast_from_hp.R

@@ -15,6 +15,8 @@
 #' @param link_source Should the input be linked to the target
 #' @param model esn or enet
 #' @param nb_iter Number of replication of the model
+#' @param lambda The lambda elastic-net (use cv.glmnet if not supplied)
+#' @param alpha The alpha parameter of elastic-net
 #'
 #' @return A dataframe with date, outcome date, forecast, outcome and current hospitalisations
 fct_iterative_forecast_from_hp <- function(data_covid,
@@ -29,7 +31,9 @@ fct_iterative_forecast_from_hp <- function(data_covid,
                                            ridge = 1e2,
                                            input_scaling = 1,
                                            seed = 1,
-                                           nb_iter = 1){
+                                           nb_iter = 1,
+                                           lambda = NULL,
+                                           alpha = 0.5){
   # iterate over each date in vecDates to train and forecast with a reservoir
   dfforecast <- lapply(vecDates,
                        function(date_i){
@@ -49,6 +53,7 @@ fct_iterative_forecast_from_hp <- function(data_covid,
                                                                                    min = 1,
                                                                                    max = 1e5)),
                                                                  no = seed)
+                                                  print(seed)
                                                   # train esn on train set and predict on test set
                                                   forecast_deriv <- fct_fit_esn(
                                                     link_source = link_source,
@@ -66,7 +71,9 @@ fct_iterative_forecast_from_hp <- function(data_covid,
                                                 } else if(model == "enet"){
                                                   forecast_deriv <- fct_fit_enet(X = ls_train_test$X,
                                                                                  Xtest = ls_train_test$Xtest,
-                                                                                 Y = ls_train_test$Y)
+                                                                                 Y = ls_train_test$Y,
+                                                                                 lambda = lambda,
+                                                                                 alpha = alpha)
                                                 } else {
                                                   stop("unknown model argument")
                                                 }

### 2023-11-17 functions/fct_objective.R

@@ -15,6 +15,8 @@
 #' @param link_source Should the input be linked to the target
 #' @param model esn or enet
 #' @param nb_iter Number of replication of the model
+#' @param lambda The lambda elastic-net (use cv.glmnet if not supplied)
+#' @param alpha The alpha parameter of elastic-net
 #'
 #' @return The mean absolute error
 fct_objective <- function(data_covid,
@@ -29,7 +31,9 @@ fct_objective <- function(data_covid,
                           ridge = 1e2,
                           input_scaling = 1,
                           seed = 1,
-                          nb_iter = 1){
+                          nb_iter = 1,
+                          lambda = NULL,
+                          alpha = 0.5){
   # iterate over each date in vecDates to train and forecast with a reservoir
   fct_iterative_forecast_from_hp(data_covid = data_covid,
                                  vecDates = vecDates,
@@ -43,7 +47,9 @@ fct_objective <- function(data_covid,
                                  ridge = ridge,
                                  input_scaling = input_scaling,
                                  seed = seed,
-                                 nb_iter = nb_iter) %>%
+                                 nb_iter = nb_iter,
+                                 lambda = lambda,
+                                 alpha = alpha) %>%
     # take the median forecast if several forecast (nb_iter > 1)
     group_by(START_DATE, outcomeDate, outcome, hosp) %>%
     summarise(forecast = median(forecast), .groups = "drop") %>%

### 2023-11-17 script/find_hyperparameter/enet_find_hp.R

@@ -1,0 +1,48 @@
+timestart = Sys.time()
+print(timestart)
+##### LOAD PACKAGES
+library(dplyr)
+library(reservoirnet)
+##### LOAD FUNCTIONS
+invisible(lapply(list.files(here::here("functions/"), full.names = TRUE), source))
+##### LOAD DATA (only data before 2021-03-01 for learning hp)
+lsdataDates <- fct_load_data_find_hp()
+##### META PARAMETERS
+n_samples = 2000
+########################## COMMON INPUT SCALING ##########################
+##### SAMPLE HYPERPARAMETERS
+dfHyperparam <- reservoirnet::random_search_hyperparam(n = n_samples,
+                                                       ls_fct = list(lambda = function(n) reservoirnet::rloguniform(n = n, min = 1e-10, max = 1e5),
+                                                                     alpha = function(n) runif(n = n, min = 0, max = 1))) %>%
+  dplyr::bind_rows() %>%
+  dplyr::select(-search_id)
+##### COMPUTE MEAN ABSOLUTE ERROR
+dfres <- parallel::mclapply(seq_len(n_samples),
+                            mc.cores = parallel::detectCores()-2,
+                            function(id_sample){
+                              print(paste0("----------------- id_sample : ", id_sample))
+                              time_start <- Sys.time()
+                              mean_absolute_error <- fct_objective(data_covid = lsdataDates$data_covid,
+                                                                   model = "enet",
+                                                                   vecDates = lsdataDates$vecDates,
+                                                                   forecast_days = 14,
+                                                                   lambda = dfHyperparam$lambda[id_sample],
+                                                                   alpha = dfHyperparam$alpha[id_sample])
+                              time_end <- Sys.time()
+                              res <- dfHyperparam[id_sample,] %>%
+                                dplyr::mutate(mean_absolute_error = mean_absolute_error,
+                                              time = difftime(time_end, time_start, units = "secs") %>% as.numeric(),
+                                              link_source = link_source,
+                                              nb_iter = 1)
+                              return(res)
+                            }) %>%
+  bind_rows()
+
+##### SAVE RESULTS
+fct_save_results(subDir = paste0("data/enet_", 1),
+                 slar_taskid = 1,
+                 object = dfres)
+
+timeend = Sys.time()
+print(timeend)
+

### 2023-11-17 script/forecast/forecast_test.R

@@ -18,7 +18,8 @@ nb_iter = 40
 hp_sets = list(common_input_scaling = "data/common_input_scaling_11423760/",
                common_input_scaling_linked_source = "data/common_input_scaling_11426140/",
                multiple_input_scaling = "data/multiple_input_scaling_11423760/",
-               multiple_input_scaling_linked_source = "data/multiple_input_scaling_11426140/") %>%
+               multiple_input_scaling_linked_source = "data/multiple_input_scaling_11426140/",
+               enet = "data/enet_1/") %>%
   lapply(function(path_i){
     list.files(path_i, full.names = TRUE) %>%
       lapply(readRDS) %>%
@@ -143,11 +144,21 @@ forecast_multiple_is_linked_source <- lapply(X = 1:nrow(best_hp_set$multiple_inp
   mutate(model = "esn_multiple_is_linked_source")
 ##### ELASTIC-NET
 message("--------- ELASTIC-NET ----------")
-forecast_enet <- fct_iterative_forecast_from_hp(data_covid = data_covid,
+forecast_enet_once <- fct_iterative_forecast_from_hp(data_covid = data_covid,
+                                                vecDates = date_i,
+                                                forecast_days = forecast_days,
+                                                lambda = best_hp_set$enet$lambda,
+                                                alpha = best_hp_set$enet$alpha,
+                                                model = "enet") %>%
+  mutate(model = "enet_hp_on_train_set",
+         hp_set = 1)
+
+message("--------- ELASTIC-NET (daily hp update) ----------")
+forecast_enet_daily <- fct_iterative_forecast_from_hp(data_covid = data_covid,
                                                 vecDates = date_i,
                                                 forecast_days = forecast_days,
                                                 model = "enet") %>%
-  mutate(model = "enet",
+  mutate(model = "enet_daily_hp_update",
          hp_set = 1)
 ##### SAVE
 message("--------- SAVE ----------")
@@ -155,7 +166,8 @@ dfres = bind_rows(forecast_common_is,
                   forecast_common_is_linked_source,
                   forecast_multiple_is,
                   forecast_multiple_is_linked_source,
-                  forecast_enet)
+                  forecast_enet_once,
+                  forecast_enet_daily)
 
 fct_save_results(subDir = paste0("data/results_forecast_testset_", slar_jobid),
                  slar_taskid = slar_taskid,

### 2023-11-17 script/forecast/forecast_test_eval.R

@@ -1,7 +1,7 @@
 library(dplyr)
 library(ggplot2)
 
-ls_files_results <- list.files("data/results_forecast_testset_11463695//", full.names = TRUE)
+ls_files_results <- list.files("data/results_forecast_testset_11463695/", full.names = TRUE)
 ls_files_results_res <- grep(ls_files_results, pattern = "hp_sets", value = TRUE, invert = TRUE)
 hp_file <- grep(ls_files_results, pattern = "hp_sets", value = TRUE)
 
@@ -133,7 +133,9 @@ model_forecast_labels <- c(
 )
 
 df_forecast <- lapply(ls_files_results_res, readRDS) %>%
-  dplyr::bind_rows() %>%
+  dplyr::bind_rows()
+
+df_forecast_aggregated <- df_forecast %>%
   group_by(START_DATE, outcomeDate, model) %>%
   summarise(outcome = unique(outcome),
             forecast = median(forecast),
@@ -143,7 +145,7 @@ df_forecast <- lapply(ls_files_results_res, readRDS) %>%
                                levels = names(model_forecast_labels),
                                labels = model_forecast_labels))
 
-table_perf_esn <- df_forecast %>%
+table_perf_esn <- df_forecast_aggregated %>%
   dplyr::mutate(absolute_error = abs(forecast - outcome),
                 relative_error = absolute_error/outcome,
                 baseline_absolute_error = abs(hosp - outcome),
@@ -156,7 +158,7 @@ table_perf_esn <- df_forecast %>%
             mre_baseline = median(absolute_error_relative_baseline, na.rm = TRUE),
             .groups = "drop")
 
-plot_forecast <- df_forecast %>%
+plot_forecast <- df_forecast_aggregated %>%
   tidyr::pivot_longer(cols = c(outcome, forecast, hosp)) %>%
   mutate(name = factor(name,
                        levels = c("outcome",
@@ -175,3 +177,66 @@ plot_forecast <- df_forecast %>%
   labs(x = "Date (month-year)",
        y = "Hospitalizations",
        color = "")
+
+##### aggregation forecast plots
+nb_aggregation <- c(1, 5, 10, 20, 30, 40)
+nb_boot <- 250
+
+df_aggregation_forecast <- parallel::mclapply(seq_len(nb_boot),
+                                              mc.cores = parallel::detectCores()-2, 
+                                              function(boot_i){
+                                                lapply(nb_aggregation,
+                                                       function(nb_aggregation_i){
+                                                         df_forecast %>%
[…]

### 2023-11-17 script/forecast/forecast_test.R
simplify test script

@@ -12,7 +12,6 @@ data_covid <- readRDS(file = "data/df_obfuscated_epidemio.rds")
 forecast_days = 14
 warmup = 30
 units = 500
-nb_hp_set = 1
 nb_iter = 40
 ##### LOAD BEST HP SETS
 hp_sets = list(common_input_scaling = "data/common_input_scaling_11423760/",
@@ -28,7 +27,7 @@ hp_sets = list(common_input_scaling = "data/common_input_scaling_11423760/",
   })
 
 best_hp_set = lapply(hp_sets,
-                     function(x) x %>% slice_min(mean_absolute_error, n = nb_hp_set))
+                     function(x) x %>% slice_min(mean_absolute_error, n = 1))
 ##### DEFINE EVALUATION SET
 vecDates <- data_covid %>%
   filter(START_DATE >= as.Date("2021-03-01") + forecast_days) %>%
@@ -36,139 +35,64 @@ vecDates <- data_covid %>%
 
 date_i <- vecDates[slar_taskid]
 message(paste0("--------- evaluation date : ", date_i, " ----------"))
-##### COMMON INPUT SCALING
-message("--------- COMMON INPUT SCALING ----------")
-forecast_common_is <- lapply(X = 1:nrow(best_hp_set$common_input_scaling),
-                             FUN = function(row_i){
-                               row = best_hp_set$common_input_scaling[row_i,]
-                               fct_iterative_forecast_from_hp(data_covid = data_covid,
-                                                              vecDates = date_i,
-                                                              warmup = warmup,
-                                                              forecast_days = forecast_days,
-                                                              units = units,
-                                                              lr = as.numeric(row[["leaking_rate"]]),
-                                                              sr = as.numeric(row[["spectral_radius"]]),
-                                                              ridge = as.numeric(row[["ridge"]]),
-                                                              input_scaling = as.numeric(row[["input_scaling"]]),
-                                                              seed = as.integer(row[["seed"]]),
-                                                              link_source = as.logical(row[["link_source"]]),
-                                                              model = "esn",
-                                                              nb_iter = nb_iter) %>%
-                                 mutate(hp_set = row[["hp_set"]])
-                             }) %>%
-  bind_rows() %>%
-  mutate(model = "esn_common_is")
-##### COMMON INPUT SCALING (linked source)
-message("--------- COMMON INPUT SCALING LS ----------")
[…]

### 2023-11-17 script/forecast/forecast_test.R
fix bug on eval

@@ -39,6 +39,7 @@ message(paste0("--------- evaluation date : ", date_i, " ----------"))
 ##### FORECAST
 dfres <- lapply(names(best_hp_set),
        function(model_set){
+         print(model_set)
          hp_set <- best_hp_set[[model_set]]
          if(model_set == "enet"){
            ##### ELASTIC-NET
@@ -58,12 +59,17 @@ dfres <- lapply(names(best_hp_set),
                     hp_set = 1)
            
            res <- bind_rows(forecast_enet_once, forecast_enet_daily)
-         } else {
+         }
+         
+         if(model_set %in% c("common_input_scaling",
+                             "common_input_scaling_linked_source",
+                             "multiple_input_scaling",
+                             "multiple_input_scaling_linked_source")){
            ##### RESERVOIR
            if(model_set %in% c("common_input_scaling", "common_input_scaling_linked_source")){
              input_scaling <- hp_set$input_scaling
-           } else if(model_set %in% c("multiple_input_scaling", "common_input_scaling_linked_source")){
-             input_scaling = row %>%
+           } else if(model_set %in% c("multiple_input_scaling", "multiple_input_scaling_linked_source")){
+             input_scaling = hp_set %>%
                dplyr::select(-c("hp_set",
                                 "ridge",
                                 "leaking_rate",
@@ -81,13 +87,13 @@ dfres <- lapply(names(best_hp_set),
                                                  sr = hp_set$spectral_radius,
                                                  ridge = hp_set$ridge,
                                                  input_scaling = input_scaling,
-                                                 seed = hp_set$seed,
                                                  link_source = hp_set$link_source,
                                                  model = "esn",
                                                  nb_iter = nb_iter) %>%
-             mutate(hp_set = row[["hp_set"]],
+             mutate(hp_set = hp_set$hp_set,
                     model = model_set)
-         } 
+         }
+         return(res)
        }) %>%
   bind_rows()
 

### 2023-11-17 draft_jss/draft_jss.Rmd
update rmd with new data

@@
 
-**TODO : extend this part and link to previous work**
-**Incorporate the following :**
-RC in high dimension
-
-The interest of RC in high dimensional setting is unclear for at least
-two reasons: (i) RC relies primarily on the projection of the low
-dimension input layer to the high dimension space of the reservoir
-neurons and (ii) the high number of features for which each can be
-associated with a different input scaling increases the hyperparameter
-search space making standard approaches such as random search or grid
-search inefficient.
-
 ## what we do
@@
 library(cowplot)
+library(patchwork)
 
@@
 
-`TODO :` add paragraph about RC in high dimension
-
-`TODO :` add paragraph about hp optimisation for RC (focus on GA). Discussed in https://ieeexplore.ieee.org/abstract/document/8481347
-
 RC can be approached as an extension of penalized linear regression where inputs are processed by a reservoir which allows memory and non-linear combinations.
@@
 
-### Hyperparameter optimisation using GA {#geneticalgo}
+### Hyperparameter optimisation using random search {#geneticalgo}
 
@@
 The pseudo-code of GA is defined as follow :
-
-1.  Initialize :
-
--   Population parameters : $N_{pop} = 200$, $N_e = 100$,
-    $N_{generation} = 30$
--   Cross-over parameters : $N_{tournament} = 2$
--   Mutation parameters : $p_{mutQuant} = .5$, $p_{mutCat} = .25$,
-    $\sigma = 1$ ($\sigma = 0.1$ for leaking rate as the hyperparameter
-    range is shorter)
--   Define the $hyperparameter\_list$, a list containing each
-    hyperparameter denoted as $param$.
--   `for` each $param$ in $hyperparameter\_list$ define hyperparameter
-    $search\_space$ :
-    -   the $lower$ and $upper$ bounds,
-    -   the $type$ (numeric, integer, categorical)
-    -   if it should be $log$ transformed (True, False)
-
-2.  Define main functions
-    -   Define $tournamentSelection(N_{pop}, N_{tournament})$:
-        -   Get the best $N_{pop}$ individuals from previous generations
-        -   Randomly select $N_{tournament}$ challengers among them
-        -   Get the best individuals from the challengers
-        -   Repeat twice to get a father and a mother
-        -   `return` $father, \: mother$
-    -   Define
-        $crossoverMutation(pere\$param\$value, mere\$param\$value, param\$search\_space)$:
-        -   `if` $search\_space\$log$ $\newline$
-            $\text{} \qquad pere\$param\$value = log_{10}(pere\$param\$value)$
-            $\newline$
-            $\text{} \qquad mere\$param\$value = log_{10}(mere\$param\$value)$
-
-        -   sample $alp$ and $mut$ probabilities from $\mathcal{U}(0,1)$
-
-        -   Cross-over, and define $child\$param\$value$ as :
-
-            -   Numerical :
-                $alp*pere\$param\$value + (1-alp)*mere\$param\$value$
-            -   Integer :
-                $round( alp*pere\$param\$value + (1-alp)*mere\$param\$value)$
-            -   Categorical :
-                $sample(x = c(pere\$param\$value, mere\$param\$value), \newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad prob = c(alp, 1-alp))$
-
-        -   `if` $mut < p_{mutQuant}$ or $mut < p_{mutCat}$ for
-            quantitative and categorical features respectively, mutate :
-
-            -   Categorical : sample another modality
-            -   Integer : add or substract 1 at random
-            -   Numerical : Add $\sigma", "imes \mathcal{N}(0,1)$
-
-        -   `if` $search\_space\$log$ :
-            $child\$param\$value = 10^{child\$param\$value}$
-
-        -   `return` $child\$param\$value$
-    -   Define $geneticSearch(N_{pop}, N_{tournament})$ :
-        -   $pere, \: mere = tournamentSelection(N_{pop}, N_{tournament})$
-        -   `for` $param$ in $hyperparameter\_list$ : $\newline$
-            $\text{} \qquad child\$param\$value = crossoverMutation(\newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad pere\$param\$value, \newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad mere\$param\$value, \newline\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad\text{}\qquad param\$search\_space)$
-        -   `return` $child$
-3.  Run the algorithm
-    -   Step 1 : Initialize genetic algorithm
-        -   Sample $N_{pop}$ individuals (i.e set of hyperparameters)
-            using random search
-    -   Step 2 : Run genetic algorithm
-        -   $generation = 1$
-        -   `while` $generation < Ngeneration$ :
-            -   $generation = generation + 1$
-            -   sample $N_{pop}$ individuals using $geneticSearch$
-                function
 
@@
 
-```{r enetNbFeatures, echo = FALSE, fig.cap = "Number of features depending elastic-net penalization.
-Data from the 100 best enet hp sets.
-Variability represented by boxplot are due to the penalization applied at different dates.", out.width="75%"}
-knitr::include_graphics(path = here::here("draft_jss/images/GA/fig-plot_enet_nb_features-1.png"))
-```
-
 ## Results
+
+```{r}
+ls_use_case <- readRDS(file = here::here("data/precomputed_results.rds"))
+```
 
@@
 
-```{r casestudyhospitalizations, echo = FALSE, fig.cap = "Hospitalizations at the university hospital of Bordeaux", out.width="75%"}
-knitr::include_graphics(path = here::here("draft_jss/images/monthly update/fig-eval-setting-1.png"))
+```{r ucPresentData, echo = FALSE, fig.cap = "Data of the use case.
+Outcome of interest is presented in orange.
+Model will be trained to forecast outcome variation but will be used to Outcome for performance and representation.
+Other features are represented in darkblue.", fig.height=10}
+ls_use_case$plot_present_data
 ```
@@
 
-#### Epidemiology subset with multiple input scaling
-
-```{r GAMultipleisEpidemioOptim, echo = FALSE, fig.cap = "Hyperparameter optimisation on epidemiological subet with multiple input scaling.
-2021-03-01 correspond to the result of approximately 3200 trials, subsequent dates corresponds to the result of approximately 1200 trials.
-For each optimisation date, only values of the 200 best individuals are shown.", fig.height=8}
-p1 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleis_epidemio/unnamed-chunk-19-1.png"))
-p2 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleis_epidemio/unnamed-chunk-20-1.png"))
-p3 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleis_epidemio/unnamed-chunk-21-1.png"))
-p4 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleis_epidemio/unnamed-chunk-22-1.png"))
-
-cowplot::plot_grid(p1, p2, p3, p4, nrow = 2, ncol = 2,
-                   labels = c("(a) Convergence",
-                              "(b) Marginal",
-                              "(c) Conditional",
-                              "(d) Feature importance (input scaling)"),
-                   label_size = 8,
-                   label_y = 1.01)
-```
-
-Figures \@ref(fig:GAMultipleisEpidemioOptim) show the marginal value of hyperparameters selected by GA over the different optimization time at panels (a) and (b).
-We note that both leaking rate and ridge values selected on the train set were not well adapted to the test set, they increased rapidly and then stabilized indicating a need leverage more information from new inputs compared to old inputs (leaking rate) and to decrease overfitting (ridge).
-The spectral radius progressively decreased over time which might indicate less importance for non-linear feature projection and a need for less chaotic reservoir.
-
-Panel (c) shows conditional values of hyperparameters.
-We observe that the outlined hyperparameters are independent one from another except for the linear association between leaking rate and ridge but which was observed only for the first updates.
-**I remember Pierrick told us that independance was a good property (probably because if two hp are correlated you can just focus on the optimisation of 1 and deduce the other) but we might need a reference here.**
-
-Panel (d) shows feature importance, depicted by the feature input scaling (i.e the higher the more important), over the different time period.
-It is interesting to note that feature importance vary over time and that important features before 2021-03-01 (IPTCC, derivative of hospitalizations) have an importance decrease of contrary to features related to fraction of positive RT-PCR in the elderly which remained important.
-**Need to change the features label of the figure.**
-
-#### GA feature selection with multiple input scaling and seed control
-
-```{r GAMultipleisGAOptim, echo = FALSE, fig.cap = "Hyperparameter optimisation with GA feature selection with multiple input scaling and seed optimisation.
-2021-03-01 correspond to the result of approximately 3200 trials, subsequent dates corresponds to the result of approximately 1200 trials.
-For each optimisation date, only values of the 200 best individuals are shown.", fig.height=8}
-p1 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleISseed_GA/unnamed-chunk-39-1.png"))
-p2 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleISseed_GA/unnamed-chunk-40-1.png"))
-p3 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleISseed_GA/unnamed-chunk-41-1.png"))
-p4 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/monthly update/GA_multipleISseed_GA/unnamed-chunk-42-1.png"))
-
-cowplot::plot_grid(p1, p2, p3, p4, nrow = 2, ncol = 2,
-                   labels = c("(a) Convergence",
-                              "(b) Marginal",
-                              "(c) Conditional",
-                              "(d) Feature selection frequency"),
-                   label_size = 8,
-                   label_y = 1.01)
-```
-
-Figure \@ref(fig:GAMultipleisGAOptim) show the marginal value of hyperparameters selected by GA over the different optimization time at panel (a) and (b).
-Leaking rate remained stable over time but again, spectral radius decreased and ridge increased indicating a lesser need for non-linear feature recombination and higher need to control for overfitting.
-Off note we outlined the seed hyperparameter here.
-Seed controls the randomness of RC connections and its value do not have epidemiological interpretation.
-Figure Panel (c) shows hyperparameter independence.
-
-Panel (d) shows feature selection over the different time period.
-It outlines that most important features (i.e the most often selected ones) are the ones related to RT-PCR, specifically for the elderly.
-We also note the importance of features related to emergency and ambulance departments.
-In addition, we observe that feature importance did not remained constant over time justifying the need for an adaptive feature selection algorithm.
-
-### Anteriority
-
-**TODO : if we keep this section, we should update with same hyperparameter (feature selection with only 30 features for GA multiple input scaling) and same aggregation mechanism as in the present forecast performance to have something clean.**
-
-Because epidemic mechanism change over time, it could of interest not to provide all the time series to the model but only its most recent data.
-We investigate whether the data anteriority we used to train and test the model had an impact on its performance.
-Because the hyperparameter optimization process is long, the hyperparameter used were optimized on all the time series (i.e 542 days).
-
-Results of the experiment are outlined at figure \@ref(fig:anteriorityPerformance).
-We show that at least 1 year of data seems necessary to leverage enough information.
-This is what will be used in the remaining sections.
-**Note that epidemio with one input scaling had very good performance on this evaluation.**
-
-```{r anteriorityPerformance, echo = FALSE, fig.cap = "Performance depending on number of days used to train the model.
-10 iterations with different seed for each of 10 best hyperparameters sets.
-Boxplots corresponds to individual models performance (no aggregation) and diamond is the median forecast of the 100 models.", out.width="75%"}
-knitr::include_graphics(path = here::here("draft_jss/images/anteriority/fig-plot_anteriority-test-set-1.png"))
+```{r ucHyperparam, fig.cap="Hyperparameter evaluation on training set by random search. Hp sets with MAE above 30 were removed for clarity of visualisation.", fig.height=12}
+ls_use_case$plot_reservoir_hp / ls_use_case$plot_feature_input_scaling + plot_annotation(tag_levels = 'A')
 ```
@@
 
-```{r modelBudget, echo = FALSE, fig.cap = "Performance dependending on number of hyperparameter sets (model) and the scenario used.
-Panel (a) explores two scenario : one where the best hyperparameter set is replicated x times with different seed, one where the x best hyperparameter sets are used with one seed each.
-Each scenario is evaluated with the aggregation of 1, 3, 5, 10, 20, 30, 40, 60 or 100 models with 250 bootstrap 95\\%CI Panel (b) explores the optimal balance between both approaches with a budget of 40 models.", fig.height=8}
-p1 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/budget/fig-plot_aggregation_convergence-1.png"), scale = 0.9)
-p2 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/budget/fig-plot_aggregation_convergence_budget-1.png"), scale = 0.9)
-
-cowplot::plot_grid(p1, p2, nrow = 2, ncol = 1,
-                   labels = c("(a) Number of models needed",
-                              "(b) Optimal budget repartition"),
-                   label_size = 8,
-                   label_y = 1.01)
-```
-
-```{r best40epidemio, echo = FALSE, fig.cap = "Performance of each of the best 40 hp sets with 40 replicate each. 250 Bootstrap 95\\%CI. Epidemiology subset with one input scaling.", out.width="75%"}
-knitr::include_graphics(path = here::here("draft_jss/images/budget/fig-plot_aggregation_best40epidemio-1.png"))
-```
-
-Figure \@ref(fig:modelBudget) shows the model performance depending on the model and the aggregation process.
-We observe that for all models, the model performance seems achieved at 30 models, therefore, for remaining results, 40 models will be used.
-All models showed that the best aggregation mechanism was to use different hyperparameter sets instead of using the best hyperparameter set with different seeds.
-The exception was the epidemiology subset with single input scaling whose best performance was achieved using only the first two models.
-This is explained by the low performance of the fourth best hyperparameter set on the test set as depicted at figure \@ref(fig:best40epidemio).
+```{r ucAggregation, echo = FALSE, fig.cap = "Aggregation of several reservoir.
+Due to its intrisic chaoticity, a same hyperparameter set can results in heterogeneous forecast.
+Panel A shows the individual forecast of 40 repetition of the same hyperparameter set for the different RC configuration.
+Panel B shows the impact of aggregating several RC with the same set of HP.
+It has small interest when the input layer was not connected to output layer but significantly improves the performance otherwise.", fig.height=8}
+ls_use_case$plot_before_aggregation_forecast + ls_use_case$plot_aggregation + plot_annotation(tag_levels = 'A')
+
+```
 
@@
 ```{r perfMonthlyUpdate, echo = FALSE}
-tibble::tribble(
-  ~"Update", ~"Feature", ~"Nb. Features", ~"MAE", ~"MRE", ~"MAEB", ~"MREB ",
-  "No", "Epi - Single IS", 14.0, 13.79, 0.34, -2.92, 0.77 ,
-  "Yes", "Epi - Single IS", 14.0, 10.16, 0.32, -5.09, 0.67 ,
-  "No", "Epi - Multiple IS", 14.0, 9.35, 0.28, -5.31, 0.66 ,
-  "Yes", "Epi - Multiple IS", 14.0, 11.83, 0.31, -2.03, 0.80 ,
-  "No", "GA - Single IS", 231.0, 10.62, 0.28, -5.05, 0.72 ,
-  "Yes", "GA - Single IS", 225.0, 9.33, 0.28, -4.76, 0.72 ,
-  "No", "GA - Multiple IS", 227.0, 9.67, 0.33, -3.64, 0.71 ,
-  "Yes", "GA - Multiple IS", 227.5, 9.45, 0.29, -3.90, 0.69 ,
-  "No", "GA - Multiple IS (seed)", 225.0, 10.68, 0.27, -4.37, 0.72 ,
-  "Yes", "GA - Multiple IS (seed)", 227.5, 10.20, 0.28, -5.33, 0.68 ,
-  "No", "Enet - Single IS", 172.0, 11.72, 0.31, -3.29, 0.79 ,
-  "Yes", "Enet - Single IS", 208.0, 13.68, 0.35, -2.58, 0.81) %>%
-  knitr::kable(caption = "Model performance with or without monthly update (Update), feature selection (Feature).
+ls_use_case$table_perf_esn %>%
+  knitr::kable(caption = "Model performance with several reservoir configuration.
 For each setting, 40 reservoir are computed and the forecast is the median of the 40 forecast.
 Results show the mean number of features and the performance metrics.
-MAE = Median Absolute Error, MRE = Median Relative Error, MAEB = Median Absolute Error to Baseline, MREB = Median Relative Error to Baseline.",
+MAE = Mean Absolute Error, MRE = Median Relative Error, MAEB = Mean Absolute Error to Baseline, MREB = Median Relative Error to Baseline.",
+               col.names = c("Model", "MAE", "MRE", "MAEB", "MREB"),
                booktabs = TRUE,
-               linesep = c('', '\\addlinespace'))
+               linesep = c('', '\\addlinespace'), digits = 2)
 ```
@@
 Grey lines are individual forecast of each of the 40 reservoir.", fig.height=8}
-p1 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/monthly update/fig-eval-prediction-noupdate-1.png"))
-p2 <- cowplot::ggdraw() +
-  cowplot::draw_image(here::here("draft_jss/images/monthly update/fig-eval-prediction-update-1.png"))
-
-cowplot::plot_grid(p1, p2, nrow = 2, ncol = 1,
-                   labels = c("(a) No update",
-                              "(b) Monthly update"),
-                   label_size = 8,
-                   label_y = 1.01)
+ls_use_case$plot_forecast
 ```
@@
 
-```{r perfPreviousExpNoFeatureSelection, echo = FALSE}
-tibble::tribble(
-  ~"Update", ~"Feature", ~"Nb. Features", ~"MAE", ~"MRE", ~"MAEB", ~"MREB ",
-  "No", "Epi - Single IS", 14.0, 12.08, 0.26, -3.04, 0.76,
-  "No", "Epi - Multiple IS", 14.0, 11.83, 0.28, -4.96, 0.67,
-  "No", "Epi - Multiple IS*", 14.0, 10.73, 0.28, -5.47, 0.65,
-  "No", "No FS - Single IS", 456.0, 9.38, 0.24, -5.23, 0.64,
-  "No", "No FS - Single IS*", 456.0, 12.16, 0.28, -4.38, 0.71,
-  "No", "Enet - Single IS", 153.5, 12.51, 0.30, -4.76, 0.76,
-  "No", "GA - Multiple IS limited to 30 features", 20.5, 10.47, 0.27, -3.03, 0.75,
-  "No", "CMAES - Multiple IS", 374.0, 13.91, 0.26, -2.62, 0.83
-  ) %>%
-  knitr::kable(caption = "Model performance depending on feature selection (Feature).
-For each setting, several reservoir are computed and the forecast is the median of the 40 forecast.
-Reservoir aggregation could be either to use the 10 best hyperparameter set once resulting in 10 models or to replicate each of the 10 best hyperparameter set 10 times (those models are denoted by a *) resulting in 100 models.
-Feature selection based on GA and CMAES use top k features with the highest input scaling instead of an indicator for each feature as previously depicted and GA was limited to 30 features.
-Models depicted as 'No FS' are models without feature selection.
-Results show the mean number of features and the performance metrics.
-MAE = Median Absolute Error, MRE = Median Relative Error, MAEB = Median Absolute Error to Baseline, MREB = Median Relative Error to Baseline.",
-               booktabs = TRUE,
-               linesep = c('\\addlinespace'))
-```
-
-Table \@ref(tab:perfPreviousExpNoFeatureSelection) shows additional results from previous experiences without monthly update of hyperparameter.
-The performance without feature selection are good compared to other results.
-Maybe we should give it a try and add it to table \@ref(tab:perfMonthlyUpdate) ?
-
 ## Discussion
 
-# Previous work
-
-Those work were done previously, the setting of train/test and hyperparameter optimisation were different and probably they should be done again in a similar setting for relevant comparison.
-
-## Raw outcome, first derivative or second derivative ?
-
-<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/explore_esn.md>
-
-## Raw feature or scaled feature ?
-
-<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/explore_esn.md>
-
-## Smoothed or not smoothed
-
-<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/explore_esn.md>
-
-## Number of nodes
-
-<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/explore_esn_r_pipeline.md>
-
-## Use of second derivative among the interesting features
-
-<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/00_reporting/tune_esn2.md>
-
-## Results of random sampling with R
-
-<https://github.com/thomasferte/PredictCovid/blob/esn_update_report/reporting/02_esn/01_explore_esn_prediction/meeting_1801/master_report.md>
-
 # Discussion and conclusion

### 2023-11-17 script/forecast/forecast_test_eval.R

@@ -1,11 +1,19 @@
+##### load packages and functions
 library(dplyr)
 library(ggplot2)
 
-ls_files_results <- list.files("data/results_forecast_testset_11463695/", full.names = TRUE)
+source("functions/fct_smoothing_derivative.R")
+
+##### load data
+ls_files_results <- list.files("data/results_forecast_testset_11495493/", full.names = TRUE)
 ls_files_results_res <- grep(ls_files_results, pattern = "hp_sets", value = TRUE, invert = TRUE)
 hp_file <- grep(ls_files_results, pattern = "hp_sets", value = TRUE)
+df_obfuscated_epidemio <- readRDS("data/df_obfuscated_epidemio.rds")
 
-##### hyperparameters plots
+##### labels and features
+hp_features_enet <- c("lambda",
+                      "alpha")
+
 hp_features_reservoir <- c("input_scaling",
                            "spectral_radius",
                            "leaking_rate",
@@ -28,6 +36,8 @@ hp_features <- c("hosp",
                  "Vaccin_1dose_rolDeriv7")
 
 feature_labels <- conservation_status <- c(
+  "outcome" = "Outcome",
+  "outcomeDeriv" = "Outcome variation",
   "input_scaling" = "Input scaling",
   "leaking_rate" = "Leaking rate",
   "ridge" = "Ridge",
@@ -54,9 +64,33 @@ model_labels <- c(
   "common_input_scaling" = "Common IS \n\n R %>>% output",
   "common_input_scaling_linked_source" = "Common IS \n\n input + R %>>% output",
   "multiple_input_scaling" = "Multiple IS \n\n R %>>% output",
-  "multiple_input_scaling_linked_source" = "Multiple IS \n\n input + R %>>% output"
+  "multiple_input_scaling_linked_source" = "Multiple IS \n\n input + R %>>% output",
+  "enet" = "Elastic-net",
+  "enet_hp_on_train_set" = "Elastic-net",
+  "enet_daily_hp_update" = "Elastic-net (alpha = 0.5, lambda = daily updated)"
 )
 
+##### present data
+data_i <- fct_smoothing_derivative(data = df_obfuscated_epidemio,
+                                   maxOutcomeDate = df_obfuscated_epidemio$START_DATE %>% max,
+                                   forecast_days = 14)
+
+plot_present_data <- data_i %>%
+  select(-START_DATE) %>%
[…]

### 2023-11-27 draft_jss/draft_jss.Rmd
update draft jss

@@
 
-## Reservoir networks presentation
-
-Reservoir Computing (RC) is a prominent machine learning methodology
+Reservoir Computing (RC) is a prominent machine learning methodology proposed by 
 that has gained significant attention in recent years for its ability to
-effectively process information generated by dynamical systems. This
+effectively process information generated by dynamical systems [@kim_health_2020, @maass_real-time_2002]. This
 innovative approach leverages the dynamics of a high-dimensional
 reservoir to perform complex computations and solve various tasks based
-on the response of this dynamical system to input signals. Reservoir
-Computing is characterized by its unique architecture, which includes a
-dynamic system, often referred to as a reservoir, that projects temporal
-input signals onto a high-dimensional feature space. This
-high-dimensional representation of input data enables Reservoir
-Computing to excel in tasks such as time series prediction, pattern
-recognition, and signal processing.
+on the response of this dynamical system to input signals.
+RC has demonstrated its efficacy in tackling various challenges, encompassing pattern classification and time series forecasting in applications ranging from electrocardiogram analysis to bird calls, power plants, internet traffic, stock prices, and beyond [@tanaka_recent_2019].
 
-**TODO : cite : Jaeger, Herbert.
-“The “echo state” approach to analysing and training recurrent neural networks-with an erratum note.” Bonn, Germany: German National Research Center for Information Technology GMD Technical Report 148.34 (2001): 13.**
-**Maass W, Natschläger T, Markram H. Real-time computing without stable states: a new framework for neural computation based on perturbations. Neural Comput. 2002;14(11):2531‐2560.**
+RC is a kind of recurrent neural networks (RNNs) where the internal states, denoted as the reservoir, is randomly generated and solely the final layer is trained.
+The reservoir projects temporal input signals onto a high-dimensional feature space facilitating the learning of non-linear and temporal interactions.
+This offers the advantage of decreasing the computing time while consistently maintaining robust performance compared to conventional RNNs [@shahi_prediction_2022, @vlachas_backpropagation_2020].
 
-## Why reservoir are interesting compared to standard approaches (penalised regression) and deep learning approaches?
+The utilization of RC poses challenges on multiple fronts.
+Firstly, its stochastic nature, attributed to the random aspect of the reservoir, introduces unpredictability into the results, even with well-chosen hyperparameters.
+Secondly, RC relies on numerous hyperparameters to establish connections between neurons.
+Despite this, guidance on determining these hyperparameters is limited, often requiring users to adopt a wrapper approach where combinations of hyperparameter performance are assessed on a training set before selecting the optimal combination.
+Thirdly, there is a current absence of implementation in R, rendering the method challenging for users unfamiliar with Python [@trouvain_reservoirpy_2020, @steiner_pyrcn_2022] or Julia [@martinuzzi_reservoircomputingjl_2022].
 
-RC exhibits several advantages over its counterparts. It excels in tasks
-that require the processing of temporal data, such as time series
-forecasting and speech recognition, thanks to its inherent memory
-capabilities. RC offers computational efficiency and reduced susceptibility
-to overfitting compared to deep learning. Moreover, RC is able to capture intricate temporal
-patterns compared to statistical methods such as penalized regression.
-
-**TODO : find ref to support it.**
-
-## examples of previous use
-
-**TODO: add ref from use case intro on use in epidemiology + find review**
-**Table 1 contains a list of use case of RC : https://www.sciencedirect.com/science/article/pii/S0893608019300784**
-
-## challenges of using reservoir
-
-RC use is still challenging for several reasons. First, its stochastic
-nature due to the reservoir random part makes the results of a given RC
-unpredictable even if the hyperparameters are well chosen. Second,
-even though, RC relies on several hyperparameter to determine the
-connections between neurons. There exist few guidance about the way to
-set the hyperparameters and most often the user must rely on a wrapper
-approach were combinations of hyperparameters performance are evaluated
-on a train set before choosing the ideal combination one. Third, most
-published paper are based on a low dimensional setting and few guidance
-exist in the context of high dimensional data, especially considering
-the choice of hyperparameters. Fourth, the use of reservoir computing in
-an epidemic setting with highly non-stationary time series is scarce
-whereas it is of tremendous interest for epidemiologists. Fifth, there
-is currently no implementation available on R making the use of this
-method challenging for user not familiar with Python.
-
-## what we do
-
-In this paper we propose to explore those challenges with extensive
-guidance in order to help new user to fully benefit from RC. First a
-general introduction to reservoir computing is provided with a tutorial
-for its use using `reservoirnet`, a R package based on `reservoirPy` Python
-module. Second, we explore the different challenges of the use of RC in
-a non-stationary high dimensional setting based on the use case of
-covid-19 hospitalizations forecast with extensive guidance on the
-modelling strategy, the choice of hyperparameters using a genetic
-algorithm and the implementation.
+In this paper, we endeavor to delve into these challenges, offering comprehensive guidance to assist new users in maximizing the benefits of RC.
+Initially, a broad introduction to reservoir computing is presented, accompanied by a tutorial on its application using `reservoirnet`, an R package built upon the `reservoirPy` Python module [@trouvain_reservoirpy_2020].
+Subsequently, we investigate the various challenges associated with the use of RC for forecasting COVID-19 hospitalizations.
+This exploration includes detailed guidance on the modeling strategy, the selection of hyperparameters, and the implementation process.
 
@@
 The parameter $\lambda$ is the ridge penalization which aims to prevent
-overfitting.
+overfitting. One can also connect the input layer to the output layer in addition to the reservoir nodes. In that case, $X$ is the accumulation of both such that :
+
+$$X = \begin{bmatrix} x(1), u(1) \\ x(2), u(2) \\ ... \\ x(T), u(T) \end{bmatrix}
+\text{ and } Y = \begin{bmatrix} y(1) \\ y(2) \\ ... \\ y(T) \end{bmatrix}$$
 
 Overall, there are four main hyperparameters to be chosen by the user: the leaking rate which defines the memory of the RC, the input scaling which define the relative importance of the features, the spectral radius which define the connections of the neurons inside the reservoir which in turn define the degree of non-linear combination of features and the ridge penalization which controls the degree of overfitting.
-The choice of hyperparameter is difficult and the use of a wrapper approach where the performance of the RC with different combinations of hyperparameters is evaluated on a train set, the best combination is chosen for the use on the test set.
+The choice of hyperparameter often rely on a wrapper approach where the performance of the RC with different combinations of hyperparameters is evaluated on a train set and the best combination is chosen to forecast on the test set.
 
@@
 
-We can proceed by loading useful packages (i.e ggplot2 [@wickham_ggplot2_2016] and dplyr [@wickham_dplyr_2023], **TODO : add cowplot**), data and define the task:
+We can proceed by loading useful packages (i.e ggplot2 [@wickham_ggplot2_2016], dplyr [@wickham_dplyr_2023] and patchwork [@pedersen_patchwork_2023]), data and define the task:
 
@@
 library(ggplot2)
+library(patchwork)
 library(reservoirnet)
-library(cowplot)
-library(patchwork)
 
@@
 
-`TODO :` check for new published forecast models
+RC can be viewed as an extension of penalized linear regression, where inputs undergo processing by a reservoir, introducing the capacity for memory and non-linear combinations.
+Given the effectiveness of penalized linear regression in COVID-19 forecasting, as highlighted in @ferte_benefit_2022, and the promising results exhibited by RC in epidemic forecasting, as demonstrated in studies such as @ghosh_reservoir_2021, @liu_nanophotonic_2023, and @ray_optimized_2021, we have opted to employ RC for the prediction of hospitalizations at 14 days at the University Hospital of Bordeaux.
 
-RC can be approached as an extension of penalized linear regression where inputs are processed by a reservoir which allows memory and non-linear combinations.
-As penalized linear regression was the best approach for covid-19 forecast in @ferte_benefit_2022 and that RC showed promising results for epidemic forecast in @ghosh_reservoir_2021, @liu_nanophotonic_2023 and @ray_optimized_2021, we proposed to use RC to forecast hospitalizations at 14 days at the University Hospital of Bordeaux.
-The main objective of this study is to evaluate the performance of RC for this task.
-Secondary objective will be (i) to assess the performance of RC depending on the input dimension, (ii) to assess the relevance of Genetic Algorithm (GA) for hyperparameter optimization and feature selection in this context.
+The primary aim of this study is to assess the performance of RC in this forecasting task.
+Secondary objectives include (i) comparing the performance of RC with that of elastic-net penalized regression (identified as the optimal model in @ferte_benefit_2022) and (ii) evaluating variations in RC performance based on different architectural choices, such as the connection between the input layer and the output layer, and the use of one input scaling per feature versus a common input scaling.
 
@@
 
-First and second derivative over the last $7$ days were computed to enrich model information.
+First derivative over the last $7$ days were computed to enrich model information.
 To take into account measurement error and daily noise variation, data were smoothed using a local polynomial regression with a span of 21 days.
+
+All data are publicly available.
+Weather data can be obtained from @smith_integrated_2011 using R package `worldmet` [@carslaw_worldmet_2023].
+[Vaccine data](https://www.data.gouv.fr/fr/datasets/donnees-relatives-aux-personnes-vaccinees-contre-la-covid-19-1/) can be downloaded from @etalab_les_2020.
+EHRs data can be downloaded on dryad [@ferte_benefit_2023].
+For privacy issues, publicly available EHRs data below 10 patients were obfuscated to 0.
+For convenience, all data were downloaded, merged and provided as replication material.
 
@@
 The task was to forecast 14 days ahead the number of hospitalized patients.
-As seen at section [Regression](#basicregression), we will train the model to predict the variation of hospitalization defined as $outcome_{t+14} = hosp_{t+14} - hosp_{t}$.
-Metrics computation and visualizations will be performed on the predicted number of hospitalizations denoted as $\hat{y_{t+14}} = \widehat{outcome_{t+14}} + hospitalizations_{t}$ and the observed number of hospitalizations at time $t+14$ denoted as $y_{t+14} = hospitalizations_{t+14}$
+As seen at section [Regression](#basicregression), we will train the model to predict the variation of hospitalization, denoted as $hosp$, defined as $outcome_{t+14} = hosp_{t+14} - hosp_{t}$.
+Metrics computation and visualizations will be performed on the predicted number of hospitalizations denoted as $\widehat{hosp_{t+14}} = \widehat{outcome_{t+14}} + hosp_{t}$.
 
-The dataset was separated into two periods.
-First period from May 16, 2020 to March 1, 2021 served to identify relevant hyperparameters.
-Remaining data was used to evaluate the model performance.
-As the time series is not stationary and most appropriate hyperparameter value might change over time, we also investigated the effect of hyperparameter update every month.
+The dataset was separated into two periods. First period from May 16, 2020 to March 1, 2021 served to identify relevant hyperparameters. Remaining data was used to evaluate the model performance.
 
-The performance of the model was evaluated according to several metrics: the median absolute error ($MAE = median(|\hat{y_t}-y_t|)$) , the median relative error ($MRE = median(|\frac{\hat{y_t}-y_t}{y_t}|)$), the median absolute error to baseline ($MAEB =  median(|\hat{y_t}-y_t| - |y_{t-14} - y_t|)$), the median relative error to baseline ($MREB = median(\frac{|\hat{y_t}-y_t|}{|y_{t-14} - y_t|})$).
-Such that the model denoted as baseline was the model defined as $\hat{y_t} = y_{t-14}$.
-All those metrics reflects different information.
-$MAE$ and $MAEB$ are highly sensitive to forecast error when the number of hospitalizations is high.
-$MAEB$ encapsulates information about the current dynamic of the hospitalizations : for the same absolute error, absolute error to baseline will be lower if the number of hospitalization is not constant.
-**MAEB = redundant with MAE ?**.
-$MRE$ is highly sensitive to forecast error when the number of hospitalizations is low because the denominator will be small.
-$MREB$ is highly sensitive to small forecast error when the number of hospitalized patients is constant over time.
-For each metric, the median was chosen over the mean to be robust to outliers.
-For instance, for relative error, when the outcome is low (e.g equal to 1 hospitalization), an error of 5 hospitalizations could results to a 400% relative error.
-**But that might favor models with chaotic metric because extreme values won't be used to compute the error ?
-=> keep median for relative error and mean of absolute error and optimize on MAE instead of MREB ?
-Also, because MAE is a difference it is similar if we compute it on the number of hospitalization or on the variation of the number of hospitalization.
-Also we might discuss to penalize more the error when the number of hospitalizations is high (already done implicitly by AE) and when there is a dynamic shift => ponderate AE by absolute value of second derivative ?**.
+The performance of the model was evaluated according to several metrics:
 
-### Feature selection with RC
+- the mean absolute error : $MAE = mean(|\widehat{hosp_{t+14}}-hosp_{t+14}|)$.
+- the median relative error : $MRE = median(|\frac{\widehat{hosp_{t+14}}-hosp_{t+14}}{hosp_{t+14}}|)$.
+- the mean absolute error to baseline : $MAEB =  mean(|\widehat{hosp_{t+14}}-hosp_{t+14}| - |hosp_{t}-hosp_{t+14}|)$.
+- the median relative error to baseline : $MREB = median(|\frac{\widehat{hosp_{t+14}}-hosp_{t+14}}{hosp_{t}-hosp_{t+14}}|)$)
 
-We investigated the performance of RC using various feature selection approach and input scaling configuration. The following setting were evaluated :
+Mean was chosen over median for $MRE$ and $MREB$ because those metrics tend to have extremely high values when the denominator is close to 0 (i.e when the number of hospitalized patients is close to 0 or the number of patients hospitalized at 14 days is close to the current number of hospitalized patients respectively).
+$MAEB$ and $MREB$ compare model performance to a baseline model which predicts the current number of hospitalized patients at 14 days.
+Those metrics help to determine the information added by the model and is a good baseline as covid-19 forecast model do not always outperform this basic forecast (@cramer_evaluation_2022).
 
-- Feature selection based on expert inputs with common input scaling
-- Feature selection based on expert inputs with specific input scaling for each feature
-- Genetic algorithm feature selection with common input scaling
-- Genetic algorithm feature selection with specific input scaling for each feature
-- Elastic-net feature selection with common input scaling
+### Models
 
-Experts inputs were a priori defined features that were judge important for epidemic forecast : hospitalizations, positive RT-PCR, positive RT-PCR among the 60 years old and more, fraction of positive RT-PCR, fraction of positive RT-PCR among the 60 years old and more, IPTCC, number of patient EHR mentioning "covid-19" in emergency units department, number of people having received at least one dose of vaccine in Gironde.
-Additionally we provided the first derivative at 7 days for each feature.
+We compared RC to elastic-net penalized regression (denoted as Enet).
+Furthermore we evaluated RC based on several architectures.
+First we compared RC with a single input scaling common to all features and a RC with on specific input scaling per feature.
+Second we compared RC where the input layer is connected to the output layer in addition to the connection between reservoir and output layer.
+Therefore, five models were evaluated :
 
-We used feature selection using elastic-net because it was the best model in previous experiment compared to other machine learning (@ferte_benefit_2022).
-This feature selection impose to select the penalty hyperparameter of elastic-net.
-This hyperparameter was selected using genetic algorithm as depicted [previously](#geneticalgo).
+- Elastic-net penalized regression denoted *Enet*
+- RC with a single input scaling and no connection between input and ouput layers denoted *Common IS R %>>% O*
+- RC with a single input scaling and connection between input and ouput layers denoted *Common IS I+R %>>% O*
+- RC with multiple input scaling and no connection between input and ouput layers denoted *Multiple IS R %>>% O*
+- RC with multiple input scaling and connection between input and ouput layers denoted *Multiple IS I+R %>>% O*
 
-Genetic algorithm feature selection was performed using a binary hyperparameter for each feature determining if it was to be selected or not.
+Because of the randomness of the reservoir, we took the median forecast of 20 reservoir on the train set and 40 reservoir on the test set.
+In addition, because covid-19 hospitalization is a non-stationary process, models were re-trained everyday using all previous days.
+To ease computation burden, only one day over two was used to find hyperparameters on the training set.
 
-In addition, we investigated if monthly update of hyperparameter was improving the performance and if optimizing the random seed (which defines the connections inside the reservoir) was improving the results.
-
-### Hyperparameter optimisation using random search {#geneticalgo}
+### Hyperparameter optimisation using random search
 
@@
 Input scaling can be either, common to all features or specific to each feature which increases the number of hyperparameter by the number of features.
-As the relevancy of feature for forecast evolves over time, we explored (i) filter based feature selection based on elastic-net before RC which has a penalization hyperparameter denoted as $enet$ and (ii) wrapper feature selection for which each feature has an additional binary hyperparameter defining if it should be selected or not.
 
-Regarding the number of hyperparameter, an automatic approach is necessary.
-We chose to use a genetic algorithm (GA) which has the benefit of making few assumptions on the link function between the vector of hyperparameter value and the objective function but might need many trials which are possible due to the fast of RC (**TODO : add ref**).
-The objective function of GA was to minimize the $MREB$ metric **(change ?)**.
-The pseudo-code of GA is defined as follow :
+Following the notation from `glmnet` package [@friedman_regularization_2010], elastic-net penalized linear regression relies on two hyperparameters, lambda (i.e the penalization parameter) and alpha (i.e the compromise between lasso and ridge penalty)
 
-For each hyperparameter, we must define the boundaries of the search space.
-We chose a large search space because we had few hypothesis to restrict it.
-Spectral radius and input scaling were defined on $[10^{-5} ; 10^5]$, leaking rate on $[10^{-5} ; 1]$, ridge on $[10^{-10} ; 10^5]$.
+Hyperparameter were selected in the training set (i.e before March 1, 2021) using a wrapper approach and a random search sampler using 2000 samples for each model.
+The sampling distribution were defined as follow :
 
-The penalization for elastic-net feature selection was initially defined on $[10^{-10} ; 10^5]$.
-However, first experiments showed that, after few generations, GA was stuck in a area were no feature selection was performed and without the ability to sample new individuals with feature selection.
-This phenomena is likely to append for evolutionary algorithm where the sampling variance is small compared to the search space.
-This phenomena is depicted at figure \@ref(fig:enetNbFeatures) were there is no feature selection performed when penalization is below $10^{-4}$.
-To mitigate this problem, we chose to restrict the search space on $[10^{-4} ; 10^5]$.
+- (RC) ridge and (Enet) lambda : log-uniform law defined between 1e-10 and 1e5
+- (RC) input scaling and spectral radius : log-uniform law defined between 1e-5 and 1e5
+- (RC) leaking rate : log-uniform law defined between 1e-3 and 1
+- (Enet) alpha : uniform defined between 0 and 1
+
+We provided large search space for all hyper-parameters.
+Search space was slightly reduced for leaking rate based on previous results and because a leaking rate of 1e-3 already imply that new inputs make the reservoir change really slowly which is not inline with the dynamic of covid-19 but would be appropriate for an application where the phenomena to forecast has a slow dynamic.
+
+Finally, we provided an additional Enet model similar to the one in @ferte_benefit_2022 where alpha was set to $0.5$ and lambda was re-evaluated everyday in the test set based on previous data using the cross-validation procedure provided by `glmnet`.
 
@@
 
-```{r}
+```{r echo=FALSE}
 ls_use_case <- readRDS(file = here::here("data/precomputed_results.rds"))
@@
 
-**TODO : change optimisation metric = mean absolute error ?**
-**TODO : optimise the seed ?**
-**TODO : compare with elastic-net alone ?**
-**TODO : compare with LSTM ?**
-**TODO : improve optimisation figures (too small)**
-**TODO : do everything with the R API ?**
-**TODO : provide clean replication code as a supplementary material (too long to be run inside a rmarkdown chunk)**
+The goal of this task is to predict 14 days ahead the hospitalization.
+Figure \@ref(fig:ucPresentData) shows both the training set (i.e before 2021-03-01) and the test set where the blue curve correspond to the input features (first derivatives are not shown) and the orange curves correspond to the outcome the model is trained on (i.e the hospitalization variation) and the hospitalizations at 14 days on which the performance metrics are computed.
+The figures outline that the relation between the input features and the outcome evolve over time and that the time series is not stationary.
+For instance IPTCC seems correlated to the outcome except that it completely miss the summer 2021 increase.
 
-The goal of this task is to predict 14 days ahead the hospitalization curve depicted at figure \@ref(fig:casestudyhospitalizations).
-It shows both the training set (i.e before 2021-03-01) and the test set encompass several slope shifts which are difficult to forecast.
-We will first introduce the results of the genetic algorithm hyperparameter optimization and then present the forecast performance.
-
-```{r ucPresentData, echo = FALSE, fig.cap = "Data of the use case.
+```{r ucPresentData, echo = FALSE, fig.cap = "Covid-19 epidemic at BUH.
 Outcome of interest is presented in orange.
-Model will be trained to forecast outcome variation but will be used to Outcome for performance and representation.
-Other features are represented in darkblue.", fig.height=10}
+Model is trained to forecast Outcome curve which corresonds to the difference between Hospitalisatiosn at 14 days and current hospitalisations.
+Other features are represented in darkblue.", fig.height=8}
 ls_use_case$plot_present_data
@@
 
-### Hyperparameter optimisation
+### Hyperparameter selection
 
-**TODO : add final choice of hyperparameter value**
+Figure \@ref(fig:ucHyperparam) shows the hyperparameter optimisation using random search for the different RC architectures.
+We note that the multiple input scaling setting with input layer connected to output layer showed poor performance on the training set which might be due to an unextensive search of hyperparameter regarding the better performance of the common input scaling with connected input layer which is a specific configuration where all input scaling are equal.
 
-In this section we will first introduce the results for the features selected based on expert knowledge with multiple input scaling and for the features selected by genetic algorithm with multiple input scaling and seed optimization.
+As expected, we observe that the optimal leaking rate is close to 1 for all RC which is coherent with the short term dynamic of covid-19 epidemic.
+Trend for other hyperparameter are less clear even though best hyperparameter sets were close for RC with common input scaling.
 
-```{r ucHyperparam, fig.cap="Hyperparameter evaluation on training set by random search. Hp sets with MAE above 30 were removed for clarity of visualisation.", fig.height=12}
-ls_use_case$plot_reservoir_hp / ls_use_case$plot_feature_input_scaling + plot_annotation(tag_levels = 'A')
+```{r ucHyperparam, echo=FALSE, fig.cap="Hyperparameter evaluation on training set by random search. Hp sets with MAE above 30 were removed for clarity of visualisation.", fig.height=8}
+ls_use_case$plot_reservoir_hp
 ```
 
-### Number of model to aggregate
+Figure \@ref(fig:ucHyperparamMultipleIS) shows the hyperparameter search for RC with multiple input scaling and connected input layer.
+We observe that the random search tend to favor low importance given to vaccine and high importance given to derivative of hospitalisations and derivative of positive RT-PCR in the elderly.
+The rest of the feature do not show clear pattern.
+The lack of importance of number of people with one dose of vaccine in this setting was discussed by @ferte_benefit_2022 and related to the lack of vaccinated people in the training set as depicted at figure \@ref(fig:ucPresentData).
 
-**TODO : if we keep this section, we should update with same hyperparameter (feature selection with only 30 features for GA multiple input scaling) and same aggregation mechanism as in the present forecast performance to have something clean.**
-**Note : if we optimize the seed, then the repetition of a given set of hyperparameter makes no sense (i.e will provide same results each time).**
-
-```{r ucAggregation, echo = FALSE, fig.cap = "Aggregation of several reservoir.
-Due to its intrisic chaoticity, a same hyperparameter set can results in heterogeneous forecast.
-Panel A shows the individual forecast of 40 repetition of the same hyperparameter set for the different RC configuration.
-Panel B shows the impact of aggregating several RC with the same set of HP.
-It has small interest when the input layer was not connected to output layer but significantly improves the performance otherwise.", fig.height=8}
-ls_use_case$plot_before_aggregation_forecast + ls_use_case$plot_aggregation + plot_annotation(tag_levels = 'A')
-
+```{r ucHyperparamMultipleIS, echo=FALSE, fig.cap="Hyperparameter evaluation on training set by random search of the model with multiple input scaling and no connection between input layer and output layer.
+Hp sets with MAE above 30 were removed for clarity of visualisation.", fig.height=8}
+ls_use_case$plot_feature_input_scaling
 ```
 
-### Present forecast performance
+### Forecast performance
 
-```{r perfMonthlyUpdate, echo = FALSE}
+```{r performance, echo = FALSE}
 ls_use_case$table_perf_esn %>%
@@
 For each setting, 40 reservoir are computed and the forecast is the median of the 40 forecast.
-Results show the mean number of features and the performance metrics.
-MAE = Mean Absolute Error, MRE = Median Relative Error, MAEB = Mean Absolute Error to Baseline, MREB = Median Relative Error to Baseline.",
+Results show the performance metrics : MAE = Mean Absolute Error, MRE = Median Relative Error, MAEB = Mean Absolute Error to Baseline, MREB = Median Relative Error to Baseline.",
                col.names = c("Model", "MAE", "MRE", "MAEB", "MREB"),
@@
 
-Table \@ref(tab:perfMonthlyUpdate) shows the performance on the test set.
-The number of selected features by elastic-net and by genetic algorithm were in the same order of magnitude.
-Updating the hyperparameter monthly did not always improve forecast accuracy (e.g epidemiological subset with multiple input scaling).
-Best model depend on the metric but epidemiological subset with multiple input scaling and no update was the best model or second to best model for all metrics.
+Table \@ref(tab:performance) shows the performance on the test set.
+Best model according to all metrics was RC with common input scaling and no connection between input and output layers.
+Having one input scaling per feature did not improve the model which might be due to low generalisability of the hyperparameter of the training set to the test set due to non-stationarity.
+Additionaly, connecting input layer to output layer did not improve the model, possibly because it add more dimension to the problem while the information provided is already inside the reservoir.
+We observe that elastic-net did not perform well and stay close to the baseline prediction.
+However, when the penalisation is updated each day, the peformance improve.
+This is emphasize the problem of the transferrability of the hyperparameters to which multiple input scaling models seem more sensitive as they have more hyperparameters.
+For instance, vaccine will keep a low weight whereas its low initial importance quickly increase to mitigate the effect of the high RT-PCR increase on the hospitalisation dynamic.
+
+Figure \@ref(fig:forecastwithorwithoutUpdate) shows the forecast of the different models.
+We note that models struggle to accurately forecast slope shifts.
+For instance, summer 2021 initial increase is missed by all models, summer decrease is partially anticipated by RC with common IS and no input layer connection, winter 2021 increase is anticipated by all models but they tend to overestimate it because of the rise of vaccine effect.
 
@@
 
-Figure \@ref(fig:forecastwithorwithoutUpdate) shows the forecast with and without monthly hyperparameter update.
-Graphically we observe that the forecast of individual models (in grey) are chaotic and that forecast aggregation (in red) is necessary.
-This is less concerning for epidemiological subset with multiple input scaling or when the seed is controlled.
-In terms of performance, we observe similar conclusion as the one depicted for table \@ref(tab:perfMonthlyUpdate).
-However, the model errors did not occur for the same period for each model.
-For instance, the epidemiological with multiple scaling and monthly update error is located mainly during the winter 2022 increase.
-Model based on all features tend to overestimate the length of the summer 2021 wave.
+### Number of model to aggregate
+
+Figures \@ref(fig:ucAggregationSpread) and figure \@ref(fig:ucAggregationNb) show 40 model forecast for each of the best hyperparameter set of the different reservoir architecture (i.e the same set of hyperparameters as used for test set forecast).
+It means at each day, the same set of hyperparameter is kept but a new reservoir is generated.
+Due to the internal random connection of the reservoir, it might result in forecast stochasticity.
+We observe that models with ridge penalty higher than 1 (i.e all models except multiple IS with input connection) tend to have more stable forecast but that the addition of input layer connection provided some extreme forecasted value.
+This extreme value phenomena is relatively rare because it is corrected when the number of model for aggregation is above 15.
+
+```{r ucAggregationSpread, echo = FALSE, fig.cap = "Individual forecast of 40 repetition of the best hyperparameter set for the different RC configuration.
+Forecast value above 200 were set to 200.", fig.height=6}
+ls_use_case$plot_before_aggregation_forecast
+```
+
+```{r ucAggregationNb, echo = FALSE, fig.cap = "It shows the impact of aggregating several RC with the best set of HP for each RC configuration.
+It has small interest when the input layer was not connected to output layer but significantly improves the performance otherwise.", fig.height=6}
+ls_use_case$plot_aggregation
+```
 
@@
 
+In this specific application, we have demonstrated that RC exhibits commendable performance in comparison to Elastic-net, which serves as the reference model.
+Furthermore, we highlight the inherent challenges in forecasting within this context, primarily stemming from the non-stationarity of the time series.
+
+This study represents a novel contribution to epidemic forecasting utilizing reservoir computing.
+Notably, previous literature predominantly focused on simpler problems characterized by fewer input features or shorter evaluation periods.
+Our findings underscore the potential of this approach for future epidemics, suggesting its potential to surpass more traditional epidemiological tools while maintaining a lightweight model structure compared to other recurrent neural networks (RNNs).
+
+All computations in this study were conducted using the reservoirnet package, and the entire codebase is accessible on GitHub (**todo : open to the public**).
+The package demonstrates its efficacy in implementing various reservoir architectures, including connection between the input layer and the output layer, as well as the utilization of several input scaling, all within the context of a real-world use case.
+
+Given the substantial number of hyperparameters involved, we acknowledge that random search may not be the most efficient optimization algorithm.
+We have retained this approach for the sake of simplicity in this tutorial paper; however, meta-heuristic approaches, particularly those utilizing evolutionary algorithms, may prove more efficient, especially when employing multiple input scaling.
+
+It is worth noting that all models, including the @ferte_benefit_2022 models, encounter challenges in accurately anticipating slope shifts, indicating the need for further investigation.
+Specifically, additional work is warranted to extend the application of RC to high-dimensional settings, building upon the insights gained from models based on a more extensive set of features.
+
 # Discussion and conclusion
 
+In this paper, we introduce the R package `reservoirnet`, which serves as a versatile tool for implementing reservoir computing based on `ReservoirPy` Python API.
+It offers flexibility in defining the reservoir architecture, including options for specifying connections between the input layer and the output layer, as well as variations in input scaling as demonstrated on a real-world use case.
+
+We provided a comprehensive overview of the basic usage of the `reservoirnet` package through illustrative examples in regression and classification tasks.
+This introductory section serves as a foundation for R users, offering step-by-step guidance on constructing and training reservoir computing models using the package.
+By demonstrating the application of RC in both regression and classification scenarios, we aim to equip users with the essential knowledge and skills needed to harness the capabilities of reservoir computing for diverse tasks.
+
+Drawing on the robust foundation of the `ReservoirPy` structure, a well-maintained library, this package inherits its reliability and longevity.
+Nevertheless, our emphasis has been on refining the primary functionalities, which may leave more advanced users desiring deeper access to core functions, such as the generation of the reservoir random connections.
+This initial version of `reservoirnet` must evolve in tandem with the growing understanding and adoption of reservoir computing within the R community.
+
 # Bibliography

### 2023-11-27 script/forecast/forecast_test_eval.R

@@ -2,7 +2,7 @@
 library(dplyr)
 library(ggplot2)
 
-source("functions/fct_smoothing_derivative.R")
+source(here::here("functions/fct_smoothing_derivative.R"))
 
 ##### load data
 ls_files_results <- list.files("data/results_forecast_testset_11495493/", full.names = TRUE)
@@ -36,38 +36,38 @@ hp_features <- c("hosp",
                  "Vaccin_1dose_rolDeriv7")
 
 feature_labels <- conservation_status <- c(
-  "outcome" = "Outcome",
-  "outcomeDeriv" = "Outcome variation",
+  "outcome" = "Hospitalisations t+14",
+  "outcomeDeriv" = "Outcome",
   "input_scaling" = "Input scaling",
   "leaking_rate" = "Leaking rate",
   "ridge" = "Ridge",
   "spectral_radius" = "Spectral radius",
   "hosp" = "Hospitalizations",
-  "P_TOUS_AGES" = "Positive RT-PCR",
-  "P_60_90_PLUS_ANS" = "Positive RT-PCR 60 yo+",
-  "FRACP_TOUS_AGES" = "% Positive RT-PCR",
-  "FRACP_60_90_PLUS_ANS" = "% Positive RT-PCR 60 yo+",
-  "URG_covid_19_COUNT" = "Emergency sojourn with covid-19",
+  "P_TOUS_AGES" = "RT-PCR+",
+  "P_60_90_PLUS_ANS" = "RT-PCR+ 60 yo+",
+  "FRACP_TOUS_AGES" = "% RT-PCR+",
+  "FRACP_60_90_PLUS_ANS" = "% RT-PCR+ 60 yo+",
+  "URG_covid_19_COUNT" = "Emergency",
   "IPTCC.mean" = "IPTCC",
   "Vaccin_1dose" = "Vaccine",
-  "hosp_rolDeriv7" = "Hospitalizations (1st deriv)",
-  "P_TOUS_AGES_rolDeriv7" = "Positive RT-PCR (1st deriv)",
-  "P_60_90_PLUS_ANS_rolDeriv7" = "Positive RT-PCR 60 yo+ (1st deriv)",
-  "FRACP_TOUS_AGES_rolDeriv7" = "% Positive RT-PCR (1st deriv)",
[…]

### 2023-11-29 draft_jss/draft_jss.Rmd
add dividing by maximum

@@
 
-Reservoir Computing (RC) is a prominent machine learning methodology proposed by 
+Reservoir Computing (RC) is a prominent machine learning methodology, proposed by @kim_health_2020 and @maass_real-time_2002,
 that has gained significant attention in recent years for its ability to
-effectively process information generated by dynamical systems [@kim_health_2020, @maass_real-time_2002]. This
+effectively process information generated by dynamical systems. This
 innovative approach leverages the dynamics of a high-dimensional
@@
 Secondly, RC relies on numerous hyperparameters to establish connections between neurons.
-Despite this, guidance on determining these hyperparameters is limited, often requiring users to adopt a wrapper approach where combinations of hyperparameter performance are assessed on a training set before selecting the optimal combination.
+Guidance on determining these hyperparameters is limited, often requiring users to adopt a wrapper approach where combinations of hyperparameter performance are assessed on a training set before selecting the optimal combination.
 Thirdly, there is a current absence of implementation in R, rendering the method challenging for users unfamiliar with Python [@trouvain_reservoirpy_2020, @steiner_pyrcn_2022] or Julia [@martinuzzi_reservoircomputingjl_2022].

### 2023-11-29 functions/fct_aggregate_forecast.R

@@ -10,10 +10,10 @@ fct_aggregate_forecast <- function(data_i,
                                    forecast_deriv){
   data_i %>%
     dplyr::mutate(forecast_deriv = forecast_deriv,
-                  forecast = forecast_deriv + hosp,
+                  forecast = forecast_deriv + outcomeHosp,
                   forecast = if_else(forecast < 0, 0, forecast)) %>%
     dplyr::slice_max(outcomeDate) %>%
-    dplyr::select(START_DATE, outcomeDate, outcome, forecast, hosp) %>%
+    dplyr::select(START_DATE, outcomeDate, outcome, forecast, hosp = outcomeHosp) %>%
     return()
 }
 

### 2023-11-29 functions/fct_smoothing_derivative.R

@@ -44,7 +44,17 @@ fct_smoothing_derivative <- function(data,
     # remove missing values due to derivative computation
     na.omit() %>%
     # get the outcome derivative
-    mutate(outcomeDeriv = outcome - hosp) %>%
+    mutate(outcomeDeriv = outcome - hosp,
+           outcomeHosp = hosp) %>%
+    mutate(across(!START_DATE & !outcome & !outcomeDate & !outcomeDeriv & !outcomeHosp,
+                  function(x){
+                    maxX <- max(abs(x))
+                    if(maxX == 0){
+                      return(x)
+                    } else {
+                      return(x/maxX)
+                    }
+                  })) %>%
     return()
 }
 

### 2023-11-29 script/forecast/forecast_test.R
new forecast with scaling

@@ -14,10 +14,10 @@ warmup = 30
 units = 500
 nb_iter = 40
 ##### LOAD BEST HP SETS
-hp_sets = list(common_input_scaling = "data/common_input_scaling_11423760/",
-               common_input_scaling_linked_source = "data/common_input_scaling_11426140/",
-               multiple_input_scaling = "data/multiple_input_scaling_11423760/",
-               multiple_input_scaling_linked_source = "data/multiple_input_scaling_11426140/",
+hp_sets = list(common_input_scaling = "data/common_input_scaling_11559350/",
+               common_input_scaling_linked_source = "data/common_input_scaling_11559390/",
+               multiple_input_scaling = "data/multiple_input_scaling_11559350/",
+               multiple_input_scaling_linked_source = "data/multiple_input_scaling_11559390/",
                enet = "data/enet_1/") %>%
   lapply(function(path_i){
     list.files(path_i, full.names = TRUE) %>%

### 2023-11-29 script/forecast/forecast_test.R
divide to conquer hp

@@ -12,7 +12,8 @@ data_covid <- readRDS(file = "data/df_obfuscated_epidemio.rds")
 forecast_days = 14
 warmup = 30
 units = 500
-nb_iter = 40
+nb_iter = 1
+nb_sample = 40
 ##### LOAD BEST HP SETS
 hp_sets = list(common_input_scaling = "data/common_input_scaling_11559350/",
                common_input_scaling_linked_source = "data/common_input_scaling_11559390/",
@@ -26,8 +27,15 @@ hp_sets = list(common_input_scaling = "data/common_input_scaling_11559350/",
       tibble::rowid_to_column(var = "hp_set")
   })
 
-best_hp_set = lapply(hp_sets,
-                     function(x) x %>% slice_min(mean_absolute_error, n = 1))
+best_hp_set = lapply(names(hp_sets),
+                     function(x){
+                       if(x == "enet"){
+                         hp_sets[[x]] %>% slice_min(mean_absolute_error, n = 1)
+                       } else {
+                         hp_sets[[x]] %>% slice_min(mean_absolute_error, n = nb_sample)
+                       }
+                     })
+names(best_hp_set) <- names(hp_sets)
 ##### DEFINE EVALUATION SET
 vecDates <- data_covid %>%
   filter(START_DATE >= as.Date("2021-03-01") + forecast_days) %>%
@@ -38,63 +46,74 @@ message(paste0("--------- evaluation date : ", date_i, " ----------"))
 
 ##### FORECAST
 dfres <- lapply(names(best_hp_set),
-       function(model_set){
-         print(model_set)
-         hp_set <- best_hp_set[[model_set]]
-         if(model_set == "enet"){
-           ##### ELASTIC-NET
-           forecast_enet_once <- fct_iterative_forecast_from_hp(data_covid = data_covid,
-                                                                vecDates = date_i,
-                                                                forecast_days = forecast_days,
-                                                                lambda = best_hp_set$enet$lambda,
-                                                                alpha = best_hp_set$enet$alpha,
-                                                                model = "enet") %>%
-             mutate(model = "enet_hp_on_train_set",
-                    hp_set = hp_set$hp_set)
-           forecast_enet_daily <- fct_iterative_forecast_from_hp(data_covid = data_covid,
-                                                                 vecDates = date_i,
-                                                                 forecast_days = forecast_days,
-                                                                 model = "enet") %>%
-             mutate(model = "enet_daily_hp_update",
-                    hp_set = 1)
[…]

### 2023-11-29 script/forecast/forecast_test_eval.R
update forecast eval

@@ -5,7 +5,7 @@ library(ggplot2)
 source(here::here("functions/fct_smoothing_derivative.R"))
 
 ##### load data
-ls_files_results <- list.files("data/results_forecast_testset_11495493/", full.names = TRUE)
+ls_files_results <- list.files("data/results_forecast_testset_11561202/", full.names = TRUE)
 ls_files_results_res <- grep(ls_files_results, pattern = "hp_sets", value = TRUE, invert = TRUE)
 hp_file <- grep(ls_files_results, pattern = "hp_sets", value = TRUE)
 df_obfuscated_epidemio <- readRDS("data/df_obfuscated_epidemio.rds")
@@ -76,7 +76,7 @@ data_i <- fct_smoothing_derivative(data = df_obfuscated_epidemio,
                                    forecast_days = 14)
 
 plot_present_data <- data_i %>%
-  select(-START_DATE, -ends_with("rolDeriv7")) %>%
+  select(-START_DATE, -outcomeHosp, -ends_with("rolDeriv7")) %>%
   tidyr::pivot_longer(cols = -"outcomeDate") %>%
   mutate(name = factor(name, levels = names(feature_labels), labels = feature_labels)) %>%
   ggplot(mapping = aes(x = outcomeDate, y = value, color = name)) +
@@ -195,6 +195,9 @@ df_forecast_aggregated <- df_forecast %>%
                                labels = model_labels))
 
 table_perf_esn <- df_forecast_aggregated %>%
+  dplyr::mutate(forecast = if_else(forecast < 10, 10, forecast),
+                hosp = if_else(hosp < 10, 10, hosp),
+                outcome = if_else(outcome < 10, 10, outcome)) %>%
   dplyr::mutate(absolute_error = abs(forecast - outcome),
                 relative_error = absolute_error/outcome,
                 baseline_absolute_error = abs(hosp - outcome),
@@ -229,7 +232,7 @@ plot_forecast <- df_forecast_aggregated %>%
 
 ##### aggregation forecast plots
 nb_aggregation <- c(1, 5, 10, 15, 20, 30, 40)
-nb_boot <- 10
+nb_boot <- 250
 
 df_aggregation_forecast <- parallel::mclapply(seq_len(nb_boot),
                                               mc.cores = parallel::detectCores()-2, 
@@ -238,12 +241,12 @@ df_aggregation_forecast <- parallel::mclapply(seq_len(nb_boot),
                                                        function(nb_aggregation_i){
                                                          df_forecast %>%
[…]

### 2023-11-29 draft_jss/draft_jss.Rmd
update manuscript

@@
   facet_grid(name ~ ., scales = "free_y") +
-  theme_bw() +
+  theme_minimal() +
   geom_vline(mapping = aes(color = "train-test sets",
@@
 
-```{r nodeactivationbad, fig.cap="Node activation over time."}
+```{r nodeactivationbad, fig.cap="20 random nodes activation over time."}
 plot(reservoir_state)
@@
 This issue can be attributed to the disparate scales of the features.
-To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby scaling them within the range of `-1` to `1`.
+To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby scaling them within the range of `0` to `1` by dividing by the maximum of the absolute value.
 
 ```{r}
-stand_max <- function(x) return(x/max(x))
+stand_max <- function(x) return(x/max(abs(x)))
 # scaled features
@@
 
-```{r nodeactivationgood, fig.cap="Node acivation over time. Scaled features"}
+```{r nodeactivationgood, fig.cap="20 random node acivation over time. Scaled features."}
 # feed them to the reservoir
@@
   scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
-  theme_bw() +
+  theme_minimal() +
   labs(color = "", x = "Date", y = "Hospitalizations")
@@
 For that step, we will learn on the `outcome_deriv` contained in `yTrain_variation` data which is defined outcome as `outcome_deriv = outcome - hosp`.
+As depicted at figure \@ref(fig:figForecastDerivvsRaw), this strategy improved the model forecast.
 
-```{r fig.cap="Covid-19 hospitalizations forecast.
+```{r figForecastDerivvsRaw, fig.cap="Covid-19 hospitalizations forecast.
 The model is either trained to forecast the number of hospitalizations (denoted Raw) or the variation of the hospitalizations compared to current level of hospitalisation (denoted Variation)"}
@@
 ```
-
-We observe an improvement of the forecast compared to the forecast of hospitalization previously discussed.
-From there, many improvement can be implemented including : frequent update of the model, leverage additional features and hyperparameter optimization.
-This notions will be discussed in the [Study case section](#studycase).
 
@@
              ncol = 2) +
+  scale_fill_gradient2(low = "#8ECAE6", high = "#FB8500", mid = "#023047",
+                       midpoint = 0) +
   theme_minimal() +
@@
 To take into account measurement error and daily noise variation, data were smoothed using a local polynomial regression with a span of 21 days.
+As previously described, input features were scaled between -1 and 1 by dividing the observed value by the maximum of the absolute value of the given input feature.
 
@@
 
+Because the outcome is obfuscated below 10 hospitalizations for privacy reason, we set both the outcome and the forecast to 10 when the observed value was 0 or the forecasted value was below 10 when evaluating the model performance.
+
 ### Models
@@
 
-Because of the randomness of the reservoir, we took the median forecast of 20 reservoir on the train set and 40 reservoir on the test set.
+Because of the randomness of the reservoir, we took the median forecast of 10 reservoir on the train set to evaluate the performance of a given hyperparameter set.
+On the test set we aggregated the forecast of 40 reservoir, each of them having one of the 40 best hyperparameter sets found on the train set.
 In addition, because covid-19 hospitalization is a non-stationary process, models were re-trained everyday using all previous days.
@@
 Model is trained to forecast Outcome curve which corresonds to the difference between Hospitalisatiosn at 14 days and current hospitalisations.
-Other features are represented in darkblue.", fig.height=8}
+Other features are scaled (divide by the maximum of the feature) represented in darkblue.", fig.height=8}
 ls_use_case$plot_present_data
@@
 Figure \@ref(fig:ucHyperparam) shows the hyperparameter optimisation using random search for the different RC architectures.
-We note that the multiple input scaling setting with input layer connected to output layer showed poor performance on the training set which might be due to an unextensive search of hyperparameter regarding the better performance of the common input scaling with connected input layer which is a specific configuration where all input scaling are equal.
+We observe that model with multiple input scaling achieved better performance on the train set compared to model with single input scaling which is expected as they can adapt more closely to the data thanks to specific input scaling for each feature.
 
-As expected, we observe that the optimal leaking rate is close to 1 for all RC which is coherent with the short term dynamic of covid-19 epidemic.
-Trend for other hyperparameter are less clear even though best hyperparameter sets were close for RC with common input scaling.
+As expected, we observe that the optimal leaking rate is above 1e-2 for all RC which is coherent with the short term dynamic of covid-19 epidemic.
+Trend for other hyperparameter are less clear even though best hyperparameter sets were close for RC with common input scaling and for RC with multiple input scaling.
 
@@
 Figure \@ref(fig:ucHyperparamMultipleIS) shows the hyperparameter search for RC with multiple input scaling and connected input layer.
-We observe that the random search tend to favor low importance given to vaccine and high importance given to derivative of hospitalisations and derivative of positive RT-PCR in the elderly.
+We observe that the random search tend to favor high importance given to derivative of positive RT-PCR (including the elderly) and the derivative of IPTCC.
 The rest of the feature do not show clear pattern.
-The lack of importance of number of people with one dose of vaccine in this setting was discussed by @ferte_benefit_2022 and related to the lack of vaccinated people in the training set as depicted at figure \@ref(fig:ucPresentData).
 
@@
 Table \@ref(tab:performance) shows the performance on the test set.
-Best model according to all metrics was RC with common input scaling and no connection between input and output layers.
+Best model according to all metrics was RC with common input scaling and connection between input and output layers.
 Having one input scaling per feature did not improve the model which might be due to low generalisability of the hyperparameter of the training set to the test set due to non-stationarity.
-Additionaly, connecting input layer to output layer did not improve the model, possibly because it add more dimension to the problem while the information provided is already inside the reservoir.
-We observe that elastic-net did not perform well and stay close to the baseline prediction.
+Additionaly, connecting input layer to output layer improved the model forecast.
+
+We observe that elastic-net did not perform well and stayed close to the baseline prediction.
 However, when the penalisation is updated each day, the peformance improve.
 This is emphasize the problem of the transferrability of the hyperparameters to which multiple input scaling models seem more sensitive as they have more hyperparameters.
-For instance, vaccine will keep a low weight whereas its low initial importance quickly increase to mitigate the effect of the high RT-PCR increase on the hospitalisation dynamic.
 
@@
 We note that models struggle to accurately forecast slope shifts.
-For instance, summer 2021 initial increase is missed by all models, summer decrease is partially anticipated by RC with common IS and no input layer connection, winter 2021 increase is anticipated by all models but they tend to overestimate it because of the rise of vaccine effect.
+For instance, summer 2021 initial increase is partially predicted by all models but its decrease is not well predicted.
+Winter 2021 increase is anticipated by all models but they tend to overestimate it because of the rise of vaccine effect.
 
@@
 
-Figures \@ref(fig:ucAggregationSpread) and figure \@ref(fig:ucAggregationNb) show 40 model forecast for each of the best hyperparameter set of the different reservoir architecture (i.e the same set of hyperparameters as used for test set forecast).
-It means at each day, the same set of hyperparameter is kept but a new reservoir is generated.
-Due to the internal random connection of the reservoir, it might result in forecast stochasticity.
-We observe that models with ridge penalty higher than 1 (i.e all models except multiple IS with input connection) tend to have more stable forecast but that the addition of input layer connection provided some extreme forecasted value.
-This extreme value phenomena is relatively rare because it is corrected when the number of model for aggregation is above 15.
+Figures \@ref(fig:ucAggregationSpread) show the individual forecast for the 40 best set of hyperparameters of each RC architecture.
+Due to the internal random connection of the reservoir, we observe forecast stochasticity and relying on only one forecast is unreliable.
+We explored the number of model needed at figure \@ref(fig:ucAggregationNb) which shows that after 10 models, forecast is stable and even 5 models for the simpler model with common input scaling which rely on less hyperparamters.
 
-```{r ucAggregationSpread, echo = FALSE, fig.cap = "Individual forecast of 40 repetition of the best hyperparameter set for the different RC configuration.
-Forecast value above 200 were set to 200.", fig.height=6}
+```{r ucAggregationSpread, echo = FALSE, fig.cap = "Individual forecast the 40 best hyperparameter sets for the different RC configuration.
+Forecast value above 200 were set to 200 for clarity.", fig.height=6}
 ls_use_case$plot_before_aggregation_forecast
@@
 
-```{r ucAggregationNb, echo = FALSE, fig.cap = "It shows the impact of aggregating several RC with the best set of HP for each RC configuration.
-It has small interest when the input layer was not connected to output layer but significantly improves the performance otherwise.", fig.height=6}
+```{r ucAggregationNb, echo = FALSE, fig.cap = "Mean absolute error depending on the number of aggregated reservoir.", fig.height=6}
 ls_use_case$plot_aggregation
@@
 
-This study represents a novel contribution to epidemic forecasting utilizing reservoir computing.
-Notably, previous literature predominantly focused on simpler problems characterized by fewer input features or shorter evaluation periods.
-Our findings underscore the potential of this approach for future epidemics, suggesting its potential to surpass more traditional epidemiological tools while maintaining a lightweight model structure compared to other recurrent neural networks (RNNs).
-
 All computations in this study were conducted using the reservoirnet package, and the entire codebase is accessible on GitHub (**todo : open to the public**).
@@
 Given the substantial number of hyperparameters involved, we acknowledge that random search may not be the most efficient optimization algorithm.
-We have retained this approach for the sake of simplicity in this tutorial paper; however, meta-heuristic approaches, particularly those utilizing evolutionary algorithms, may prove more efficient, especially when employing multiple input scaling.
+We have retained this approach for the sake of simplicity in this tutorial paper; however, meta-heuristic approaches, particularly those utilizing evolutionary algorithms, may prove more efficient, especially when employing multiple input scaling [@bala_applications_2018].
+
+This study represents a novel contribution to epidemic forecasting utilizing reservoir computing.
+Notably, previous literature predominantly focused on simpler problems characterized by fewer input features or shorter evaluation periods [@ghosh_reservoir_2021, @liu_nanophotonic_2023, @ray_optimized_2021].
+Our findings underscore the potential of this approach for future epidemics, suggesting its potential to surpass more traditional epidemiological tools while maintaining a lightweight model structure compared to other recurrent neural networks (RNNs).
 

### 2023-11-29 script/forecast/forecast_test_eval.R

@@ -99,9 +99,9 @@ df_hp <- readRDS(hp_file) %>%
                 n_try = n(),
                 rank_perf_color = case_when(rank_perf == 1 ~ "Best",
                                             rank_perf <= 10 ~ "Top 10",
-                                            rank_perf <=  100 ~ "Top 100",
-                                            rank_perf > 100 ~ "Others"),
-                rank_perf_color = factor(rank_perf_color, levels = c("Best", "Top 10", "Top 100", "Others")),
+                                            rank_perf <=  40 ~ "Top 40",
+                                            rank_perf > 40 ~ "Others"),
+                rank_perf_color = factor(rank_perf_color, levels = c("Best", "Top 10", "Top 40", "Others")),
                 mean_absolute_error = if_else(mean_absolute_error >= 30, 30, mean_absolute_error)) %>%
   ungroup() %>%
   dplyr::select(all_of(c("mean_absolute_error",
@@ -160,7 +160,7 @@ plot_reservoir_hp <- df_hp %>%
 
 plot_feature_input_scaling <- df_hp %>%
   filter(HP %in% hp_features,
-         model == "multiple_input_scaling") %>%
+         model == "multiple_input_scaling_linked_source") %>%
   ggplot(mapping = aes(x = HP_value, 
                        y = mean_absolute_error,
                        color = rank_perf_color)) +

### 2023-12-18 draft_jss/draft_jss.Rmd
update draft

@@
 
-Reservoir Computing (RC) is a prominent machine learning methodology, proposed by @kim_health_2020 and @maass_real-time_2002,
+Reservoir Computing (RC) is a prominent machine learning method, proposed by @jaeger__2001, @maass_real-time_2002 and @lukosevicius_reservoir_2009
 that has gained significant attention in recent years for its ability to
-effectively process information generated by dynamical systems. This
+efficiently process information generated by dynamical systems. This
 innovative approach leverages the dynamics of a high-dimensional
-reservoir to perform complex computations and solve various tasks based
+"reservoir" (we define it below) to perform complex computations and solve various tasks based
 on the response of this dynamical system to input signals.
-RC has demonstrated its efficacy in tackling various challenges, encompassing pattern classification and time series forecasting in applications ranging from electrocardiogram analysis to bird calls, power plants, internet traffic, stock prices, and beyond [@tanaka_recent_2019].
+RC has demonstrated its efficacy in tackling various challenges, encompassing pattern classification and time series forecasting in applications ranging from electrocardiogram analysis to bird calls @trouvain_canary_2021, language processing @hinaut_real-time_2013, power plants, internet traffic, stock prices, and beyond [@tanaka_recent_2019, @lukosevicius_reservoir_2009].
 
-RC is a kind of recurrent neural networks (RNNs) where the internal states, denoted as the reservoir, is randomly generated and solely the final layer is trained.
+Originally, the RC paradigm was implemented in artificial firing-rate neurons  (“Echo State Networks”, @jaeger__2001) and spiking neurons ("Liquid State Machine”, @maass_real-time_2002) as a recurrent neural network (RNN) where the internal recurrent connections, denoted as the reservoir, are randomly generated and only the output layer (named “read-out”) is trained.
 The reservoir projects temporal input signals onto a high-dimensional feature space facilitating the learning of non-linear and temporal interactions.
-This offers the advantage of decreasing the computing time while consistently maintaining robust performance compared to conventional RNNs [@shahi_prediction_2022, @vlachas_backpropagation_2020].
+Thus, this recurrent layer contains are high-dimensional non-linear recombination of the inputs and past states: it is a “reservoir of computations” from which useful information can be linearly extracted (or “read-out”) to provide the desired outputs.
+This offers the advantage of decreasing the computing time while consistently maintaining robust performance compared to conventional RNNs [@vlachas_backpropagation_2020].
+In parallel, this RC paradigm gained an increasing interest because of its ability to be implemented not only on classical computers.
+Indeed, the hidden recurrent layer can be kept untrained and thus a wide range of physical media can be used to replace it.
+Recently, @tanaka_recent_2019 reviewed this prolific field: from FPGA hardware [@penkovsky_efficient_2018], to spin waves using magnetic properties [@nakane_reservoir_2018], skrymions [@prychynenko_magnetic_2018] or optical implementations [@rafayelyan_large-scale_2020].
+This provides interesting and potentially more efficient alternative to traditional machine learning.
+
+RC leverages various hyperparameters to introduce prior knowledge about the relationship between input variables and output targets.
+However, since the connections within the reservoir are randomly initialized, the same set of hyperparameters may exhibit diverse behaviors across different realizations of the reservoir connections.
+This unpredictability makes it challenging to anticipate the performance of a particular hyperparameter setting, as identical settings may produce varying outcomes when applied to distinct instances of the reservoir.
+Moreover, selecting the most suitable hyperparameters often requires researchers to experiment with multiple combinations on a training dataset and evaluate their performance on a separate test set.
+Although this approach can be resource-intensive and time-consuming, it is a compromise that is acceptable considering the rapid simulation capabilities offered by RC.
+
+*XH : Commentaires généraux sur le paragraphe : il faut ne faut pas insister sur les hyperparamètres comme si c’était négatif.
+Plutôt dire qu’il y a des optimisations et tricks à connaitre, et que c’est intéressant d’avoir une librairie pour ça, et aussi à des fins de reproductibilité.
+Dire que c’est un trade-off car c’est les a priori nécessaire qu’on rajoute dans l’algo car on gagne du temps sur l’apprentissage.
+C’est ambigu de parler de stochaticité ici je trouve, car le processus n’est pas stochastic une fois les poids fixés.*
+
+*TF : j'ai modifié, mieux ?*
 
@@
 Secondly, RC relies on numerous hyperparameters to establish connections between neurons.
-Guidance on determining these hyperparameters is limited, often requiring users to adopt a wrapper approach where combinations of hyperparameter performance are assessed on a training set before selecting the optimal combination.
-Thirdly, there is a current absence of implementation in R, rendering the method challenging for users unfamiliar with Python [@trouvain_reservoirpy_2020, @steiner_pyrcn_2022] or Julia [@martinuzzi_reservoircomputingjl_2022].
+Guidance on determining these hyperparameters is limited, often requiring users to evaluate the performance of different combinations of hyperparameters on a train set before selecting the optimal combination to forecast on the test set^[In this article, we employ the term "train set" to refer to the combined dataset consisting of both the training and validation sets, which are cycled through in a cross-validation manner.].
+Thirdly, there is a current absence of implementation in R, rendering the method challenging for users unfamiliar with Python [@trouvain_reservoirpy_2022] or Julia [@martinuzzi_reservoircomputingjl_2022].
+
+*XH : Je pense que ça embrouille là : c’est train/test/validation Les hyperparamètres sont entrainé sur le train set puis évalués sur le validation set.
+Et ensuite le test set.
+Par abus de langage on dit souvent « test set » au lieu de « validation set » quand on optimise les paramètres.
+Moi ça me choque pas, mais dans un papier vaut mieux utiliser les termes adéquats, ou en tous cas toujours utiliser les memes, donc si tu n’as jamais de validation set (enfin le vrai test set), ça peut passer (pas besoin de tout changer).
+Met juste une note de bas de page pour préciser que en général on devrait l’appeler le validation set.
+On verra ce que les reviewers disent, pas besoin de perdre du temps maintenant.*
+
+*TF : ok, j'ai rajouté une note de bas de page ci-dessous*
 
 In this paper, we endeavor to delve into these challenges, offering comprehensive guidance to assist new users in maximizing the benefits of RC.
-Initially, a broad introduction to reservoir computing is presented, accompanied by a tutorial on its application using `reservoirnet`, an R package built upon the `reservoirPy` Python module [@trouvain_reservoirpy_2020].
+Initially, a broad introduction to reservoir computing is presented, accompanied by a tutorial on its application using `reservoirnet`, an R package built upon the `reservoirPy` Python module [@trouvain_reservoirpy_2020, @trouvain_create_2022, @trouvain_reservoirpy_2022].
 Subsequently, we investigate the various challenges associated with the use of RC for forecasting COVID-19 hospitalizations.
@@
 
-RC is a machine learning algorithm composed of an input layer denoted
-$u(t)$ which is randomly transformed by the reservoir into an activation
-state denoted $x(t)$ which is fed to a ridge penalized linear regression
-trained to forecast the outcome denoted as $y(t)$ as depicted at figure
-\@ref(fig:rcpresentation).
+RC is a machine learning paradigm which is most often implemented as Echo State Networks (ESNs; @jaeger__2001), the firing-rate neuron version.
+An ESN is described by three matrices of connectivity: an input layer $W_{in}$, a recurrent layer $W$ and an output layer $W_{out}$.
+At each time step, the input vector $u_t$ is projected into the reservoir which is also combined with reservoir past state $x(t-1)$ through the recurrent connections.
+The output $y(t)$ is linearly read-out from the reservoir.
+Input $W_{in}$ and recurrent $W$ matrices are kept random; only the output matrix $W_{out}$ is trained in an offline or online method.
+Often a ridge regression (i.e.
+a regularized linear regression) is used to obtain the desired outputs $y(t)$ from the reservoir states $x(t)$.
+Figure \@ref(fig:rcpresentation) depicts the architecture.
+For simplicity, we will use the term "reservoir computing” for “Echo State Network” in the remainder of the paper.
 
@@
 Connection between input layer and reservoir and inside reservoir are random.
-Only the output layer is optimized based on a ridge penalized linear regression."}
+Only the output layer is optimized based on a ridge regression."}
 knitr::include_graphics("images/schema.png/image1.png")
@@
 
-The leaking rate alpha define the rate of update of the nodes. The
+The leaking rate $\alpha$ defines the rate of update of the nodes. The
 closer $\alpha$ is to $1$, the higher the reservoir is sensitive to new
@@
 (i.e $x(t)$) and the new inputs (i.e $u(t+1)$). Both $W_{in}$ and $W$
-are random matrices of size $Nres \times M$ and $Nres \times Nres$
+are random matrices of size $N_{res} \times M$ and $N_{res} \times N_{res}$
 respectively.
 
-$W_{in}$ is a dense matrix generated using a Bernoulli distribution
-where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an
-equal probability where $m = 1, …, M$ corresponds to a given feature in
-the input layer. The input scaling, denoted $I_{scale}$, is an
-hyperparameter coefficient which can be common to all features from the
-input layer or specific to each feature $m$. In that case, the more
-important the feature is, the greater should be its input scaling. $W$
-is a sparse matrix where values are generated from a Gaussian
-distribution $\mathcal{N}(0,1)$. Then the matrix $W$ is scaled
-according to the spectral radius, an hyperparameter defining the highest
-eigen value of $W$.
+The leaking rate $\alpha$ defines the update rate of the nodes.
+The closer $\alpha$ is to $1$, the more the reservoir is sensitive to new inputs $u(t)$.
+Therefore, the reservoir state at time $t+1$ denoted $x(t+1)$ depends on the reservoir state at the previous time $x(t)$ and the new inputs $u(t+1)$.
+Both $W_{in}$ and $W$ are random matrices of size $N_{res} \times M$ and $N_{res} \times N_{res}$ respectively.
+
+*XH : Ah tu utilises une matrice dense ?
+si oui, t’as fait ça dans toutes tes simulations ?
+habituellement on utilise plutôt 10 ou 20% de connectivité, surtout si on prend un distribution bimodale comme bernouilli.
+En tous cas il faut pas préciser que soit doit être dense ici.*
+
+*TF : on a pris les paramètres par défaut de reservoirpy, si j'ai bien compris, la matrice est générée par la fonction reservoirpy.mat_gen.bernoulli qui par défaut est dense (connectivity = 1)*
+
+$W_{in}$ is a dense matrix generated using a Bernoulli (bimodal) distribution where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an equal probability where  $m = 1, …, M$ corresponds to a given feature in the input layer.
+The input scaling, denoted $I_{scale}$, is a hyperparameter coefficient often common to all features from the input layer or specific to each feature $m$.
+In that case, the more important the feature is, the greater should be its input scaling.
+$W$ is a matrix (usually sparse) where values are generated from a Gaussian distribution $\mathcal{N}(0,1)$.
+Then, the $W$ matrix is scaled according to the defined spectral radius, a hyperparameter defining the highest eigen value of $W$.
 
@@
 
-The parameter $\lambda$ is the ridge penalization which aims to prevent
-overfitting. One can also connect the input layer to the output layer in addition to the reservoir nodes. In that case, $X$ is the accumulation of both such that :
+The parameter $\lambda$ is the ridge penalization which aims to prevent overfitting.
+Additionally, one can also connect the input layer to the output layer to the reservoir nodes.
+In that case, $X$ is the accumulation of both such that :
 
@@
 Overall, there are four main hyperparameters to be chosen by the user: the leaking rate which defines the memory of the RC, the input scaling which define the relative importance of the features, the spectral radius which define the connections of the neurons inside the reservoir which in turn define the degree of non-linear combination of features and the ridge penalization which controls the degree of overfitting.
-The choice of hyperparameter often rely on a wrapper approach where the performance of the RC with different combinations of hyperparameters is evaluated on a train set and the best combination is chosen to forecast on the test set.
+The choice of hyperparameter often requires the user to evaluate the performance of different combinations of hyperparameters on a validation set before selecting the optimal combination to forecast on the test set.
 
@@
   theme_minimal() +
-  geom_vline(mapping = aes(color = "train-test sets",
-                           xintercept = traintest_date)) +
   labs(color = "") +
@@
 
--   Number of nodes (`units`) : it corresponds to the number of nodes inside the reservoir. Usually, the more the better but more nodes increases the computation time.
+-   Number of nodes (`units`) : it corresponds to the number of nodes inside the reservoir. Usually, the more the better, but more nodes increases the computation time.
 -   Leaking rate (`lr`) : the leaking rate corresponds to the balance between the new inputs and the previous state. A leaking rate of 1 only consider information from new inputs.
 -   Spectral radius (`sr`): the spectral radius is the maximum absolute eigenvalue of the reservoir connectivity matrix.
-A small spectral radius induces stable dynamics inside the reservoir, a high spectral radius induces chaotic regimen inside the reservoir.
+A small spectral radius induces stable dynamics inside the reservoir, a high spectral radius induces a chaotic regime inside the reservoir.
 -   Input scaling (`input_scaling`): the input scaling is a gain applied to the input features of the reservoir.
--   Seed (`seed`): because the reservoir connections are set at random, setting the seed is a good approach to ensure   reproducibility.
-Another approach would be to use several different reservoir but it increases computation time.
+- Warmup (`warmup`) : it corresponds to the number of time step during which the data are propagating into the reservoir but not used to fit the output layer.
+This hyperparameter is set in the `reservoirR_fit()` function.
+
+In addition, we can set the seed (`seed`). Because the reservoir connections are set at random, setting the seed is a good approach to ensure reproducibility.
 
@@
 
-Then we can feed the data to the reservoir and see the activation depending on time of the reservoir. To do so, we first prepare the data and transform it to a matrix :
+Then we can feed the data to the reservoir and see the activation state of the reservoir $x(t)$. To do so, we first prepare the data and transform it to a matrix:
 
@@
 Numerous nodes within the system exhibit a consistent equilibrium state.
-The challenge arises when the ridge output layer attempts to acquire knowledge from these nodes, as they do not convey meaningful information.
+The challenge arises when the output layer attempts to extract knowledge from these nodes, as they do not convey meaningful information.
 This issue can be attributed to the disparate scales of the features.
-To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby scaling them within the range of `0` to `1` by dividing by the maximum of the absolute value.
+To address this concern, a practical approach involves normalizing the features by dividing each of them by their respective maximum values, thereby scaling them within the range of $-1$ to $1$ by dividing by the maximum of the absolute value.
+Of note, here the features will be scaled between $0$ and $1$ because all features are positive.
 
@@
 
-```{r nodeactivationgood, fig.cap="20 random node acivation over time. Scaled features."}
+```{r nodeactivationgood, fig.cap="20 random node activation over time. Scaled features."}
 # feed them to the reservoir
@@
 
-To perform this task, we need to modify the training and testing process.
-Leveraging the inherent echo property of the RC, information from preceding time steps is preserved within the reservoir, effectively endowing the RC with a form of memory.
+To perform this task, we need to modify the training and testing processes.
+Leveraging the inherent inertia of the reservoir, information from preceding time steps is preserved, effectively endowing the RC with a form of memory.
 Consequently, the final state vector encapsulates insights gathered from all antecedent states.
@@
 This process is executed as follows:
+
+*XH : Markovian or “echo state” property ? c’est pour dire que les états finissent par être oubliés au bout d’un certain temps, donc c’est pas ce que tu veux dire là je pense*
+
+*TF : j'ai remplacé par inertia qui parait effectivement plus approprié. Mon idée c'est que l'information s'accumule dans le réservoir au fur et à mesure qu'il "lit" la séquence de données.*
 
@@
 
-In this paper, we introduce the R package `reservoirnet`, which serves as a versatile tool for implementing reservoir computing based on `ReservoirPy` Python API.
+In this paper, we introduce the R package `reservoirnet`, which serves as a versatile tool for implementing reservoir computing based on `ReservoirPy`'s Python API.
 It offers flexibility in defining the reservoir architecture, including options for specifying connections between the input layer and the output layer, as well as variations in input scaling as demonstrated on a real-world use case.
@@
 Drawing on the robust foundation of the `ReservoirPy` structure, a well-maintained library, this package inherits its reliability and longevity.
-Nevertheless, our emphasis has been on refining the primary functionalities, which may leave more advanced users desiring deeper access to core functions, such as the generation of the reservoir random connections.
-This initial version of `reservoirnet` must evolve in tandem with the growing understanding and adoption of reservoir computing within the R community.
+We have focused on providing access to the fundamental features, building upon the strong base provided by `ReservoirPy.` Therefore, this initial version of `reservoirnet` must evolve in tandem with the growing understanding and adoption of reservoir computing within the R community.
+
+*XH : Il faut rephrase mais je suis pas sur de ce qui tu veux dire.*
+
+*TF : j'ai reformulé ce dernier paragraphe en espérant que ça soit mieux.
+Mon idée c'était d'expliquer qu'on a focalisé le package sur les fonctions principales mais que tout n'est pas dispo depuis R (par exemple on ne peut pas choisir comment sont générées les matrices de connection aléatoires du réservoir, on les a laissé aux paramètres par défaut).*
 

### 2023-12-19 draft_jss/draft_jss.Rmd
update draft

@@
 abstract: >
-  The abstract of the article.
+  Reservoir Computing (RC) is a machine learning method that uses the dynamics of a high-dimensional "reservoir" to efficiently process information generated by dynamical systems.
+It has been successful in solving various tasks including time epidemic forecasting, language processing or bird calls.
+RC is implemented in Python and Julia but not in R.
+This article introduces **reservoirnet**, a R package providing access to the Python API **ReservoirPy**, allowing R users to harness the power of reservoir computing.
+This article provides an introduction to the fundamentals of RC and showcases its real-world applicability through three distinct sections.
+First, we cover the foundational concepts of RC, setting the stage for understanding its capabilities.
+Next, we delve into the practical usage of **reservoirnet**  through two illustrative examples.
+These examples demonstrate how it can be applied to real-world problems, specifically, regression of COVID-19 hospitalizations and classification of Japanese vowels.
+Finally, we present a comprehensive analysis of a real-world application of **reservoirnet**, where it was used to forecast COVID-19 hospitalizations at Bordeaux University Hospital using public data and electronic health records.
 keywords:
   # at least one keyword must be supplied
-  formatted: [keywords, not capitalized, "\\proglang{Java}"]
+  formatted: ["\\textbf{reservoirnet}", "\\textbf{ReservoirPy}", Reservoir computing, COVID-19, time series]
   plain:     [keywords, not capitalized, Java]
@@
 
-Numerous COVID-19 forecasting algorithms have been proposed using different methods (e.g ensemble, deep learning, compartmental), yet none has proven entirely satisfactory [@rahimi_review_2021, @cramer_evaluation_2022].
+Numerous COVID-19 forecasting algorithms have been proposed using different methods (e.g ensemble, deep learning, compartmental), yet none has proven entirely satisfactory [@rahimi_review_2021, @cramer_evaluation_2022-1].
 In France, short-term forecasts  with different methods have been evaluated with similar results [@pottier_forecast_2021, @paireau_ensemble_2022, @carvalho_analysis_2021, @mohimont_convolutional_2021].
@@
 $MAEB$ and $MREB$ compare model performance to a baseline model which predicts the current number of hospitalized patients at 14 days.
-Those metrics help to determine the information added by the model and is a good baseline as covid-19 forecast model do not always outperform this basic forecast (@cramer_evaluation_2022).
+Those metrics help to determine the information added by the model and is a good baseline as covid-19 forecast model do not always outperform this basic forecast (@cramer_evaluation_2022-1).
 

### 2023-12-19 script/forecast/forecast_test.R
add importance in functions

@@ -14,6 +14,7 @@ warmup = 30
 units = 500
 nb_iter = 1
 nb_sample = 40
+importance = TRUE
 ##### LOAD BEST HP SETS
 hp_sets = list(common_input_scaling = "data/common_input_scaling_11559350/",
                common_input_scaling_linked_source = "data/common_input_scaling_11559390/",
@@ -45,82 +46,92 @@ date_i <- vecDates[slar_taskid]
 message(paste0("--------- evaluation date : ", date_i, " ----------"))
 
 ##### FORECAST
-dfres <- lapply(names(best_hp_set),
-                function(model_set){
-                  print(model_set)
-                  hp_set <- best_hp_set[[model_set]]
-                  if(model_set == "enet"){
-                    ##### ELASTIC-NET
-                    forecast_enet_once <- fct_iterative_forecast_from_hp(data_covid = data_covid,
-                                                                         vecDates = date_i,
-                                                                         forecast_days = forecast_days,
-                                                                         lambda = best_hp_set$enet$lambda,
-                                                                         alpha = best_hp_set$enet$alpha,
-                                                                         model = "enet") %>%
-                      mutate(model = "enet_hp_on_train_set",
-                             hp_set = hp_set$hp_set)
-                    forecast_enet_daily <- fct_iterative_forecast_from_hp(data_covid = data_covid,
-                                                                          vecDates = date_i,
-                                                                          forecast_days = forecast_days,
-                                                                          model = "enet") %>%
-                      mutate(model = "enet_daily_hp_update",
-                             hp_set = 1)
+lsres <- list(dfImp = NULL, dfPred = NULL)
+
+lapply(names(best_hp_set),
+       function(model_set){
+         print(model_set)
+         hp_set <- best_hp_set[[model_set]]
+         if(model_set == "enet"){
+           ##### ELASTIC-NET
+           forecast_enet_once <- fct_iterative_forecast_from_hp(data_covid = data_covid,
+                                                                vecDates = date_i,
+                                                                forecast_days = forecast_days,
+                                                                lambda = best_hp_set$enet$lambda,
+                                                                alpha = best_hp_set$enet$alpha,
+                                                                model = "enet",
+                                                                importance = importance)
+           lsres$dfPred <<- forecast_enet_once$dfPred %>%
+             mutate(model = "enet_hp_on_train_set",
[…]

### 2023-12-19 draft_jss/draft_jss.Rmd
update model result ongoing work

@@
 
+### Input feature importance
+
+For the model where the input layer is directly connected to the output layer, we can compare the coefficient of the output layer estimated for the input layer and the reservoir nodes.
+We can also compare the coefficient given to the input layer by the output layer in the reservoir and the coefficient estimated by the elastic-net model.
+
+Figure \@ref(fig:importanceReservoir) outlines the rank of input layer compared to all the connections to the output layer including the 500 nodes and the 16 features (bias is not represented).
+We observe that the model with common input scaling tend to give less weight to the raw features compared to the model with multiple input scaling.
+
+```{r importanceReservoir, fig.cap="Mean feature importance of the 40 best hyperparameter sets by model.
+Only models with connection between input and output layer are included.
+The more weight the output layer attributes to the input layer, the closer its rank will be to 1 and the more the feature is important.
+Rank is obtained by comparing the feature output layer coefficient and all other connection coefficient."}
+ls_use_case$plot_reservoir_importance
+```
+
+Figure \@ref(fig:importanceReservoirvsEnet)
+
+```{r importanceReservoirvsEnet, fig.cap="Mean feature coefficient of the 40 best hyperparameter sets by model and the elastic-net model.
+Only models with connection between input and output layer are included."}
+ls_use_case$plot_importance
+```
+
 ### Forecast performance

### 2023-12-19 script/forecast/forecast_test_eval.R

@@ -1,11 +1,12 @@
 ##### load packages and functions
 library(dplyr)
 library(ggplot2)
+library(scales)
 
 source(here::here("functions/fct_smoothing_derivative.R"))
 
 ##### load data
-ls_files_results <- list.files("data/results_forecast_testset_11561202/", full.names = TRUE)
+ls_files_results <- list.files("data/results_forecast_testset_11922514/", full.names = TRUE)
 ls_files_results_res <- grep(ls_files_results, pattern = "hp_sets", value = TRUE, invert = TRUE)
 hp_file <- grep(ls_files_results, pattern = "hp_sets", value = TRUE)
 df_obfuscated_epidemio <- readRDS("data/df_obfuscated_epidemio.rds")
@@ -180,8 +181,76 @@ plot_feature_input_scaling <- df_hp %>%
        y = "Mean absolute error",
        color = "")
 
+##### importance plots
+df_importance <- lapply(ls_files_results_res, function(x) readRDS(x)$dfImp) %>%
+  dplyr::bind_rows()
+
+df_importance_no_reservoir <- df_importance %>%
+  dplyr::mutate(model = factor(model,
+                               levels = names(model_labels),
+                               labels = model_labels),
+                feature = factor(feature,
+                                 levels = names(feature_labels),
+                                 labels = feature_labels)) %>%
+  filter(!is.na(feature)) %>%
+  group_by(feature, model, hp_set) %>%
+  summarise(importance = mean(importance), .groups = "drop")
+
+ordered_feature_imp <- df_importance_no_reservoir %>%
+  filter(model == "Elastic-net") %>%
+  arrange(importance) %>%
+  pull(feature) %>%
+  as.character()
+
+S_sqrt <- function(x){sign(x)*sqrt(abs(x))}
+IS_sqrt <- function(x){x^2*sign(x)}
+S_sqrt_trans <- function() trans_new("S_sqrt",S_sqrt,IS_sqrt)
+
+# mean feature coefficient by hp set and by model.
+plot_importance <- df_importance_no_reservoir %>%
+  mutate(feature = factor(feature, levels = ordered_feature_imp)) %>%
+  ggplot(mapping = aes(y = feature, color = model, x = importance)) +
+  geom_boxplot(position = position_dodge(width = 0.9)) +
+  geom_vline(xintercept = 0, color = "black", linetype = 2) +
+  scale_alpha_manual(values = c(0.2, 0.2, 1)) +
[…]

### 2023-12-20 draft_jss/draft_jss.Rmd
update draft

--- 2023-12-19 draft_jss/draft_jss.Rmd
@@
 Additionaly, connecting input layer to output layer improved the model forecast.
-
-We observe that elastic-net did not perform well and stayed close to the baseline prediction.
-However, when the penalisation is updated each day, the peformance improve.
-This is emphasize the problem of the transferrability of the hyperparameters to which multiple input scaling models seem more sensitive as they have more hyperparameters.
+All RC models performed better than the elastic-net model.
 
@@
 ls_use_case$plot_aggregation
+```
+
+### Input feature importance
+
+We compared the coefficients of the output layer estimated for the input layer and the reservoir nodes.
+Additionally, we compared the coefficient given to the input layer by the output layer in the reservoir and the coefficient estimated by the elastic-net model.
+
+Figure \@ref{fig:importanceReservoir} illustrates the ranking of input layer compared to all connections to the output layer, including the 500 reservoir nodes and the 16 features of the input layer (excluding bias).
+The figure shows that the model with common input scaling tends to assign less weight to input layer compared to the model with multiple input scaling.
+This suggests that the reservoir with common input scaling provides more information than the reservoir with multiple input scaling, which aligns with its better performance, as shown in Table \@ref{tab:performance}.
+
+```{r importanceReservoir, fig.cap="Mean feature importance of the 40 best hyperparameter sets by model, focus on the connection between the input and output layers.
+Models with direct connection between input and output layer are included.
+The rank is obtained by comparing the feature input layer and all other connection coefficients (both input and reservoir corresponding coefficients) attributed by the output layer at each date for each hyperparameter set.
+The higher the output layer's coefficient for the input layer, the closer its rank will be to 1 and the more important the feature is.", fig.height=6}
+ls_use_case$plot_reservoir_importance
+```
+
+Furthermore, Figure \@ref{fig:importanceReservoirvsEnet} compares the coefficients assigned to input features by the elastic-net model and the RC models.
+While the coefficients are generally consistent across RC models, there are some notable differences with elastic-net.
+Specifically, certain features deemed important by the elastic-net model, such as the derivative of RT-PCR, and the derivative of Vaccine, are less important for the reservoir computing model.
+This may indicate that these features predictive ability is better conveyed by their relationship with other features, which is captured by the reservoir computing model but might not be by the elastic-net model.
+Conversely, emergency, IPTCC, proportion of positive RT-PCR, and hospitalizations are more important for the reservoir computing model than for the elastic-net model.
+
+```{r importanceReservoirvsEnet, fig.cap="Mean feature coefficient of the 40 best hyperparameter sets by model and the elastic-net model.
+Only models with direct connection between input and output layer are included.
+The coefficients were calculated as the average value across all dates for each feature, model and hyperparamet set.", fig.height=6}
+ls_use_case$plot_importance
 ```

### 2023-12-20 script/forecast/forecast_test_eval.R

@@ -217,6 +217,7 @@ plot_importance <- df_importance_no_reservoir %>%
   scale_color_manual(values = c("#023047", "#8ECAE6", "#FB8500")) +
   theme_minimal() +
   theme(legend.position = "bottom") +
+  guides(color=guide_legend(nrow=2)) +
   labs(x = "Coefficient", y = "", color = "", fill = "")
 
 df_importance_reservoir <- df_importance %>%
@@ -232,11 +233,9 @@ df_importance_reservoir <- df_importance %>%
   mutate(rank = dense_rank(desc(abs(importance)))) %>%
   ungroup() %>%
   filter(feature %in% feature_labels) %>%
-  group_by(model, hp_set, feature) %>%
-  summarise(rank = mean(rank), .groups = "drop") %>%
   mutate(feature = factor(feature, levels = ordered_feature_imp))
 
-# mean feature importance by hp set and by esn model.
+# feature importance by hp set and by esn model.
 nb_features = df_importance$feature %>% unique() %>% length()
 plot_reservoir_importance <- df_importance_reservoir %>%
   ggplot(mapping = aes(y = feature, color = model, x = rank)) +

### 2023-12-20 draft_jss/draft_jss.Rmd
update before 2nd round

@@
 
-*TF : on a pris les paramètres par défaut de reservoirpy, si j'ai bien compris, la matrice est générée par la fonction reservoirpy.mat_gen.bernoulli qui par défaut est dense (connectivity = 1)*
+*TF : my bad, j'avais mal lu la doc. On utilise l'implémentation de base de reservoirpy qui est effectivement sparse à 0.1 par défaut.*
 
-$W_{in}$ is a dense matrix generated using a Bernoulli (bimodal) distribution where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an equal probability where  $m = 1, …, M$ corresponds to a given feature in the input layer.
+$W_{in}$ is a matrix (usually sparse) generated using a Bernoulli (bimodal) distribution where each value can be either $-I_{scale}(m)$ or $I_{scale}(m)$ with an equal probability where $m = 1, …, M$ corresponds to a given feature in the input layer.
 The input scaling, denoted $I_{scale}$, is a hyperparameter coefficient often common to all features from the input layer or specific to each feature $m$.

### 2024-05-23 draft_jss/draft_jss.Rmd
update results after BH review

@@
 
-```{r fig.cap="Forecast"}
+```{r fig.cap="Forecast", fig.height=4, fig.width=7}
 dfOutcome %>%
@@
                           color = "forecast")) +
-  geom_vline(mapping = aes(color = "train-test sets",
-                           xintercept = traintest_date)) +
-  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
+  annotate("rect",
+           xmin = traintest_date,
+           xmax = max(dfOutcome$outcomeDate, na.rm = T),
+           ymin = 0,
+           ymax = max(dfOutcome$outcome, na.rm = T)*1.1, 
+           alpha = .2) +
+  annotate("text", label = "Test set", x = as.Date("2022-08-01"), y = 2200, size = 7) +
+  annotate("text", label = "Train set", x = as.Date("2021-03-01"), y = 2200, size = 7) +
+  scale_color_manual(values = c("#3772ff", "#080708")) +
   theme_minimal() +
@@
 ```{r figForecastDerivvsRaw, fig.cap="Covid-19 hospitalizations forecast.
-The model is either trained to forecast the number of hospitalizations (denoted Raw) or the variation of the hospitalizations compared to current level of hospitalisation (denoted Variation)"}
+The model is either trained to forecast the number of hospitalizations (denoted Raw) or the variation of the hospitalizations compared to current level of hospitalisation (denoted Variation)", fig.height=4, fig.width=7}
 ## Fit reservoir on outcome variation instead of raw outcome
@@
                           color = "Forecast")) +
-  geom_vline(mapping = aes(color = "train-test sets",
-                           xintercept = traintest_date)) +
+  annotate("rect",
+           xmin = traintest_date,
+           xmax = max(dfOutcome$outcomeDate, na.rm = T),
+           ymin = 0,
+           ymax = max(dfOutcome$outcome, na.rm = T)*1.1, 
+           alpha = .2) +
+  annotate("text", label = "Test set", x = as.Date("2022-08-01"), y = 2200, size = 5) +
+  annotate("text", label = "Train set", x = as.Date("2021-03-01"), y = 2200, size = 5) +
   facet_wrap(Outcome_type ~ .,
              labeller = label_bquote(cols = "Outcome" : .(Outcome_type))) +
-  scale_color_manual(values = c("#3772ff", "#080708", "#df2935")) +
+  scale_color_manual(values = c("#3772ff", "#080708")) +
   theme_minimal() +
+  theme(legend.position = "bottom") +
   labs(color = "", x = "Date", y = "Hospitalizations")

### 2024-05-23 script/forecast/forecast_test_eval.R

@@ -141,7 +141,8 @@ plot_reservoir_hp <- df_hp %>%
   filter(HP %in% hp_features_reservoir) %>%
   ggplot(mapping = aes(x = HP_value, 
                        y = mean_absolute_error,
-                       color = rank_perf_color)) +
+                       color = rank_perf_color,
+                       alpha = rank_perf_color)) +
   geom_point() +
   facet_grid(model ~ HP, scales = "free_x",
              labeller = labeller(model = model_labels,
@@ -150,21 +151,25 @@ plot_reservoir_hp <- df_hp %>%
                 breaks = c(1e-10, 1e-5, 1e0, 1e5, 1e10),
                 minor_breaks = 10^(seq(-10, 5))) +
   scale_color_manual(values = c("blue", "#E76F51", "#E9C46A", "#2A9D8F")) +
+  scale_alpha_manual(values = c(1, 1, 1, 0.1)) +
   lims(y = c(15, 30)) +
   theme_minimal() + 
-  theme(strip.text.y = element_text(angle = 0),
+  theme(legend.position = "bottom",
+        axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0),
+        strip.text.y = element_text(angle = 0),
         strip.text.x = element_text(angle = 90)) +
-  theme(legend.position = "bottom") +
   labs(x = "Hyperparameter value", 
        y = "Mean absolute error",
-       color = "") 
+       color = "",
+       alpha = "") 
 
 plot_feature_input_scaling <- df_hp %>%
   filter(HP %in% hp_features,
          model == "multiple_input_scaling_linked_source") %>%
   ggplot(mapping = aes(x = HP_value, 
                        y = mean_absolute_error,
-                       color = rank_perf_color)) +
+                       color = rank_perf_color,
+                       alpha = rank_perf_color)) +
   geom_point() +
   facet_wrap(HP ~ ., scales = "free_x",
              labeller = labeller(model = model_labels,
@@ -175,8 +180,10 @@ plot_feature_input_scaling <- df_hp %>%
                 minor_breaks = 10^(seq(-10, 5))) +
   scale_y_continuous(breaks = c(15, 20, 30)) +
   scale_color_manual(values = c("blue", "#E76F51", "#E9C46A", "#2A9D8F")) +
+  scale_alpha_manual(values = c(1, 1, 1, 0.1)) +
   theme_minimal() + 
-  theme(legend.position = "bottom") +
[…]

### 2024-05-23 draft_jss/draft_jss.Rmd

--- 2023-12-20 draft_jss/draft_jss.Rmd
@@
 ```{r eval=FALSE}
+# Install reservoirnet package from CRAN
 install.packages("reservoirnet")
@@
 ```{r message=FALSE, warning=FALSE}
+# Load usefull packages
 library(dplyr)
@@
 library(reservoirnet)
-
+# load dfCovid data from the reservoirnet package which contains Covid data
 data("dfCovid")
+# Set the forecast horizon to 14 days
 dist_forecast = 14
+# Set the train-test split to 2022-01-01
 traintest_date = as.Date("2022-01-01")
@@
   mutate(outcome = lead(x = hosp, n = dist_forecast),
+         # Create a new column 'outcome' which contains the number of hospitalizations ('hosp')
+         # shifted forward by 'dist_forecast' days (14 days). This represents the outcome we want to predict.
+
          outcomeDate = date + dist_forecast,
+         # Create a new column 'outcomeDate' which is the current date plus the forecast period (14 days).
+
          outcome_deriv = outcome - hosp) %>%
+         # Create a new column 'outcome_deriv' which is the difference between the predicted outcome and current hospitalizations.
+         # This represents the change in hospitalizations over the forecast period.
+
   # rolling average for tested and positive_pcr
@@
                                                   .f = mean))
+            # Apply a rolling mean (7-day average) to the 'Positive' and 'Tested' columns.
+            # The 'slider::slide_dbl' function is used to calculate the mean over a window of 7 days (current day + 6 days before).
+            # This smooths out daily fluctuations and provides a better trend indicator.
+
 ```
@@
 ```{r covidintro, fig.cap="Hospitalizations, IPTCC and positive PCR of Bordeaux University Hospital.", echo = FALSE}
+# Create a figure with hospitalizations, positive and tested RT-PCR from covid data
 dfOutcome %>%
@@
 ```{r}
+# Create a reservoir computing node using the 'createNode' function from the reservoirnet package.
+# Arguments:
+# - nodeType = "Reservoir": Specify the type of node to be a reservoir.
+# - seed = 1: Set the seed for reproducibility, ensuring consistent results when the model is run multiple times.
+# - units = 500: Set the number of reservoir units (neurons) to 500.
+# - lr = 0.7: Set the leakage rate (lr) of the reservoir, which controls how quickly the reservoir state decays over time.
+# - sr = 1: Set the spectral radius (sr) of the reservoir, which influences the stability and memory capacity of the reservoir.
+# - input_scaling = 1: Set the input scaling factor, which scales the input signal before it is fed into the reservoir.
+
 reservoir <- reservoirnet::createNode(nodeType = "Reservoir",
@@
 ```{r}
-## select explanatory and transform it to an array
+## select explanatory features of the train set and transform it to an array
 X <- dfOutcome %>%
@@
 ```{r}
+# Generate the state of the reservoir using the 'predict_seq' function from the reservoirnet package.
+# Arguments:
+# - node = reservoir: The reservoir computing node created earlier.
+# - X = X: The input data matrix containing the features 'hosp', 'Positive', and 'Tested'.
+# The function computes the state of the reservoir for each time step in the input sequence,
+# effectively transforming the input data into the reservoir's high-dimensional state space.
+
 reservoir_state <- predict_seq(node = reservoir, X = X)
@@
 ```{r nodeactivationbad, fig.cap="20 random nodes activation over time."}
+# Plot the reservoir state activation over time
 plot(reservoir_state)
@@
 ```{r}
+# Standardise features by dividing by the maximum value can improve performance
+# After standardisation, all features are on a similar scale which helps RC
 stand_max <- function(x) return(x/max(abs(x)))
@@
 ```{r nodeactivationgood, fig.cap="20 random node activation over time. Scaled features."}
-# feed them to the reservoir
+# feed the scaled features to the reservoir
 reservoir_state_stand <- predict_seq(node = reservoir,
@@
                                     ridge = 1e3)
+# Create a readout node using ridge regression with the 'createNode' function from the reservoirnet package.
+# Arguments:
+# - nodeType = "Ridge": Specify the type of node to be a ridge regression readout.
+# - ridge = 1e3: Set the regularization parameter (ridge) for the ridge regression to 1000.
+# Ridge regression is used to prevent overfitting by adding a penalty on the size of the coefficients.
+
 model <- reservoirnet::link(reservoir, readout)
+# Link the reservoir and readout nodes to form a complete reservoir computing model.
+# The 'link' function connects the high-dimensional state generated by the reservoir to the readout layer,
+# allowing the model to learn the mapping from the reservoir states to the target outputs.
 ```
@@
 ```{r}
+# Perform some data management to isolate train and test sets
 # train set
@@
 ```{r}
+# Forecast with the trained reservoir on the test data
 vec_pred <- reservoirnet::predict_seq(node = fit$fit,
@@
 
-```{r fig.cap="Forecast"}
+```{r fig.cap="Forecast", fig.height=4, fig.width=7}
+# Make figure to represent forecast on the train and test sets.
+
 dfOutcome %>%
@@
                                                  reset = TRUE)
-## Transform the outome variation forecast into hospitalisation forecast
+## Transform the outcome variation forecast into hospitalization forecast
 vec_pred2 <- vec_pred2_variation + xTest$hosp
@@
 ```{r}
+# Get the Japanese vowels dataset using the 'generate_data' function from the reservoirnet package.
+# The dataset contains preprocessed features and labels for classification.
+# Then we isolate train and test sets
 japanese_vowels <- reservoirnet::generate_data(dataset = "japanese_vowels")[[1]]
@@
 ```{r vowelpresentation, fig.cap="Vowel dataset, sample with 3 speakers and 2 utterance each.", echo = FALSE}
+# Make a figure of the Vowel dataset of selected participants.
 vec_sample <- c(1, 2, 41, 42, 71, 72)
@@
                                       seed = 1)
+# Create a reservoir computing node with 500 units using the 'createNode' function from the reservoirnet package.
+# Arguments:
+# - units = 500: Set the number of reservoir units (neurons) to 500.
+# - lr = 0.1: Set the leakage rate (lr) of the reservoir to 0.1, controlling how quickly the reservoir state decays over time.
+# - sr = 0.9: Set the spectral radius (sr) of the reservoir to 0.9, influencing the stability and memory capacity of the reservoir.
+# - seed = 1: Set the seed for reproducibility, ensuring consistent results when the model is run multiple times.
 readout <- reservoirnet::createNode("Ridge",ridge=1e-6)
+# Create a readout node using ridge regression with the 'createNode' function from the reservoirnet package.
+# Arguments:
+# - ridge = 1e-6: Set the regularization parameter (ridge) for the ridge regression to 1e-6.
+# Ridge regression is used to prevent overfitting by adding a penalty on the size of the coefficients.
 ```
@@
 for (x in X_train) {
+  # Loop over each training sample in X_train.
   states <- reservoirnet::predict_seq(node = reservoir, X = x,
                                       reset=TRUE)
+  # Generate the reservoir states for the current training sample using the 'predict_seq' function.
   states_train[[k]] <- t(as.matrix(states[nrow(states),]))
+  # Extract the final reservoir state for the current sample and store it in 'states_train'.
   k <- k+1
@@
 ```{r}
+# Fit the reservoir using the last state vector (each observation is the whole vowel sequence)
 res <- reservoirnet::reservoirR_fit(readout,X = states_train, Y = Y_train)
@@
 ```{r}
+# The operation is repeated for the test set :
+# - the last state vector is extracted for each utterance
+# - the model is used to forecast the utterance from this last state vector
 Y_pred <- list()
@@
 For each of the 6 utterance, the model correctly identifies the speaker."}
+# A figure represents the performance on the test set
 dfplotseqtovec <- lapply(vec_sample,
@@
 ```{r}
+# The overall accuracy is evaluated
 accuracy <- function(pred, truth) mean(pred == truth)
@@
 ```{r}
+# For this new task where we want to forecast for each time step (instead of each utterance)
+# we start by getting the data in the appropriate format
+# Then we split the train and test data
 japanese_vowels <- reservoirnet::generate_data(
@@
 ```{r}
+# Create an input, a reservoir and an output layers
 source <- createNode("Input")
@@
 reservoir <- createNode("Reservoir",units = 500,lr=0.1, sr=0.9, seed = 1)
-
+# Connect the input layer to the reservoir and connect both the input layer and the reservoir to the output layer
 model <- list(source %>>% reservoir, source) %>>% readout
@@
 ```{r}
+# Fit the RC model
 model_fit <- reservoirnet::reservoirR_fit(node = model,
@@
                                           warmup = 2)
-
+# Predict with the fitted model
 Y_pred <- reservoirnet::predict_seq(node = model_fit$fit,
@@
 ```{r figseqtoseq, fig.cap="Prediction in a sequence-to-sequence approach 6 samples with 3 speakers and 2 utterance each. The higher the score of the speaker, the lighter the color.", fig.height=6}
+# Make a graph with a label for each time of each utterance
 dfplotseqtoseq <- lapply(vec_sample,
@@
 ```{r}
+# Compute the accuracy
 Y_pred_class <- sapply(Y_pred, FUN = function(x) apply(as.matrix(x),
@@
 ```{r echo=FALSE}
+# because computing time for this section is long, all results were computed
+# using a calculation server and stored into a rds file
 ls_use_case <- readRDS(file = here::here("data/precomputed_results.rds"))
@@
 Other features are scaled (divide by the maximum of the feature) represented in darkblue.", fig.height=8}
+# descriptive plot of the data
 ls_use_case$plot_present_data
@@
 ```{r ucHyperparam, echo=FALSE, fig.cap="Hyperparameter evaluation on training set by random search. Hp sets with MAE above 30 were removed for clarity of visualisation.", fig.height=8}
+# hyperparameter search
 ls_use_case$plot_reservoir_hp
@@
 Hp sets with MAE above 30 were removed for clarity of visualisation.", fig.height=8}
+# hyperparameter search, focus on input scaling
 ls_use_case$plot_feature_input_scaling
@@
 ```{r performance, echo = FALSE}
+# performance of the different model
 ls_use_case$table_perf_esn %>%
@@
 Grey lines are individual forecast of each of the 40 reservoir.", fig.height=8}
+# plot of the forecast of the different models
 ls_use_case$plot_forecast
@@
 Forecast value above 200 were set to 200 for clarity.", fig.height=6}
+# plot of different models without forecast aggregation.
 ls_use_case$plot_before_aggregation_forecast
@@
 ```{r ucAggregationNb, echo = FALSE, fig.cap = "Mean absolute error depending on the number of aggregated reservoir.", fig.height=6}
+# plot models depending on number of aggregated reservoir.
 ls_use_case$plot_aggregation
@@
 The higher the output layer's coefficient for the input layer, the closer its rank will be to 1 and the more important the feature is.", fig.height=6}
+# Plot raw feature importance
 ls_use_case$plot_reservoir_importance
@@
 The coefficients were calculated as the average value across all dates for each feature, model and hyperparamet set.", fig.height=6}
+# Plot feature importance of different models
 ls_use_case$plot_importance

### 2024-05-24 script/forecast/forecast_test_eval.R
update figure for legend adaptation

@@ -186,7 +186,8 @@ plot_feature_input_scaling <- df_hp %>%
         axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0)) +
   labs(x = "Hyperparameter value", 
        y = "Mean absolute error",
-       color = "")
+       color = "",
+       alpha = "")
 
 ##### importance plots
 df_importance <- lapply(ls_files_results_res, function(x) readRDS(x)$dfImp) %>%
